---
title: "Zero Inflated Beta Regression"
author: "Jonas J. R. Koberschinski"
editor: visual
---

## Prep

```{r}
# 
library(haven) # For reading SPSS files
library(tidyverse) # For data manipulation and visualization
library(dplyr) # For data manipulation

# Excel
library(readxl) # For reading Excel files
library(writexl) # For writing Excel files

# 
library(lme4) # For linear mixed-effects models
library(nlme) # For linear and nonlinear mixed effects models
library(glmmTMB) # For generalized linear mixed models
library(pscl) # For zero-inflated models

# 
library(performance) # For model performance assessment
library(DHARMa) # For residual diagnostics

library(ggplot2) # For data visualization



```

Zero-inflated Beta-Binomial models with the2
scDMV R package

## Data import

```{r}
 ## Load the BDNF data from a .sav file
 ### ??? more information !!!  restructured without methylation mean!
bdnf <- read_excel("~/Work/TBM-R-Analysis/tables/Rohdaten/BDNF_Neuroplastizität_Alter_CpGs_ohne_m11_und_p36_mit_verhalten_vollständig_umstrukturiert.xlsx")

tau <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/TAU_CpGs_mit_verhalten_ohne_p61_p59_p34_p29_umstruk.xlsx"))

psd95 <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/PSD95_CpGs_mit_verhalten_größer_umstrukturiert.xlsx"))

 ##first look at BDNF data
head(bdnf)
summary(bdnf)
## Check the structure of the dataset
str(bdnf)

```

## Subset Data and Drop NA

```{r}
# Remove rows with NA values in the 'methylierung' column

##BDNF
bdnf_subset <- subset(bdnf, select = c(ID, Standort, Zeitpunkt, Gruppe, Sex, Age_start, CogTel, CpG_position, methylierung))
bdnf_nona <- drop_na(bdnf_subset)
```

## BDNF (all cpgs)
### Normality check

```{r}
# Descriptive statistics for methylierung
mean(bdnf_nona$methylierung)
sd(bdnf_nona$methylierung)
median(bdnf_nona$methylierung)

# Another Histogram
hist(bdnf_nona$methylierung, breaks = 60, main = "BDNF Methylation distribution", xlab = "Methylation")
hist(tau$methylation, breaks = 60, main = "TAU Methylation distribution", xlab = "Methylation")
hist(psd95$methylation, breaks = 60, main = "PSD95 Methylation distribution", xlab = "Methylation")

# QQ plot
qqnorm(bdnf_nona$methylierung)
qqline(bdnf_nona$methylierung, col = "red")

# Shapiro-Wilk normality test
shapiro.test(bdnf_nona$methylierung)

# Boxplot to visualize methylierung distribution
boxplot(bdnf_nona$methylierung, main="Boxplot of BDNF Methylation Levels", ylab="Methylation Levels")
```


zero-inflated Poisson distributions {\displaystyle \mu } = 0,114 {\displaystyle \pi } = 426


```{r}
#summary(m1 <- zeroinfl(bdnf_nona$methylierung ~ bdnf_nona$Zeitpunkt + bdnf_nona$Group | #bdnf_nona$id_umstrukturiereninst, data = bdnf_nona))
```


## Descriptive statistics for zero-inflation
```{r}
# Total count
bdnf_nona %>%
  count() # 3603


# Count of zeros
bdnf_nona %>%
  filter(methylierung == 0) %>%
  count() # 1643

# Count of non-zeros
bdnf_nona %>%
  filter(methylierung > 0) %>%
  count() # 1960

# Proportion of zeros
bdnf_nona %>%
  summarise(prop_zero = mean(methylierung == 0))


```

```{r}
bdnf_nona %>%
  summarise(
    total_mean = mean(methylierung),
    nonzero_mean = mean(methylierung[methylierung > 0]),
    n_total = n(),
    n_zero = sum(methylierung == 0),
    prop_zero = mean(methylierung == 0)
  )

```








## Zero-inflated Beta-Binomial models
```{r}
#library(gamlss)

# Beta regression mit Zero-Inflation (BEINF-Family)
#zibeta_mod <- gamlss(
#  methylierung ~ Standort + Zeitpunkt + Gruppe + Sex + Age_start + CogTel + CpG_position + random(as.factor(ID)) + random(as.factor(Zeitpunkt)),
#  family = BEINF(),
#  data = bdnf_nona
#)
#summary(zibeta_mod)
```

```{r}
library(glmmTMB)

# Zero-inflated Beta
mod <- glmmTMB(
  methylierung ~ Standort + Zeitpunkt + Gruppe + Sex + Age_start + CogTel + (1|ID),
  ziformula = ~1,        # model for zeros
  family = beta_family(link = "logit"),
  data = bdnf_nona
)
summary(mod)
#preds <- predict(mod, type="response")
#range(preds)
```


```{r}
library(glmmTMB)

# Zero-inflated Beta
mod_CpG_ID <- glmmTMB(
  methylierung ~ Standort + Gruppe + Sex + Age_start + CogTel + Zeitpunkt + (1|CpG_position) + (1|ID),
  ziformula = ~ ID + Zeitpunkt + Gruppe,        # model for zeros
  family = beta_family(link = "logit"),
  data = bdnf_nona
)
summary(mod_CpG_ID)
#preds <- predict(mod, type="response")
#range(preds)
```
ID as random effect improved model slightly


Reduktion in Odds!

```{r}
library(dplyr)
library(ggplot2)

# Daten vorbereiten
bdnf_plot <- bdnf_nona %>%
  mutate(is_zero = ifelse(methylierung == 0, 1, 0)) %>%
  group_by(CpG_position, Gruppe) %>%
  summarise(
    mean_methyl = mean(methylierung[methylierung > 0], na.rm = TRUE),
    prop_zero = mean(is_zero, na.rm = TRUE),
    .groups = "drop"
  )

# Beta-Teil (Mittelwert der Methylierung)
p1 <- ggplot(bdnf_plot, aes(x = CpG_position, y = mean_methyl, fill = Gruppe)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Beta-Teil: Mittelwert der Methylierung (>0)",
       y = "Mittelwert", x = "CpG Position") +
  scale_fill_brewer(palette = "Set1")

# Zero-Inflation Teil
p2 <- ggplot(bdnf_plot, aes(x = CpG_position, y = prop_zero, fill = Gruppe)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Zero-Inflation: Anteil Nullwerte",
       y = "Proportion Nullwerte", x = "CpG Position") +
  scale_fill_brewer(palette = "Set2")

# Plots anzeigen
library(gridExtra)
grid.arrange(p1, p2, nrow = 2)

```

```{r}
testDispersion(mod)
```


```{r}
library(DHARMa)

# Simulate residuals
sim_res <- simulateResiduals(fittedModel = mod_CpG_ID, n = 1000)
residuals(sim_res)

# Plot residuals
plot(sim_res)
```

https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html

```{r}
check_model(mod)
check_model(mod_CpG_ID)
```





## BDNF (High Variance CpGs)

```{r}
bdnf_reduced <- bdnf %>%
  dplyr::select(methylierung, Group, Zeitpunkt, Standort, Geschlecht, Age_start, CogTel, id_umstrukturieren, CpG_position) %>%
  dplyr::filter(CpG_position %in% c("m148", "m066", "m111", "m058", "m039")) #( maybe add m035)
```

```{r}
hist(bdnf_reduced$methylierung, breaks = 60, main = "BDNF Methylation distribution", xlab = "Methylation")
```

```{r}
library(glmmTMB)

# Zero-inflated Beta
mod <- glmmTMB(
  methylierung ~ Standort + Zeitpunkt + Group + Geschlecht + Age_start + CogTel,
  ziformula = ~ Group + Zeitpunkt,        # model for zeros
  family = beta_family(link = "logit"),
  data = bdnf_reduced
)
summary(mod)
#preds <- predict(mod, type="response")
#range(preds)
```


Bestes Modell (Geringste AIC)
methylierung ~ Standort + Gruppe + Sex + Age_start + CogTel + Zeitpunkt + (1 | CpG_position) + (1 | ID)

