--- 
title: "Generalized Linear Models"
author: "Jonas J. R. Koberschinski"
date: "20-10-2025"
output: html
---


## Setup

```{r}
#Prep
 ## Set working directory
#setwd("~/Work/Promotionsdatananalyse mit Excel und Martin/Rohdaten/")
#print(getwd())

 ## Load necessary packages
library(haven)
library(tidyverse)
library(dplyr)
library(readxl)
library(writexl)
library(nlme)
library(lme4)

```

## Data import

```{r}
 ## Load the BDNF data from a .sav file
 ### ??? more information !!!  restructured without methylation mean!
bdnf <- read_excel("~/Work/TBM-R-Analysis/tables/Rohdaten/BDNF_Neuroplastizität_Alter_CpGs_ohne_m11_und_p36_mit_verhalten_vollständig_umstrukturiert.xlsx")

tau <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/TAU_CpGs_mit_verhalten_ohne_p61_p59_p34_p29_umstruk.xlsx"))

psd95 <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/PSD95_CpGs_mit_verhalten_größer_umstrukturiert.xlsx"))

 ##first look at BDNF data
head(bdnf)
summary(bdnf)
 ## Check the structure of the dataset
str(bdnf)

```


## Subset Data and Drop NA
```{r}
# Remove rows with NA values in the 'methylierung' column

##BDNF
BDNF2 <- bdnf %>% drop_na(methylierung)

# Summarize the number of NA values in the 'methylierung' column (commented out)
# sum(is.na(BDNF$methylierung))

#subset of BDNF2 with only the values "m148" "m066" "m111" "m058" of the column "CpG_Position"
bdnf_reduced <- BDNF2 %>%
  dplyr::select(methylierung, Group, Zeitpunkt, Standort, Geschlecht, Age_start, CogTel, id_umstrukturieren, CpG_position) %>%
  dplyr::filter(CpG_position %in% c("m148", "m066", "m111", "m058", "m039")) #maybe drop m039


# without NA
str(bdnf_reduced)

tau2 <- tau %>% drop_na(methylation)

psd952  <- psd95  %>% drop_na(methylation)
```


## BDNF
### Normality check
```{r}
# Descriptive statistics for methylierung
mean(bdnf_reduced$methylierung)
sd(bdnf_reduced$methylierung)
median(bdnf_reduced$methylierung)

# Histogram to check normality of methylierung
ggplot(bdnf_reduced, aes(x = methylierung)) +
  geom_histogram(binwidth = 0.005, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of BDNF Methylation Levels", x = "Methylation Levels", y = "Frequency") +
  theme_minimal()

# Another Histogram
hist(bdnf$methylierung, breaks = 30, main = "Methylation distribution", xlab = "Methylation")

# QQ plot
qqnorm(bdnf$methylierung)
qqline(bdnf$methylierung, col = "red")

# Shapiro-Wilk normality test
shapiro.test(bdnf$methylierung)

# Boxplot to visualize methylierung distribution
boxplot(bdnf_reduced$methylierung, main="Boxplot of BDNF Methylation Levels", ylab="Methylation Levels")
```

Methylation values are significantly non-normal
(see zero-inflated Poisson (ZIP) models)


### GLM
```{r}
bdnf_glm_nointeraction <- lme(
  fixed = methylierung ~ as.factor(Group) + as.factor(Zeitpunkt) + as.factor(Standort) + as.factor(Geschlecht) + Age_start + CogTel,
  random = ~ 1 | id_umstrukturieren,
  method = "ML",
  data = bdnf_reduced
)

summary(bdnf_glm_nointeraction)

res <- resid(bdnf_glm_nointeraction)
shapiro.test(res)
qqnorm(res)
qqline(res, col = "blue")
```

```{r}
bdnf_glm_time_nointeraction <- lme(
  fixed = methylierung ~ as.factor(Group) + as.factor(Zeitpunkt) + as.factor(Standort) + as.factor(Geschlecht) + Age_start + CogTel,
  random = ~ as.factor(Zeitpunkt) | id_umstrukturieren,
  method = "ML",
  data = bdnf_reduced
)

summary(bdnf_glm_time_nointeraction)

res_time <- resid(bdnf_glm_nointeraction)
shapiro.test(res_time)
qqnorm(res_time)
qqline(res_time, col = "blue")
```

also resuidulas are non-normal --> try logit transformation, nlme or glmmTMB below

### Model for group
```{r}
# Split the data by group
BDNF_controll = bdnf_reduced[bdnf_reduced$Group=='Control',]

BDNF_piano    = bdnf_reduced[bdnf_reduced$Group=='Piano',]


# Define a mixed model for the Piano group?????
bdnf_mixed_model_piano <- lme(fixed= methylierung ~ as.factor(Zeitpunkt),
                                  random=~1|id_umstrukturieren,
                                  method="ML",data=BDNF_piano)

bdnf_mixed_model_controll <- lme(fixed= methylierung ~ as.factor(Zeitpunkt),
                                  random=~1|id_umstrukturieren,
                                  method="ML",data=BDNF_controll)


# Summarize the Piano group model
summary(bdnf_mixed_model_piano)
summary(bdnf_mixed_model_controll)

```


### Interaction plot
```{r}

# Interaction plot showing methylation levels over time

plot_bdnf <- ggplot(bdnf_reduced, 
                    aes(x = Zeitpunkt, y = methylierung, group = 1)) +
  geom_line(stat = "summary", fun = mean) +
  geom_point(stat = "summary", fun = mean) +
  stat_summary(
    fun.data = mean_se,      # mean ± standard error
    geom = "errorbar",
    width = 0.1              # width of the error bar caps
  ) +
  theme_linedraw() +
  labs(
    title = "Interaction Plot of BDNF Methylation Levels over Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(plot_bdnf)
  
```

```{r}
# Interaction plot showing methylation levels by group and time
interaction_plot_bdnf <- ggplot(bdnf_reduced, aes(x = Zeitpunkt, y = methylierung, group = Group, colour = Group)) +
  geom_line(stat = "summary", fun = mean) +
  geom_point(stat = "summary", fun = mean) +
  theme_linedraw() +
  labs(
    title = "Interaction Plot of BDNF Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_bdnf)

# Interaction plot with error bars for BDNF
interaction_plot_bdnf <- ggplot(
  bdnf_reduced,
  aes(x = Zeitpunkt, y = methylierung, group = Group, colour = Group)
) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(
    fun.data = mean_se,      # mean ± standard error
    geom = "errorbar",
    width = 0.1              # width of the error bar caps
  ) +
  theme_linedraw() +
  labs(
    title = "Interaction Plot of BDNF Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_bdnf)


```

### Logit


```{r}
library(nlme)

# 1. Transform methylation
bdnf_reduced$Methyl_arcsin <- asin(sqrt(bdnf_reduced$methylierung))

# 2. Fit linear mixed-effects model
bdnf_lme <- lme(
  fixed = Methyl_arcsin ~ as.factor(Group) + as.factor(Zeitpunkt) +
            as.factor(Standort) + as.factor(Geschlecht) + Age_start + CogTel,
  random = ~1 | id_umstrukturieren,
  method = "ML",
  data = bdnf_reduced
)

# 3. Check results
summary(bdnf_lme)

qqnorm(resid(bdnf_lme))
qqline(resid(bdnf_lme), col = "blue")
shapiro.test(resid(bdnf_lme))
```


```{r}
# Logit transformation of methylierung
bdnf_log <- BDNF2 %>%
  mutate(
    methylierung_logit = log((methylierung + 1e-6) / (1 - methylierung + 1e-6))
  )

shapiro.test(bdnf_log$methylierung_logit)
hist(bdnf_log$methylierung_logit, breaks = 30, main = "Logit Transformed Methylation distribution", xlab = "Logit Transformed Methylation")
```

```{r}
# Define mixed models without interaction for logit transformed data
library(nlme)

bdnf_glm_logit_nointeraction <- lme(
  fixed = methylierung_logit ~ as.factor(Group) + as.factor(Zeitpunkt) + 
    as.factor(Standort) + as.factor(Geschlecht) + Age_start + CogTel,
  random = ~ 1 | id_umstrukturieren,
  method = "ML",
  data = bdnf_log
)

summary(bdnf_glm_logit_nointeraction)

res_log <- resid(bdnf_glm_logit_nointeraction)
shapiro.test(res_log)
qqnorm(res_log)
qqline(res_log, col = "blue")

```

```{r}
bdnf_glm_logit_time_nointeraction <- lme(
  fixed = methylierung_logit ~ as.factor(Group) + as.factor(Zeitpunkt) + 
    as.factor(Standort) + as.factor(Geschlecht) + Age_start + CogTel,
  random = ~ as.factor(Zeitpunkt) | id_umstrukturieren,
  method = "ML",
  data = bdnf_log
)

summary(bdnf_glm_logit_time_nointeraction)
```

```{r}
anova(bdnf_glm_logit_time_nointeraction, bdnf_glm_logit_nointeraction)

```


```{r}
library(nlme)

bdnf_glm_logit_var <- lme(
  fixed = methylierung_logit ~ as.factor(Group) + as.factor(Zeitpunkt) +
    as.factor(Standort) + as.factor(Geschlecht) + Age_start + CogTel,
  random = ~ 1 | id_umstrukturieren,
  weights = varPower(form = ~ fitted(.)),  # variance depends on fitted values
  method = "ML",
  data = bdnf_log
)

summary(bdnf_glm_logit_var)

qqnorm(resid(bdnf_glm_logit_var))
qqline(resid(bdnf_glm_logit_var))
```



```{r}
library(glmmTMB)

# Zero-inflated beta GLMM
model <- glmmTMB(
  methylierung ~ Group + Zeitpunkt + Standort + Geschlecht + Age_start + CogTel + (1|id_umstrukturieren),
  ziformula = ~1,          # models the probability of zero
  family = beta_family(link = "logit"),  # beta regression for values in (0,1)
  data = bdnf_reduced
)
summary(model)
```

## Tau

### Subset Data for relevant CpGs

```{r}
tau2 <- tau %>% drop_na(methylation)

#subset of tau2 with only the values "m331", "m337", "m310", "m315", "m323" of the column "CpG_Position"
tau_reduced <- tau2 %>%
  dplyr::select(methylation, Gruppe, Zeit, standort_aus_andere_quelle, Sex, Age_start, CogTel, id_umstruk, CpG_position)   %>%
  dplyr::filter(CpG_position %in% c("m331", "m337", "m310", "m315", "m323", "m306")) #maybe drop m306

# without NA
str(tau_reduced)
```


```{r}
# Define mixed models without interaction

tau_glm_nointeraction <- lme(
  fixed = methylation ~ as.factor(Gruppe) + as.factor(Zeit) + as.factor(standort_aus_andere_quelle) + as.factor(Sex) + Age_start + as.numeric(CogTel),
  random = ~ 1 | id_umstruk,
  method = "ML",
  data = tau_reduced
)

summary(tau_glm_nointeraction)
```


```{r}
tau_glm_interaction <- lmer(
  formula = methylation ~ as.factor(Gruppe) + as.factor(Zeit) + as.factor(standort_aus_andere_quelle) + as.factor(Sex) + Age_start +
                         as.numeric(CogTel)  +
                        (1 | CpG_position),
  data = tau_reduced,
  REML = TRUE
)

summary(tau_glm_nointeraction)
summary(tau_glm_interaction)

library(car)
vif(tau_glm_interaction) # because of 


```

### Interaction plot

```{r}
# Plot showing methylation levels over time (Tau)
interaction_plot_tau <- ggplot(
  tau_reduced,
  aes(x = Zeit, y = methylation, group = 1)  # make sure variable names match
) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point", size = 2) +
  theme_linedraw() +
  labs(
    title = "Plot of Tau Methylation Levels over Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_tau)
```


```{r}
# Interaction plot showing methylation levels by group and time (Tau)
interaction_plot_tau <- ggplot(
  tau_reduced,
  aes(x = Zeit, y = methylation, group = Gruppe, colour = Gruppe)  # make sure variable names match
) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(
    fun.data = mean_se,      # mean ± standard error
    geom = "errorbar",
    width = 0.1              # width of the error bar caps
  ) +
  theme_linedraw() +
  labs(
    title = "Interaction Plot of Tau Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_tau)



```


## PSD95 


```{r}
psd952  <- psd95  %>% drop_na(methylation)
#subset of BDNF2 with only the values "m148" "m066" "m111" "m058" of the column "CpG_Position"
psd95_reduced <- psd952 %>%
  dplyr::select(methylation, Gruppe, Zeitpunkt, standort_aus_andere_quelle, Sex, Age_start, CogTel, id_umstrukturiert, CpG_Position) %>%
  dplyr::filter(CpG_Position %in% c("p95", "p60", "p23")) #maybe drop m039

# without NA
str(psd95_reduced)


```

```{r}
# Define mixed models without interaction
psd95_glm_nointeraction <- lme(
  fixed = methylation ~ as.factor(Gruppe) + as.factor(Zeitpunkt) + as.factor(standort_aus_andere_quelle) + as.factor(Sex) + Age_start + as.numeric(CogTel),
  random = ~ 1 | id_umstrukturiert,
  method = "ML",
  data = psd952
)

summary(psd95_glm_nointeraction)

psd95_glm_interaction <- lmer(
  formula = methylation ~ as.factor(Gruppe) + as.factor(Zeitpunkt) + as.factor(standort_aus_andere_quelle) + as.factor(Sex) + Age_start + as.numeric(CogTel) + Age_start * as.factor(Sex) +
                        (1 | CpG_Position),
  data = psd952,
  REML = TRUE
)


summary(psd95_glm_interaction)
```

### Interaction plot

```{r}
# Interaction plot showing methylation levels by group and time (Tau)
interaction_plot_tau <- ggplot(
  psd95_reduced,
  aes(x = Zeitpunkt, y = methylation, group = 1)  # make sure variable names match
) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point", size = 2) +
  theme_linedraw() +
  labs(
    title = "Interaction Plot of Tau Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_tau)
```


#### Comparison between groups

```{r}
library(ggplot2)

# Interaction plot for PSD95 with mean ± SE
interaction_plot_psd95 <- ggplot(
  psd95_reduced,
  aes(x = Zeitpunkt, y = methylation, group = Gruppe, colour = Gruppe)
) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(
    fun.data = mean_se,   # mean ± standard error
    geom = "errorbar",
    width = 0.1           # width of the error bar caps
  ) +
  theme_linedraw() +
  labs(
    title = "Interaction Plot of PSD95 Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_psd95)

```

#### Comparison between sexes

```{r}
library(ggplot2)

# Interaction plot for PSD95 with mean ± SE
interaction_plot_psd95 <- ggplot(
  psd95_reduced,
  aes(x = Zeitpunkt, y = methylation, group = Sex, colour = Sex)
) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun = mean, geom = "point", size = 2) +
  stat_summary(
    fun.data = mean_se,   # mean ± standard error
    geom = "errorbar",
    width = 0.1           # width of the error bar caps
  ) +
  theme_linedraw() +
  labs(
    title = "Interaction Plot of PSD95 Methylation Levels by Sex and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_psd95)

```


