---
title: "Principal Component Analysis (PCA) using Covariance Matrix"
author: "Jonas J. R. Koberschinski"
date: "16-10-2025"

format: html
---

## Data import

```{r}
#| echo: true
#| include: false

#data set with all CpGs in columns for EDA and PCA
library(tidyverse)

# load data
 ## BDNF
bdnf <- as_tibble(read.table('/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/bdnf.txt', 
                   sep = "\t", 
                   header = TRUE, 
                   row.names = NULL, 
                   quote = ""))
bdnf[1:5,]

 ##Tau
tau <- as_tibble(read.table('/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/tau.txt', 
                   sep = "\t", 
                   header = TRUE, 
                   row.names = NULL, 
                   quote = ""))
tau[1:5,]

 ## PSD95
psd95 <- as_tibble(read.table('/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/psd95.txt', 
                   sep = "\t", 
                   header = TRUE, 
                   row.names = NULL, 
                   quote = ""))
psd95[1:5,]
```

Raw (unscaled) CpG methylation values. --\>PCA is therefore using the **covariance matrix** of those CpGs.

Chat: "You can interpret large‐magnitude loadings as CpGs most responsible for differences among your samples." Watch out!

## BDNF
### Data selection
```{r}
cpg_bdnf <- bdnf[11:20]
```
### PCA computation
```{r}
pca_bdnf <- prcomp(na.omit(cpg_bdnf), scale = FALSE) # Raw (unscaled) methylation values --> PCA with covariance matrix of CpGs
# prcomp() centers by default!
#For CpG methylation data (which often varies by scale and variance across loci): center = TRUE and scale = TRUE are recommended (your current setting is correct).

# Extract principal component scores
pc_scores_bdnf <- as.data.frame(pca_bdnf$x)
```
### Normality diagnostics
```{r}
# Perform Shapiro-Wilk test for normality on each principal component
shapiro_p_bdnf <- sapply(pc_scores_bdnf, function(z) {
  n <- length(z)
  if (n >= 3 && n <= 5000) shapiro.test(z)$p.value else NA_real_
})
shapiro_tab_bdnf <- data.frame(PC = names(shapiro_p_bdnf),
                          p_value = as.numeric(shapiro_p_bdnf),
                          BH_adj = p.adjust(shapiro_p_bdnf, method = "BH"))
print(shapiro_tab_bdnf)

# Quick QQ-plots for the first two PCs
par(mfrow = c(1, 2))
qqnorm(pc_scores_bdnf[[1]], main = "QQ plot: PC1"); qqline(pc_scores_bdnf[[1]])
qqnorm(pc_scores_bdnf[[2]], main = "QQ plot: PC2"); qqline(pc_scores_bdnf[[2]])
par(mfrow = c(1, 1))

```

```{r}
# Extract principal component scores
pc_scores <- as.data.frame(pca_bdnf$x)
```



### Visualizations
```{r}
## plot pc1 and pc2
plot(pca_bdnf$x[,1], pca_bdnf$x[,2])# pca_bdnf$x holds the principal component scores (i.e., the transformed sample coordinates in PC space).
# Each column corresponds to a principal component, and each row corresponds to a sample.
# plot the first principal component (PC1) against the second principal component (PC2) to visualize the distribution of samples in the space defined by these two components.



## make a scree plot
pca_bdnf_var <- pca_bdnf$sdev^2
# The standard deviations of the principal components are stored in the sdev attribute of the pca_bdnf object.
# To understand the amount of variance explained by each principal component, you can square these standard deviations
# to obtain the variance explained by each component.


pca_bdnf_var_per <- round(pca_bdnf_var/sum(pca_bdnf_var)*100, 1)
# To express the variance explained by each principal component as a percentage of the total variance,
# you can divide each component's variance by the total variance (sum of all variances)
# and multiply by 100. The round function is used to round the percentages to one decimal place for better readability.

barplot(pca_bdnf_var_per, main="BDNF Scree Plot", xlab="Principal Component", ylab="Percent Variation")
# The barplot (Scree Plot) shows how much information each PC captures — usually you look for the “elbow point” to decide how many PCs to keep.
# In this plot, each bar represents a principal component, and its height corresponds to the percentage of total variance explained by that component.
# The x-axis represents the principal components (PC1, PC2, etc.), and the y-axis represents the percentage of variance explained by each component.
# The scree plot helps to visualize the relative importance of each principal component in terms of the variance it captures from the original data.


## now make a fancy looking plot that shows the PCs and the variation:
library(ggplot2)


pca_bdnf_data <- as.data.frame(pca_bdnf$x)
pca_bdnf_data$ID <- rownames(pca_bdnf)

str(pca_bdnf_data)

ggplot(pca_bdnf_data, aes(PC1, PC2)) +
  geom_point() +
  xlab(paste("PC1 - ", pca_bdnf_var_per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca_bdnf_var_per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("BDNF PCA Graph")


bdnf_loading_scores <- pca_bdnf$rotation[,1]
bdnf_methylation_scores <- abs(bdnf_loading_scores) # what to do of bdnf_methylation_scores?
```


```{r}
library(ggfortify)
pca_bdnf_res <- prcomp(na.omit(bdnf[c(11:20)]), scale = FALSE)

autoplot(pca_bdnf_res, data = na.omit(bdnf[c(1:20)]), colour = 'Zeitpunkt',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         frame = TRUE, frame.type = "norm"
)

autoplot(pca_bdnf_res, data = na.omit(bdnf[c(1:20)]), colour = 'Gruppe', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         frame = TRUE, frame.type = "norm"
)
autoplot(pca_bdnf_res, data = na.omit(bdnf[c(1:22)]), colour = 'Sex', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         frame = TRUE, frame.type = "norm"
)
```



```{r}
#| label: fig-bdnf-pca-group-timepoint
#| fig-cap: "BDNF PCA biplot. Points colored by Gruppe, faceted by Zeitpunkt; loadings shown as arrows with labels; 95% normal ellipses per group."
#| fig-width: 15
#| fig-height: 6

library(ggfortify)
library(ggplot2)

bdnf$Age_start <- as.numeric(bdnf$Age_start)
bdnf$CogTel    <- as.numeric(bdnf$CogTel)

autoplot(
  pca_bdnf_res,
  data = na.omit(bdnf[c(1:22)]),
  colour = "Gruppe",
  loadings = TRUE,
  loadings.colour = "steelblue",
  loadings.label = TRUE,
  loadings.label.colour = "firebrick",
  loadings.label.size = 3,
  frame = TRUE,
  frame.type = "norm",
  frame.level = 0.95,
  frame.alpha = 0.15
) +
  scale_size_continuous(range = c(2, 6)) +
  facet_wrap(~Zeitpunkt) +
  theme_classic() +
  labs(title = "BDNF PCA by Group (faceted by timepoint)")


```
```{r}

library(ggfortify)
library(ggplot2)

# Ensure numeric variables for size
bdnf$Age_start <- as.numeric(bdnf$Age_start)
bdnf$CogTel <- as.numeric(bdnf$CogTel)

autoplot(
  pca_bdnf_res,
  data = na.omit(bdnf[c(1:22)]),
  colour = "Gruppe",            # point color by group
  loadings = TRUE,
  loadings.colour = "steelblue",        # <-- arrows
  loadings.label = TRUE,
  loadings.label.colour = "firebrick",  # <-- text labels
  loadings.label.size = 3,
    frame = TRUE,                 # << draw group frames
  frame.type = "norm",          # "norm", "t", or "convex"
  frame.level = 0.95,           # 95% ellipse
  frame.alpha = 0.15            # transparency for overlap
) +
  scale_size_continuous(range = c(2, 6)) +      # min/max point size
  facet_wrap(~Zeitpunkt) +                       # separate panels for each timepoint
  theme_classic() +
  ggtitle("BDNF PCA: Color by Group Facet by Timepoint")

```

```{r}
library(dplyr)
library(ggplot2)
library(ggrepel)

## --- 1) Prepare matched data and run PCA ---
cpg_cols <- which(names(bdnf) == "m148"):which(names(bdnf) == "p020")

X <- bdnf[, cpg_cols] |>
  mutate(across(everything(), ~ suppressWarnings(as.numeric(.))))
keep <- complete.cases(X, bdnf[, c("Gruppe","Sex","Zeitpunkt","Age_start")])
X    <- as.matrix(X[keep, , drop = FALSE])
meta <- bdnf[keep, c("Gruppe","Sex","Zeitpunkt","Age_start")]

pca <- prcomp(X, center = TRUE, scale. = TRUE)

## --- 2) Scores and (shorter) top loadings ---
scores <- as.data.frame(pca$x[, 1:2]) |>
  setNames(c("PC1","PC2")) |>
  bind_cols(meta)

# top 8 loadings, scaled shorter so they don't swamp points
load <- as.data.frame(pca$rotation[, 1:2])
load$CpG <- rownames(load)
load$mag <- sqrt(load$PC1^2 + load$PC2^2)
load_top <- load |> dplyr::slice_max(mag, n = 8)

rng    <- apply(scores[, c("PC1","PC2")], 2, \(z) diff(range(z)))
arrow_scale <- 0.45 * min(rng) / max(abs(as.matrix(load_top[, c("PC1","PC2")])))

load_top_scaled <- transform(load_top,
                             PC1 = PC1 * arrow_scale,
                             PC2 = PC2 * arrow_scale)

## --- 3) Plot: arrows first (light), then points (bigger), then labels ---
ggplot() +
  # arrows BEHIND points
  geom_segment(
    data = load_top_scaled,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    color = "steelblue3", linewidth = 0.5, alpha = 0.5,
    arrow = arrow(length = unit(5, "pt"))
  ) +
  # points ON TOP
  geom_point(
    data = scores,
    aes(PC1, PC2, color = Gruppe, shape = Sex, size = Age_start),
    alpha = 0.85, stroke = 0.3
  ) +
  # loadings labels (repelled, small)
  geom_text_repel(
    data = load_top_scaled,
    aes(PC1, PC2, label = CpG),
    size = 3, color = "firebrick", max.overlaps = 50
  ) +
  stat_ellipse(
    data = scores,
    aes(PC1, PC2, color = Gruppe),
    level = 0.68, linewidth = 0.6, type = "norm", show.legend = FALSE
  ) +
  facet_wrap(~ Zeitpunkt) +
  coord_equal() +
  scale_color_brewer(palette = "Dark2") +
  scale_shape_manual(values = c(16, 17)) +
  scale_size_continuous(range = c(1.8, 4.2)) +
  theme_classic() +
  labs(
    title = "BDNF PCA: color=Group, shape=Sex, size=Age, facets=Zeitpunkt",
    x = paste0("PC1 (", scales::percent(summary(pca)$importance[2, 1]), ")"),
    y = paste0("PC2 (", scales::percent(summary(pca)$importance[2, 2]), ")")
  )


```



### Ranking BDNF loadings

```{r}
# Rank CpGs by influence on PC1
ranked_bdnf_cpgs <- sort(abs(bdnf_loading_scores), decreasing =
 TRUE)
head(ranked_bdnf_cpgs)
```

| Rank | CpG | |Loading| |
|------|------|-----------:|
| 1 | m148 | 0.612 |
| 2 | m066 | 0.453 |
| 3 | m111 | 0.431 |
| 4 | m058 | 0.422 |
| - | ---- | ----- |
| 5 | m039 | 0.215 |
| 6 | m035 | 0.076 |
| 7 | m087 | 0.053 |
| 8 | m024 | 0.029 |
| 9 | p020 | 0.022 |
| 10 | p018 | 0.013 |


### Visualize ranked loadings

```{r}
plot(ranked_bdnf_cpgs, type = "b", main = "Absolute PC1 Loadings (BDNF CpGs)",
     ylab = "|Loading|", xlab = "Ranked CpGs")
abline(v = 4.5, col = "red", lty = 2, lwd = 2)
```


Die ersten 4 CpG scheinen die wesentlichen zu sein, wie in den Boxplots zu erkennen!

## Tau

```{r}
cpg_tau <- tau[c(9:57)]
pca_tau <- prcomp(na.omit(cpg_tau), scale = FALSE)
# Extract principal component scores
pc_scores_tau <- as.data.frame(pca_tau$x)
```

```{r}
## Normality diagnostics
# Perform Shapiro-Wilk test for normality on each principal component
shapiro_p_tau <- sapply(pc_scores_tau, function(z) {
  n <- length(z)
  if (n >= 3 && n <= 5000) shapiro.test(z)$p.value else NA_real_
})
shapiro_tab_tau <- data.frame(PC = names(shapiro_p_tau),
                          p_value = as.numeric(shapiro_p_tau),
                          BH_adj = p.adjust(shapiro_p_tau, method = "BH"))
print(shapiro_tab_tau)

# Quick QQ-plots for the first two PCs
par(mfrow = c(1, 2))
qqnorm(pc_scores_tau[[1]], main = "QQ plot: PC1"); qqline(pc_scores_tau[[1]])
qqnorm(pc_scores_tau[[2]], main = "QQ plot: PC2"); qqline(pc_scores_tau[[2]])
par(mfrow = c(1, 1))

```


```{r}
## plot pc1 and pc2
plot(pca_tau$x[,1], pca_tau$x[,2])

## make a scree plot
pca_tau_var <- pca_tau$sdev^2
pca_tau_var_per <- round(pca_tau_var/sum(pca_tau_var)*100, 1)

barplot(pca_tau_var_per, main="Tau Scree Plot", xlab="Principal Component", ylab="Percent Variation")


## now make a fancy looking plot that shows the PCs and the variation:
library(ggplot2)

pca_tau_data <- as.data.frame(pca_tau$x)
pca_tau_data$ID <- rownames(pca_tau)

str(pca_tau_data)

ggplot(pca_tau_data, aes(PC1, PC2)) +
  geom_point() +
  xlab(paste("PC1 - ", pca_tau_var_per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca_tau_var_per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("Tau PCA Graph")


tau_loading_scores <- pca_tau$rotation[,1]
tau_methylation_scores <- abs(tau_loading_scores) # what to do of tau_methylation_scores?
```


```{r}
library(ggfortify)
pca_tau_res <- prcomp(na.omit(tau[c(9:57)]), scale = FALSE)

autoplot(pca_tau_res, data = na.omit(tau[c(1:57)]), colour = 'Zeit',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         frame = TRUE, frame.type = "norm"
         )

autoplot(pca_tau_res, data = na.omit(tau[c(1:57)]), colour = 'Gruppe', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         frame = TRUE, frame.type = "norm"
)

autoplot(pca_tau_res, data = na.omit(tau[c(1:57,357)]), colour = 'Sex', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         frame = TRUE, frame.type = "norm"
)

```

```{r}
autoplot(
  pca_tau_res,
  data = na.omit(tau[c(1:57,357)]),
  colour = "Gruppe",            # point color by group
  shape  = "Sex",               # point shape by sex
  loadings = TRUE,
  loadings.colour = "steelblue",        # <-- arrows
  loadings.label = TRUE,
  loadings.label.colour = "firebrick",  # <-- text labels
  loadings.label.size = 3,
  frame = TRUE,                 # << draw group frames
  frame.type = "norm",          # "norm", "t", or "convex"
  frame.level = 0.95,           # 95% ellipse
  frame.alpha = 0.15            # transparency for overlap
) +
  scale_shape_manual(values = c(16, 17)) +      # custom shapes for Sex
  scale_size_continuous(range = c(2, 6)) +      # min/max point size
  facet_wrap(~Zeit) +                       # separate panels for each timepoint
  theme_classic() +
  ggtitle("Tau PCA: Color by Group, Shape by Sex, Size by Age, Facet by Timepoint")


```


### Ranking Tau loadings
```{r}
# Rank CpGs by influence on PC1
ranked_tau_cpgs <- sort(abs(tau_loading_scores), decreasing = TRUE)
head(ranked_tau_cpgs)

```

Die ersten 10 CpG scheinen die wesentlichen zu sein, wie in den Boxplots zu erkennen! m277 = 0.1086290423\
m210 = 0.0785399456

| Rank | CpG  |        |
|------|------|-------:|
| 1    | m331 | 0.4438 |
| 2    | m337 | 0.4236 |
| 3    | m310 | 0.3856 |
| 4    | m315 | 0.3843 |
| 5    | m323 | 0.3771 |
| -    | ---- | ------ |
| 6    | m306 | 0.2459 |
| 7    | m219 | 0.1652 |
| 8    | m300 | 0.1489 |
| 9    | m293 | 0.1295 |
| 10   | m207 | 0.1246 |
| 11   | m249 | 0.1151 |
| 12   | m277 | 0.1086 |
| 13   | m210 | 0.0785 |



### Visualize ranked loadings

There are two clear gaps:

Between m323 (0.377) and m306 (0.246) → Δ = 0.131

Between m277 (0.109) and m210 (0.079) → Δ = 0.030

```{r}
plot(ranked_tau_cpgs, type = "b", main = "Absolute PC1 Loadings (Tau CpGs)",
     ylab = "|Loading|", xlab = "Ranked CpGs")
abline(v = 5.5, col = "red", lty = 2, lwd = 2)

```




## PSD95

```{r}
cpg_psd95 <- psd95[10:54]
pca_psd95 <- prcomp(na.omit(cpg_psd95), scale = FALSE)
pc_scores_psd95 <- as.data.frame(pca_psd95$x)
```

```{r}
## Normality diagnostics
# Perform Shapiro-Wilk test for normality on each principal component
shapiro_p_psd95 <- sapply(pc_scores_psd95, function(z) {
  n <- length(z)
  if (n >= 3 && n <= 5000) shapiro.test(z)$p.value else NA_real_
})
shapiro_tab_psd95 <- data.frame(PC = names(shapiro_p_psd95),
                          p_value = as.numeric(shapiro_p_psd95),
                          BH_adj = p.adjust(shapiro_p_psd95, method = "BH"))
print(shapiro_tab_psd95)

# Quick QQ-plots for the first two PCs
par(mfrow = c(1, 2))
qqnorm(pc_scores_psd95[[1]], main = "QQ plot: PC1"); qqline(pc_scores_psd95[[1]])
qqnorm(pc_scores_psd95[[2]], main = "QQ plot: PC2"); qqline(pc_scores_psd95[[2]])
par(mfrow = c(1, 1))

```
```{r}
## plot pc1 and pc2
plot(pca_psd95$x[,1], pca_psd95$x[,2])

## make a scree plot
pca_psd95_var <- pca_psd95$sdev^2
pca_psd95_var_per <- round(pca_psd95_var/sum(pca_psd95_var)*100, 1)

barplot(pca_psd95_var_per, main="PSD95 Scree Plot", xlab="Principal Component", ylab="Percent Variation")


## now make a fancy looking plot that shows the PCs and the variation:
library(ggplot2)


pca_psd95_data <- as.data.frame(pca_psd95$x)
pca_psd95_data$ID <- rownames(pca_psd95)

str(pca_psd95_data)

ggplot(pca_psd95_data, aes(PC1, PC2)) +
  geom_point() +
  xlab(paste("PC1 - ", pca_psd95_var_per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca_psd95_var_per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("PSD95 PCA Graph")


psd95_loading_scores <- pca_psd95$rotation[,1]
psd95_methylation_scores <- abs(psd95_loading_scores) # what to do of tau_methylation_scores?
```


```{r}
library(ggfortify)
pca_psd95_res <- prcomp(na.omit(psd95[c(10:54)]), scale = FALSE)

autoplot(pca_psd95_res, data = na.omit(psd95[c(1:54)]), colour = 'Zeitpunkt',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         frame = TRUE, frame.type = "norm"
         )

autoplot(pca_psd95_res, data = na.omit(psd95[c(1:54)]), colour = 'Gruppe', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         frame = TRUE, frame.type = "norm"
         )

autoplot(pca_psd95_res, data = na.omit(psd95[c(1:54,353)]), colour = 'Sex', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3,
         frame = TRUE, frame.type = "norm"
)
```


```{r}
autoplot(
  pca_psd95_res,
  data = na.omit(psd95[c(1:57,353)]),
  colour = "Gruppe",            # point color by group
  loadings = TRUE,
  loadings.colour = "steelblue",        # <-- arrows
  loadings.label = TRUE,
  loadings.label.colour = "firebrick",  # <-- text labels
  loadings.label.size = 3,
  frame = TRUE,                 # << draw group frames
  frame.type = "norm",          # "norm", "t", or "convex"
  frame.level = 0.95,           # 95% ellipse
  frame.alpha = 0.15            # transparency for overlap
) +
  scale_size_continuous(range = c(2, 6)) +      # min/max point size
  facet_wrap(~Zeitpunkt) +                       # separate panels for each timepoint
  theme_classic() +
  ggtitle("PSD95 PCA: Color by Group, Shape by Sex, Size by Age, Facet by Timepoint")


```

### Ranking PSD95 loadings
```{r}
# Rank CpGs by influence on PC1
ranked_psd95_cpgs <- sort(abs(psd95_loading_scores), decreasing = TRUE)
head(ranked_psd95_cpgs)
```

| Rank | CpG | |Loading| |
|------|------|-----------:|
| 1 | p95 | 0.664 |
| 2 | p60 | 0.484 |
| 3 | p23 | 0.445 |
| - | --- | ----- |
| 4 | m35 | 0.153 |
| 5 | m28 | 0.103 |
| 6 | m418 | 0.097 |
| 7 | m24 | 0.094 |
| 8 | p91 | 0.094 |
| 9 | m400 | 0.086 |
| 10 | m11 | 0.078 |

ChatGPT:
What to report

“PC1 loadings show a sharp elbow after 3 CpGs (figure). We therefore analyze the top-3 CpGs that explain the majority of variability and provide sensitivity analyses using (i) all CpGs and (ii) a PC1-based composite.”

### Visualize ranked loadings

```{r}
#| label: fig-psd95-loadings
#| fig-cap: "Absolute PC1 Loadings (PSD95 CpGs)"
plot(ranked_psd95_cpgs, type = "b", main = "Absolute PC1 Loadings (PSD95 CpGs)",
     ylab = "|Loading|", xlab = "Ranked CpGs")
abline(v = 3.5, col = "red", lty = 2, lwd = 2)
```
Die ersten 3 CpG scheinen die wesentlichen zu sein, wie in den Boxplots zu erkennen!








## all for manuscript figures

### biplot
```{r}
#| label: fig-pca-genes
#| fig-cap: "PCA biplots for BDNF, Tau, and PSD95. Points colored by group; faceted by timepoint; loadings shown; 95% normal ellipses."
#| fig-width: 16
#| fig-height: 14

library(ggfortify)
library(ggplot2)
library(patchwork)

p_bdnf <- autoplot(
  pca_bdnf_res,
  data = na.omit(bdnf[1:22]),
  colour = "Gruppe",
  loadings = TRUE,
  loadings.colour = "steelblue",
  loadings.label = TRUE,
  loadings.label.colour = "firebrick",
  loadings.label.size = 3,
  frame = TRUE,
  frame.type = "norm",
  frame.level = 0.95,
  frame.alpha = 0.15
) +
  facet_wrap(~Zeitpunkt) +
  theme_classic() +
  labs(title = "BDNF")

p_tau <- autoplot(
  pca_tau_res,
  data = na.omit(tau[1:57]),
  colour = "Gruppe",
  loadings = TRUE,
  loadings.colour = "steelblue",
  loadings.label = TRUE,
  loadings.label.colour = "firebrick",
  loadings.label.size = 3,
  frame = TRUE,
  frame.type = "norm",
  frame.level = 0.95,
  frame.alpha = 0.15
) +
  facet_wrap(~Zeit) +   # keep your column name here
  theme_classic() +
  labs(title = "Tau")

p_psd95 <- autoplot(
  pca_psd95_res,
  data = na.omit(psd95[1:54]),
  colour = "Gruppe",
  loadings = TRUE,
  loadings.colour = "steelblue",
  loadings.label = TRUE,
  loadings.label.colour = "firebrick",
  loadings.label.size = 3,
  frame = TRUE,
  frame.type = "norm",
  frame.level = 0.95,
  frame.alpha = 0.15
) +
  facet_wrap(~Zeitpunkt) +
  theme_classic() +
  labs(title = "PSD95")

# combine into one figure (3 rows)
(p_bdnf / p_tau / p_psd95) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

```

### Scree plots

```{r}
#| label: fig-scree-all
#| fig-cap: "Scree plots showing percent variance explained by principal components for BDNF, Tau, and PSD95."
#| fig-width: 10
#| fig-height: 11

op <- par(no.readonly = TRUE)
par(mfrow = c(3, 1), mar = c(4.2, 4.5, 3, 1))

barplot(pca_bdnf_var_per,
        main = "BDNF Scree Plot",
        xlab = "Principal Component",
        ylab = "Percent Variation")

barplot(pca_tau_var_per,
        main = "Tau Scree Plot",
        xlab = "Principal Component",
        ylab = "Percent Variation")

barplot(pca_psd95_var_per,
        main = "PSD95 Scree Plot",
        xlab = "Principal Component",
        ylab = "Percent Variation")

par(op)

```


### loadings



```{r}
#| label: fig-pc1-loadings
#| fig-cap: "Absolute PC1 loadings ranked for BDNF, Tau, and PSD95 CpGs. Red dashed line indicates the cutoff."
#| fig-width: 10
#| fig-height: 9

op <- par(no.readonly = TRUE)

par(
  mfrow = c(3, 1),          # 3 rows, 1 column
  mar   = c(4.2, 4.2, 3, 1) # margins: bottom, left, top, right
)

plot(ranked_bdnf_cpgs, type = "b",
     main = "Absolute PC1 Loadings (BDNF CpGs)",
     ylab = "|Loading|", xlab = "Ranked CpGs")
abline(v = 4.5, col = "red", lty = 2, lwd = 2)

plot(ranked_tau_cpgs, type = "b",
     main = "Absolute PC1 Loadings (Tau CpGs)",
     ylab = "|Loading|", xlab = "Ranked CpGs")
abline(v = 5.5, col = "red", lty = 2, lwd = 2)

plot(ranked_psd95_cpgs, type = "b",
     main = "Absolute PC1 Loadings (PSD95 CpGs)",
     ylab = "|Loading|", xlab = "Ranked CpGs")
abline(v = 3.5, col = "red", lty = 2, lwd = 2)

par(op)

```


