---
title: "Principal Component Analysis"
author: "Jonas J.R. Koberschinski"
date: "16-10-2025"
format: html
---

## Data import

```{r}
#| echo: true
#| include: false

#data set with all CpGs in columns for EDA and PCA
library(tidyverse)

# load data
 ## BDNF
bdnf <- as_tibble(read.table('/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/bdnf.txt', 
                   sep = "\t", 
                   header = TRUE, 
                   row.names = NULL, 
                   quote = ""))
bdnf[1:5,]

 ##Tau
tau <- as_tibble(read.table('/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/tau.txt', 
                   sep = "\t", 
                   header = TRUE, 
                   row.names = NULL, 
                   quote = ""))
tau[1:5,]

 ## PSD95
psd95 <- as_tibble(read.table('/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/psd95.txt', 
                   sep = "\t", 
                   header = TRUE, 
                   row.names = NULL, 
                   quote = ""))
psd95[1:5,]
```

Raw (unscaled) CpG methylation values. --\>PCA is therefore using the **covariance matrix** of those CpGs.

Chat: "You can interpret large‐magnitude loadings as CpGs most responsible for differences among your samples." Watch out!

## BDNF

```{r}
cpg_bdnf <- bdnf[11:20]

pca_bdnf <- prcomp(na.omit(cpg_bdnf), scale = FALSE) # Raw (unscaled) methylation values --> PCA with covariance matrix of CpGs
# prcomp() centers by default!
#For CpG methylation data (which often varies by scale and variance across loci): center = TRUE and scale = TRUE are recommended (your current setting is correct).






## plot pc1 and pc2
plot(pca_bdnf$x[,1], pca_bdnf$x[,2])# pca_bdnf$x holds the principal component scores (i.e., the transformed sample coordinates in PC space).
# Each column corresponds to a principal component, and each row corresponds to a sample.
# plot the first principal component (PC1) against the second principal component (PC2) to visualize the distribution of samples in the space defined by these two components.



## make a scree plot
pca_bdnf_var <- pca_bdnf$sdev^2
# The standard deviations of the principal components are stored in the sdev attribute of the pca_bdnf object.
# To understand the amount of variance explained by each principal component, you can square these standard deviations
# to obtain the variance explained by each component.


pca_bdnf_var_per <- round(pca_bdnf_var/sum(pca_bdnf_var)*100, 1)
# To express the variance explained by each principal component as a percentage of the total variance,
# you can divide each component's variance by the total variance (sum of all variances)
# and multiply by 100. The round function is used to round the percentages to one decimal place for better readability.

barplot(pca_bdnf_var_per, main="BDNF Scree Plot", xlab="Principal Component", ylab="Percent Variation")
# The barplot (Scree Plot) shows how much information each PC captures — usually you look for the “elbow point” to decide how many PCs to keep.
# In this plot, each bar represents a principal component, and its height corresponds to the percentage of total variance explained by that component.
# The x-axis represents the principal components (PC1, PC2, etc.), and the y-axis represents the percentage of variance explained by each component.
# The scree plot helps to visualize the relative importance of each principal component in terms of the variance it captures from the original data.


## now make a fancy looking plot that shows the PCs and the variation:
library(ggplot2)


pca_bdnf_data <- as.data.frame(pca_bdnf$x)
pca_bdnf_data$ID <- rownames(pca_bdnf)

str(pca_bdnf_data)

ggplot(pca_bdnf_data, aes(PC1, PC2)) +
  geom_point() +
  xlab(paste("PC1 - ", pca_bdnf_var_per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca_bdnf_var_per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("BDNF PCA Graph")


bdnf_loading_scores <- pca_bdnf$rotation[,1]
bdnf_methylation_scores <- abs(bdnf_loading_scores) # what to do of bdnf_methylation_scores?
  
library(ggfortify)
pca_bdnf_res <- prcomp(na.omit(bdnf[c(11:20)]), scale = FALSE)

autoplot(pca_bdnf_res, data = na.omit(bdnf[c(1:20)]), colour = 'Zeitpunkt',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)
autoplot(pca_bdnf_res, data = na.omit(bdnf[c(1:20)]), colour = 'Gruppe', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)
autoplot(pca_bdnf_res, data = na.omit(bdnf[c(1:22)]), colour = 'Sex', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)
```



```{r}
library(ggfortify)
library(ggplot2)

# Ensure numeric variables for size
bdnf$Age_start <- as.numeric(bdnf$Age_start)
bdnf$CogTel <- as.numeric(bdnf$CogTel)

autoplot(
  pca_bdnf_res,
  data = na.omit(bdnf[c(1:22)]),
  colour = 'Gruppe',        # group coloring
  shape = 'Sex',            # shape by Sex
  loadings = TRUE,
  loadings.colour = 'blue',
  loadings.label = TRUE,
  loadings.label.size = 3
) +
  scale_shape_manual(values = c(16, 17)) +      # custom shapes for Sex
  scale_size_continuous(range = c(2, 6)) +      # min/max point size
  facet_wrap(~Zeitpunkt) +                       # separate panels for each timepoint
  theme_classic() +
  ggtitle("BDNF PCA: Color by Group, Shape by Sex, Size by Age, Facet by Timepoint")

```




### Ranking BDNF loadings
```{r}
# Rank CpGs by influence on PC1
ranked_bdnf_cpgs <- sort(abs(bdnf_loading_scores), decreasing =
 TRUE)
head(ranked_bdnf_cpgs)
```

| Rank | CpG | |Loading| |
|------|------|-----------:|
| 1 | m148 | 0.612 |
| 2 | m066 | 0.453 |
| 3 | m111 | 0.431 |
| 4 | m058 | 0.422 |
| - | ---- | ----- |
| 5 | m039 | 0.215 |
| 6 | m035 | 0.076 |
| 7 | m087 | 0.053 |
| 8 | m024 | 0.029 |
| 9 | p020 | 0.022 |
| 10 | p018 | 0.013 |


### Visualize ranked loadings
```{r}
plot(ranked_bdnf_cpgs, type = "b", main = "Absolute PC1 Loadings (BDNF CpGs)",
     ylab = "|Loading|", xlab = "Ranked CpGs")
abline(v = 4.5, col = "red", lty = 2, lwd = 2)
```


Die ersten 4 CpG scheinen die wesentlichen zu sein, wie in den Boxplots zu erkennen!

## Tau

```{r}
cpg_tau <- tau[c(9:57)]
pca_tau <- prcomp(na.omit(cpg_tau), scale = FALSE)

## plot pc1 and pc2
plot(pca_tau$x[,1], pca_tau$x[,2])

## make a scree plot
pca_tau_var <- pca_tau$sdev^2
pca_tau_var_per <- round(pca_tau_var/sum(pca_tau_var)*100, 1)

barplot(pca_tau_var_per, main="Tau Scree Plot", xlab="Principal Component", ylab="Percent Variation")


## now make a fancy looking plot that shows the PCs and the variation:
library(ggplot2)

pca_tau_data <- as.data.frame(pca_tau$x)
pca_tau_data$ID <- rownames(pca_tau)

str(pca_tau_data)

ggplot(pca_tau_data, aes(PC1, PC2)) +
  geom_point() +
  xlab(paste("PC1 - ", pca_tau_var_per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca_tau_var_per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("Tau PCA Graph")


tau_loading_scores <- pca_tau$rotation[,1]
tau_methylation_scores <- abs(tau_loading_scores) # what to do of tau_methylation_scores?
  
library(ggfortify)
pca_tau_res <- prcomp(na.omit(tau[c(9:57)]), scale = FALSE)

autoplot(pca_tau_res, data = na.omit(tau[c(1:57)]), colour = 'Zeit',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)
autoplot(pca_tau_res, data = na.omit(tau[c(1:57)]), colour = 'Gruppe', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)
autoplot(pca_tau_res, data = na.omit(tau[c(1:57,357)]), colour = 'Sex', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)


```

### Ranking Tau loadings
```{r}
# Rank CpGs by influence on PC1
ranked_tau_cpgs <- sort(abs(tau_loading_scores), decreasing = TRUE)
head(ranked_tau_cpgs)

```

Die ersten 10 CpG scheinen die wesentlichen zu sein, wie in den Boxplots zu erkennen! m277 = 0.1086290423\
m210 = 0.0785399456

| Rank | CpG  |        |
|------|------|-------:|
| 1    | m331 | 0.4438 |
| 2    | m337 | 0.4236 |
| 3    | m310 | 0.3856 |
| 4    | m315 | 0.3843 |
| 5    | m323 | 0.3771 |
| -    | ---- | ------ |
| 6    | m306 | 0.2459 |
| 7    | m219 | 0.1652 |
| 8    | m300 | 0.1489 |
| 9    | m293 | 0.1295 |
| 10   | m207 | 0.1246 |
| 11   | m249 | 0.1151 |
| 12   | m277 | 0.1086 |
| 13   | m210 | 0.0785 |



### Visualize ranked loadings

There are two clear gaps:

Between m323 (0.377) and m306 (0.246) → Δ = 0.131

Between m277 (0.109) and m210 (0.079) → Δ = 0.030

```{r}
plot(ranked_tau_cpgs, type = "b", main = "Absolute PC1 Loadings (Tau CpGs)",
     ylab = "|Loading|", xlab = "Ranked CpGs")
abline(v = 5.5, col = "red", lty = 2, lwd = 2)

```




## PSD95

```{r}
cpg_psd95 <- psd95[10:54]
pca_psd95 <- prcomp(na.omit(cpg_psd95), scale = FALSE)

## plot pc1 and pc2
plot(pca_psd95$x[,1], pca_psd95$x[,2])

## make a scree plot
pca_psd95_var <- pca_psd95$sdev^2
pca_psd95_var_per <- round(pca_psd95_var/sum(pca_psd95_var)*100, 1)

barplot(pca_psd95_var_per, main="PSD95 Scree Plot", xlab="Principal Component", ylab="Percent Variation")


## now make a fancy looking plot that shows the PCs and the variation:
library(ggplot2)


pca_psd95_data <- as.data.frame(pca_psd95$x)
pca_psd95_data$ID <- rownames(pca_psd95)

str(pca_psd95_data)

ggplot(pca_psd95_data, aes(PC1, PC2)) +
  geom_point() +
  xlab(paste("PC1 - ", pca_psd95_var_per[1], "%", sep="")) +
  ylab(paste("PC2 - ", pca_psd95_var_per[2], "%", sep="")) +
  theme_bw() +
  ggtitle("PSD95 PCA Graph")


psd95_loading_scores <- pca_psd95$rotation[,1]
psd95_methylation_scores <- abs(psd95_loading_scores) # what to do of tau_methylation_scores?
  
library(ggfortify)
pca_psd95_res <- prcomp(na.omit(psd95[c(10:54)]), scale = FALSE)

autoplot(pca_psd95_res, data = na.omit(psd95[c(1:54)]), colour = 'Zeitpunkt',
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)
autoplot(pca_psd95_res, data = na.omit(psd95[c(1:54)]), colour = 'Gruppe', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)
autoplot(pca_psd95_res, data = na.omit(psd95[c(1:54,353)]), colour = 'Sex', 
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 3)
```

### Ranking PSD95 loadings
```{r}
# Rank CpGs by influence on PC1
ranked_psd95_cpgs <- sort(abs(psd95_loading_scores), decreasing = TRUE)
head(ranked_psd95_cpgs)
```

| Rank | CpG | |Loading| |
|------|------|-----------:|
| 1 | p95 | 0.664 |
| 2 | p60 | 0.484 |
| 3 | p23 | 0.445 |
| - | --- | ----- |
| 4 | m35 | 0.153 |
| 5 | m28 | 0.103 |
| 6 | m418 | 0.097 |
| 7 | m24 | 0.094 |
| 8 | p91 | 0.094 |
| 9 | m400 | 0.086 |
| 10 | m11 | 0.078 |

### Visualize ranked loadings

```{r}
plot(ranked_psd95_cpgs, type = "b", main = "Absolute PC1 Loadings (PSD95 CpGs)",
     ylab = "|Loading|", xlab = "Ranked CpGs")
abline(v = 4, col = "red", lty = 2, lwd = 2)
```
Die ersten 3 CpG scheinen die wesentlichen zu sein, wie in den Boxplots zu erkennen!