---
title: "Exploratory Data Analysis"
author: "Jonas J.R. Koberschinski"
date: "14-11-2024"
last_modified: "2026-01-29"
output: html
---

```{r}
#| label: setup
#| include: false
# Global setup: libraries & options

library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(forcats)
library(rlang)

library(psych)
library(psychTools)
library(DescTools)

library(crosstable)
library(flextable)

# Global options
options(dplyr.summarise.inform = FALSE)
```


## Data import
```{r}
#| label: data-import
#| echo: true

# load mean&median data
# List of file names (without paths and extension)
file_names <- c(
  "bdnf_mean.txt",
  "bdnf_median.txt",
  
  "tau_mean.txt",
  "tau_median.txt",
  
  "psd95_mean.txt",
  "psd95_median.txt"
)

# Folder where the files are located
subfolder <- "/home/kobi/Work/TBM-R-Analysis/tables/mean&median_tables"

# Loop through the file names, read each table, and assign it to a variable
for (file_name in file_names) {
  # Construct the full file path by combining the subfolder and file name
  file_path <- file.path(subfolder, file_name)
  
  # Remove the .txt extension to get the base name for the variable
  var_name <- tools::file_path_sans_ext(file_name)
  
  # Read the table and assign it to a variable with the same name
  assign(var_name, read.table(file_path, sep = "\t", header = TRUE, row.names = NULL, quote = ""))
}


# load cpg-data
cpg <- read.table(file = '/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/cpg.txt', sep = "\t", header = TRUE, row.names = NULL, quote = "")


# load wide data
# List of file names (without paths and extension)
file_names_wide <- c(
  "bdnf_mean_wide.txt",
  "bdnf_median_wide.txt",
  
  "tau_mean_wide.txt",
  "tau_median_wide.txt",
  
  "psd95_mean_wide.txt",
  "psd95_median_wide.txt"
)

# Folder where the files are located
subfolder_wide <- "/home/kobi/Work/TBM-R-Analysis/tables/wide_format/"

# Loop through the file names, read each table, and assign it to a variable
for (file_name in file_names_wide) {
  # Construct the full file path by combining the subfolder and file name
  file_path <- file.path(subfolder_wide, file_name)
  
  # Remove the .txt extension to get the base name for the variable
  var_name <- tools::file_path_sans_ext(file_name)
  
  # Read the table and assign it to a variable with the same name
  assign(var_name, read.table(file_path, sep = "\t", header = TRUE, row.names = NULL, quote = ""))
}

```

# EDA (Exploratory Data Analysis)

## Basic demopgrahics (Code to table summarized by chatgpt beyogen auf bdnf)

```{r}
#| label: tbl-demographics
#| results: asis

mean_sd <- function(x, digits_mean = 2, digits_sd = 1) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)
  sprintf(paste0("%.", digits_mean, "f \u00B1 %.", digits_sd, "f"), m, s)
}

min_max <- function(x, digits = 0) {
  mn <- min(x, na.rm = TRUE)
  mx <- max(x, na.rm = TRUE)
  sprintf(paste0("%.", digits, "f / %.", digits, "f"), mn, mx)
}

female_male_pct <- function(sex_vec, digits = 1) {
  tab <- prop.table(table(sex_vec)) * 100
  f <- unname(tab["F"]); m <- unname(tab["M"])
  if (is.na(f)) f <- 0
  if (is.na(m)) m <- 0
  sprintf(paste0("%.", digits, "f / %.", digits, "f"), f, m)
}

group_stats <- function(df) {
  list(
    mean_age_sd = mean_sd(df$Age_start, 2, 1),
    min_max     = min_max(df$Age_start, 0),
    n           = nrow(df),
    fm_pct      = female_male_pct(df$Sex, 1)
  )
}

bdnf_piano   <- subset(bdnf_median_wide, Gruppe == "Piano")
bdnf_control <- subset(bdnf_median_wide, Gruppe == "Controll")

p <- group_stats(bdnf_piano)
c <- group_stats(bdnf_control)

cat(paste0(
  "| Characteristic        | Piano Group | Control Group |\n",
  "|-----------------------|------------:|--------------:|\n",
  "| Mean age \u00B1 SD (years) | ", p$mean_age_sd, " | ", c$mean_age_sd, " |\n",
  "| Min / Max             | ", p$min_max, " | ", c$min_max, " |\n",
  "| N                     | ", p$n, " | ", c$n, " |\n",
  "| Female / Male (%)     | ", p$fm_pct, " | ", c$fm_pct, " |\n"
))

```


### Extended (bezogen auf bdnf!)

```{r}
#| label: table1-demographics
#| results: asis

# ---- (5) robust labels: standardize Gruppe and Sex ----
df <- bdnf_median_wide
df$Gruppe <- trimws(df$Gruppe)
df$Sex    <- trimws(df$Sex)

df$Sex <- toupper(df$Sex)
df$Sex[df$Sex %in% c("FEMALE")] <- "F"
df$Sex[df$Sex %in% c("MALE")]   <- "M"

g_piano   <- "Piano"
g_control <- "Controll"

df_piano   <- subset(df, Gruppe == g_piano)
df_control <- subset(df, Gruppe == g_control)
df_overall <- df

# ---- helpers ----
fmt_mean_sd <- function(x, dm = 2, ds = 1) {
  m <- mean(x, na.rm = TRUE); s <- sd(x, na.rm = TRUE)
  sprintf(paste0("%.", dm, "f \u00B1 %.", ds, "f"), m, s)
}
fmt_median_iqr <- function(x, d = 2) {
  med <- median(x, na.rm = TRUE)
  q <- quantile(x, probs = c(0.25, 0.75), na.rm = TRUE, names = FALSE)
  sprintf(paste0("%.", d, "f (%.", d, "f, %.", d, "f)"), med, q[1], q[2])
}
fmt_min_max <- function(x, d = 0) {
  mn <- min(x, na.rm = TRUE); mx <- max(x, na.rm = TRUE)
  sprintf(paste0("%.", d, "f / %.", d, "f"), mn, mx)
}
n_nonmiss <- function(x) sum(!is.na(x))

fmt_fm_pct <- function(sex_vec, d = 1) {
  tab <- prop.table(table(sex_vec)) * 100
  f <- unname(tab["F"]); m <- unname(tab["M"])
  if (is.na(f)) f <- 0
  if (is.na(m)) m <- 0
  sprintf(paste0("%.", d, "f / %.", d, "f"), f, m)
}
fmt_missing <- function(x) {
  n_miss <- sum(is.na(x))
  n_tot  <- length(x)
  sprintf("%d (%0.1f%%)", n_miss, 100 * n_miss / n_tot)
}

summ_one <- function(d) {
  list(
    n_participants = nrow(d),
    age_nonmiss    = n_nonmiss(d$Age_start),
    age_mean_sd    = fmt_mean_sd(d$Age_start, 2, 1),
    age_median_iqr = fmt_median_iqr(d$Age_start, 2),
    age_min_max    = fmt_min_max(d$Age_start, 0),
    sex_fm_pct     = fmt_fm_pct(d$Sex, 1),
    sex_missing    = fmt_missing(d$Sex),

    t0_n           = n_nonmiss(d$T0),
    t1_n           = n_nonmiss(d$T1),
    t2_n           = n_nonmiss(d$T2),
    t0_missing     = fmt_missing(d$T0),
    t1_missing     = fmt_missing(d$T1),
    t2_missing     = fmt_missing(d$T2),
    total_samples  = n_nonmiss(d$T0) + n_nonmiss(d$T1) + n_nonmiss(d$T2)
  )
}

P <- summ_one(df_piano)
C <- summ_one(df_control)
O <- summ_one(df_overall)

cat(paste0(
  "| Characteristic | Piano | Control | Overall |\n",
  "|---|---:|---:|---:|\n",
  "| Participants (N) | ", P$n_participants, " | ", C$n_participants, " | ", O$n_participants, " |\n",
  "| Age non-missing (n) | ", P$age_nonmiss, " | ", C$age_nonmiss, " | ", O$age_nonmiss, " |\n",
  "| Mean age \u00B1 SD (years) | ", P$age_mean_sd, " | ", C$age_mean_sd, " | ", O$age_mean_sd, " |\n",
  "| Median (IQR) age (years) | ", P$age_median_iqr, " | ", C$age_median_iqr, " | ", O$age_median_iqr, " |\n",
  "| Min / Max age (years) | ", P$age_min_max, " | ", C$age_min_max, " | ", O$age_min_max, " |\n",
  "| Female / Male (%) | ", P$sex_fm_pct, " | ", C$sex_fm_pct, " | ", O$sex_fm_pct, " |\n",
  "| Sex missing, n (%) | ", P$sex_missing, " | ", C$sex_missing, " | ", O$sex_missing, " |\n",
  "| BDNF samples at T0 (n) | ", P$t0_n, " | ", C$t0_n, " | ", O$t0_n, " |\n",
  "| BDNF samples at T1 (n) | ", P$t1_n, " | ", C$t1_n, " | ", O$t1_n, " |\n",
  "| BDNF samples at T2 (n) | ", P$t2_n, " | ", C$t2_n, " | ", O$t2_n, " |\n",
  "| Total BDNF samples (T0+T1+T2) | ", P$total_samples, " | ", C$total_samples, " | ", O$total_samples, " |\n",
  "| Missing T0, n (%) | ", P$t0_missing, " | ", C$t0_missing, " | ", O$t0_missing, " |\n",
  "| Missing T1, n (%) | ", P$t1_missing, " | ", C$t1_missing, " | ", O$t1_missing, " |\n",
  "| Missing T2, n (%) | ", P$t2_missing, " | ", C$t2_missing, " | ", O$t2_missing, " |\n"
))

```

### Table with all genes

```{r}
#| label: table-samples-bdnf-tau-psd95
#| results: asis

# helper: robust trimming
prep_df <- function(df) {
  df$Gruppe <- trimws(df$Gruppe)
  df
}

n_nonmiss <- function(x) sum(!is.na(x))

# summarize one biomarker wide-df into a named row
summ_marker <- function(df, marker, g_piano = "Piano", g_control = "Controll") {
  df <- prep_df(df)

  dP <- subset(df, Gruppe == g_piano)
  dC <- subset(df, Gruppe == g_control)
  dO <- df

  c(
    Marker = marker,

    Piano_T0 = n_nonmiss(dP$T0),
    Piano_T1 = n_nonmiss(dP$T1),
    Piano_T2 = n_nonmiss(dP$T2),
    Piano_Total = n_nonmiss(dP$T0) + n_nonmiss(dP$T1) + n_nonmiss(dP$T2),

    Control_T0 = n_nonmiss(dC$T0),
    Control_T1 = n_nonmiss(dC$T1),
    Control_T2 = n_nonmiss(dC$T2),
    Control_Total = n_nonmiss(dC$T0) + n_nonmiss(dC$T1) + n_nonmiss(dC$T2),

    Overall_T0 = n_nonmiss(dO$T0),
    Overall_T1 = n_nonmiss(dO$T1),
    Overall_T2 = n_nonmiss(dO$T2),
    Overall_Total = n_nonmiss(dO$T0) + n_nonmiss(dO$T1) + n_nonmiss(dO$T2)
  )
}

rows <- rbind(
  summ_marker(bdnf_median_wide,  "BDNF"),
  summ_marker(tau_median_wide,   "TAU"),
  summ_marker(psd95_median_wide, "PSD95")
)

# render as markdown table
cat(paste0(
  "| Marker | Piano T0 | Piano T1 | Piano T2 | Piano Total | Control T0 | Control T1 | Control T2 | Control Total | Overall T0 | Overall T1 | Overall T2 | Overall Total |\n",
  "|---|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|\n",
  paste(apply(rows, 1, function(r) {
    paste0("| ", r["Marker"], " | ",
           r["Piano_T0"], " | ", r["Piano_T1"], " | ", r["Piano_T2"], " | ", r["Piano_Total"], " | ",
           r["Control_T0"], " | ", r["Control_T1"], " | ", r["Control_T2"], " | ", r["Control_Total"], " | ",
           r["Overall_T0"], " | ", r["Overall_T1"], " | ", r["Overall_T2"], " | ", r["Overall_Total"], " |")
  }), collapse = "\n"),
  "\n"
))

```
### another demopgrahic (bezogen auf BDNF! villeicht ändern)

```{r}
summ_one <- function(d) {
  list(
    n_participants = nrow(d),
    age_nonmiss    = n_nonmiss(d$Age_start),
    age_mean_sd    = fmt_mean_sd(d$Age_start, 2, 1),
    age_median_iqr = fmt_median_iqr(d$Age_start, 2),
    age_min_max    = fmt_min_max(d$Age_start, 0),
    sex_fm_pct     = fmt_fm_pct(d$Sex, 1),
    sex_missing    = fmt_missing(d$Sex)
  )
}

P <- summ_one(df_piano)
C <- summ_one(df_control)
O <- summ_one(df_overall)

cat(paste0(
  "| Characteristic | Piano | Control | Overall |\n",
  "|---|---:|---:|---:|\n",
  "| Participants (N) | ", P$n_participants, " | ", C$n_participants, " | ", O$n_participants, " |\n",
  "| Age non-missing (n) | ", P$age_nonmiss, " | ", C$age_nonmiss, " | ", O$age_nonmiss, " |\n",
  "| Mean age \u00B1 SD (years) | ", P$age_mean_sd, " | ", C$age_mean_sd, " | ", O$age_mean_sd, " |\n",
  "| Median (IQR) age (years) | ", P$age_median_iqr, " | ", C$age_median_iqr, " | ", O$age_median_iqr, " |\n",
  "| Min / Max age (years) | ", P$age_min_max, " | ", C$age_min_max, " | ", O$age_min_max, " |\n",
  "| Female / Male (%) | ", P$sex_fm_pct, " | ", C$sex_fm_pct, " | ", O$sex_fm_pct, " |\n",
  "| Sex missing, n (%) | ", P$sex_missing, " | ", C$sex_missing, " | ", O$sex_missing, " |\n"
))

```

### all genes with missing

```{r}
#| label: table-genes-missing
#| results: asis

prep_df <- function(df) {
  df$Gruppe <- trimws(df$Gruppe)
  df
}

n_nonmiss <- function(x) sum(!is.na(x))

fmt_missing <- function(x) {
  n_miss <- sum(is.na(x))
  n_tot  <- length(x)
  sprintf("%d (%0.1f%%)", n_miss, 100 * n_miss / n_tot)
}

summ_marker <- function(df, marker, g_piano = "Piano", g_control = "Controll") {
  df <- prep_df(df)

  dP <- subset(df, Gruppe == g_piano)
  dC <- subset(df, Gruppe == g_control)
  dO <- df

  # counts
  P_t0 <- n_nonmiss(dP$T0); P_t1 <- n_nonmiss(dP$T1); P_t2 <- n_nonmiss(dP$T2)
  C_t0 <- n_nonmiss(dC$T0); C_t1 <- n_nonmiss(dC$T1); C_t2 <- n_nonmiss(dC$T2)
  O_t0 <- n_nonmiss(dO$T0); O_t1 <- n_nonmiss(dO$T1); O_t2 <- n_nonmiss(dO$T2)

  # missing (%)
  P_m0 <- fmt_missing(dP$T0); P_m1 <- fmt_missing(dP$T1); P_m2 <- fmt_missing(dP$T2)
  C_m0 <- fmt_missing(dC$T0); C_m1 <- fmt_missing(dC$T1); C_m2 <- fmt_missing(dC$T2)
  O_m0 <- fmt_missing(dO$T0); O_m1 <- fmt_missing(dO$T1); O_m2 <- fmt_missing(dO$T2)

  list(
    marker = marker,
    Piano   = list(t0 = P_t0, t1 = P_t1, t2 = P_t2,
                   tot = P_t0 + P_t1 + P_t2,
                   m0 = P_m0, m1 = P_m1, m2 = P_m2),
    Control = list(t0 = C_t0, t1 = C_t1, t2 = C_t2,
                   tot = C_t0 + C_t1 + C_t2,
                   m0 = C_m0, m1 = C_m1, m2 = C_m2),
    Overall = list(t0 = O_t0, t1 = O_t1, t2 = O_t2,
                   tot = O_t0 + O_t1 + O_t2,
                   m0 = O_m0, m1 = O_m1, m2 = O_m2)
  )
}

rows <- list(
  summ_marker(bdnf_median_wide,  "BDNF"),
  summ_marker(tau_median_wide,   "TAU"),
  summ_marker(psd95_median_wide, "PSD95")
)

cat(paste0(
  "| Marker | Piano T0 n | Piano T1 n | Piano T2 n | Piano Total | Piano missing T0 | Piano missing T1 | Piano missing T2 | Control T0 n | Control T1 n | Control T2 n | Control Total | Control missing T0 | Control missing T1 | Control missing T2 | Overall T0 n | Overall T1 n | Overall T2 n | Overall Total | Overall missing T0 | Overall missing T1 | Overall missing T2 |\n",
  "|---|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|\n",
  paste(vapply(rows, function(r) {
    paste0(
      "| ", r$marker,
      " | ", r$Piano$t0, " | ", r$Piano$t1, " | ", r$Piano$t2, " | ", r$Piano$tot,
      " | ", r$Piano$m0, " | ", r$Piano$m1, " | ", r$Piano$m2,
      " | ", r$Control$t0, " | ", r$Control$t1, " | ", r$Control$t2, " | ", r$Control$tot,
      " | ", r$Control$m0, " | ", r$Control$m1, " | ", r$Control$m2,
      " | ", r$Overall$t0, " | ", r$Overall$t1, " | ", r$Overall$t2, " | ", r$Overall$tot,
      " | ", r$Overall$m0, " | ", r$Overall$m1, " | ", r$Overall$m2,
      " |"
    )
  }, character(1)), collapse = "\n"),
  "\n"
))

```

```{r}
#| label: tables-per-marker
#| results: asis

# ---------- helpers ----------
fmt_mean_sd <- function(x, dm = 2, ds = 1) {
  m <- mean(x, na.rm = TRUE); s <- sd(x, na.rm = TRUE)
  sprintf(paste0("%.", dm, "f \u00B1 %.", ds, "f"), m, s)
}
fmt_median_iqr <- function(x, d = 2) {
  med <- median(x, na.rm = TRUE)
  q <- quantile(x, probs = c(0.25, 0.75), na.rm = TRUE, names = FALSE)
  sprintf(paste0("%.", d, "f (%.", d, "f, %.", d, "f)"), med, q[1], q[2])
}
fmt_min_max <- function(x, d = 0) {
  mn <- min(x, na.rm = TRUE); mx <- max(x, na.rm = TRUE)
  sprintf(paste0("%.", d, "f / %.", d, "f"), mn, mx)
}
n_nonmiss <- function(x) sum(!is.na(x))
fmt_missing <- function(x) {
  n_miss <- sum(is.na(x)); n_tot <- length(x)
  sprintf("%d (%0.1f%%)", n_miss, 100 * n_miss / n_tot)
}
fmt_fm_pct <- function(sex_vec, d = 1) {
  sex_vec <- trimws(toupper(sex_vec))
  sex_vec[sex_vec %in% c("FEMALE")] <- "F"
  sex_vec[sex_vec %in% c("MALE")]   <- "M"
  tab <- prop.table(table(sex_vec)) * 100
  f <- unname(tab["F"]); m <- unname(tab["M"])
  if (is.na(f)) f <- 0
  if (is.na(m)) m <- 0
  sprintf(paste0("%.", d, "f / %.", d, "f"), f, m)
}

summ_one <- function(d) {
  list(
    n_participants = nrow(d),
    age_nonmiss    = n_nonmiss(d$Age_start),
    age_mean_sd    = fmt_mean_sd(d$Age_start, 2, 1),
    age_median_iqr = fmt_median_iqr(d$Age_start, 2),
    age_min_max    = fmt_min_max(d$Age_start, 0),
    sex_fm_pct     = fmt_fm_pct(d$Sex, 1),
    sex_missing    = fmt_missing(d$Sex),

    t0_n           = n_nonmiss(d$T0),
    t1_n           = n_nonmiss(d$T1),
    t2_n           = n_nonmiss(d$T2),
    t0_missing     = fmt_missing(d$T0),
    t1_missing     = fmt_missing(d$T1),
    t2_missing     = fmt_missing(d$T2),
    total_samples  = n_nonmiss(d$T0) + n_nonmiss(d$T1) + n_nonmiss(d$T2)
  )
}

# ---------- main function: prints ONE table for one marker ----------
print_marker_table <- function(df_wide, marker_name,
                               g_piano = "Piano", g_control = "Controll") {
  df <- df_wide
  df$Gruppe <- trimws(df$Gruppe)
  df$Sex    <- trimws(df$Sex)

  df_piano   <- subset(df, Gruppe == g_piano)
  df_control <- subset(df, Gruppe == g_control)
  df_overall <- df

  P <- summ_one(df_piano)
  C <- summ_one(df_control)
  O <- summ_one(df_overall)

  cat("\n\n### ", marker_name, "\n\n", sep = "")
  cat(paste0(
    "| Characteristic | Piano | Control | Overall |\n",
    "|---|---:|---:|---:|\n",
    "| Participants (N) | ", P$n_participants, " | ", C$n_participants, " | ", O$n_participants, " |\n",
    "| Age non-missing (n) | ", P$age_nonmiss, " | ", C$age_nonmiss, " | ", O$age_nonmiss, " |\n",
    "| Mean age \u00B1 SD (years) | ", P$age_mean_sd, " | ", C$age_mean_sd, " | ", O$age_mean_sd, " |\n",
    "| Median (IQR) age (years) | ", P$age_median_iqr, " | ", C$age_median_iqr, " | ", O$age_median_iqr, " |\n",
    "| Min / Max age (years) | ", P$age_min_max, " | ", C$age_min_max, " | ", O$age_min_max, " |\n",
    "| Female / Male (%) | ", P$sex_fm_pct, " | ", C$sex_fm_pct, " | ", O$sex_fm_pct, " |\n",
    "| Sex missing, n (%) | ", P$sex_missing, " | ", C$sex_missing, " | ", O$sex_missing, " |\n",
    "| ", marker_name, " samples at T0 (n) | ", P$t0_n, " | ", C$t0_n, " | ", O$t0_n, " |\n",
    "| ", marker_name, " samples at T1 (n) | ", P$t1_n, " | ", C$t1_n, " | ", O$t1_n, " |\n",
    "| ", marker_name, " samples at T2 (n) | ", P$t2_n, " | ", C$t2_n, " | ", O$t2_n, " |\n",
    "| Total ", marker_name, " samples (T0+T1+T2) | ", P$total_samples, " | ", C$total_samples, " | ", O$total_samples, " |\n",
    "| Missing T0, n (%) | ", P$t0_missing, " | ", C$t0_missing, " | ", O$t0_missing, " |\n",
    "| Missing T1, n (%) | ", P$t1_missing, " | ", C$t1_missing, " | ", O$t1_missing, " |\n",
    "| Missing T2, n (%) | ", P$t2_missing, " | ", C$t2_missing, " | ", O$t2_missing, " |\n"
  ))
}

# ---------- print the three separate tables ----------
print_marker_table(bdnf_median_wide,  "BDNF")
print_marker_table(tau_median_wide,   "TAU")
print_marker_table(psd95_median_wide, "PSD95")

```
### Union demographics across all genes
```{r}
#| label: table-union-demographics-genes
#| results: asis

library(dplyr)

g_piano   <- "Piano"
g_control <- "Controll"

# ---- helpers ----
mean_sd <- function(x) sprintf("%.2f \u00B1 %.1f", mean(x, na.rm=TRUE), sd(x, na.rm=TRUE))
median_iqr <- function(x) {
  q <- quantile(x, c(.25,.5,.75), na.rm=TRUE, names = FALSE)
  sprintf("%.2f (%.2f, %.2f)", q[2], q[1], q[3])
}
min_max <- function(x) sprintf("%.0f / %.0f", min(x, na.rm=TRUE), max(x, na.rm=TRUE))

fmt_missing <- function(x) sprintf("%d (%.1f%%)", sum(is.na(x)), 100 * mean(is.na(x)))

fm_pct <- function(sex_vec) {
  s <- toupper(trimws(sex_vec))
  s[s %in% c("FEMALE")] <- "F"
  s[s %in% c("MALE")]   <- "M"
  tab <- prop.table(table(s)) * 100
  f <- unname(tab["F"]); m <- unname(tab["M"])
  if (is.na(f)) f <- 0
  if (is.na(m)) m <- 0
  sprintf("%.1f / %.1f", f, m)
}

n_nonmiss <- function(x) sum(!is.na(x))

# ---- UNION master for demographics (unique IDs across all genes) ----
demo_union <- bind_rows(
  bdnf_median_wide  %>% select(ID, Gruppe, Sex, Age_start),
  tau_median_wide   %>% select(ID, Gruppe, Sex, Age_start),
  psd95_median_wide %>% select(ID, Gruppe, Sex, Age_start)
) %>%
  mutate(
    Gruppe = trimws(Gruppe),
    Sex    = trimws(Sex)
  ) %>%
  group_by(ID) %>%
  summarise(
    Gruppe    = first(na.omit(Gruppe)),
    Sex       = first(na.omit(Sex)),
    Age_start = first(na.omit(Age_start)),
    .groups = "drop"
  )

summ_demo <- function(d) list(
  N = nrow(d),
  age_n = sum(!is.na(d$Age_start)),
  mean_sd = mean_sd(d$Age_start),
  med_iqr = median_iqr(d$Age_start),
  minmax = min_max(d$Age_start),
  fm = fm_pct(d$Sex),
  sex_miss = fmt_missing(d$Sex)
)

P <- summ_demo(filter(demo_union, Gruppe == g_piano))
C <- summ_demo(filter(demo_union, Gruppe == g_control))
O <- summ_demo(demo_union)

# ---- gene rows: n and missing per timepoint for each marker ----
prep_gene <- function(df) mutate(df, Gruppe = trimws(Gruppe))

gene_stats <- function(df, marker) {
  df <- prep_gene(df)
  P <- filter(df, Gruppe == g_piano)
  C <- filter(df, Gruppe == g_control)
  O <- df

  list(
    marker = marker,
    P = list(t0=n_nonmiss(P$T0), t1=n_nonmiss(P$T1), t2=n_nonmiss(P$T2),
             m0=fmt_missing(P$T0), m1=fmt_missing(P$T1), m2=fmt_missing(P$T2)),
    C = list(t0=n_nonmiss(C$T0), t1=n_nonmiss(C$T1), t2=n_nonmiss(C$T2),
             m0=fmt_missing(C$T0), m1=fmt_missing(C$T1), m2=fmt_missing(C$T2)),
    O = list(t0=n_nonmiss(O$T0), t1=n_nonmiss(O$T1), t2=n_nonmiss(O$T2),
             m0=fmt_missing(O$T0), m1=fmt_missing(O$T1), m2=fmt_missing(O$T2))
  )
}

genes <- list(
  gene_stats(bdnf_median_wide,  "BDNF"),
  gene_stats(tau_median_wide,   "TAU"),
  gene_stats(psd95_median_wide, "PSD95")
)

# ---- print unified table ----
cat(paste0(
  "| Characteristic | Piano | Control | Overall |\n",
  "|---|---:|---:|---:|\n",
  "| Participants (N) | ", P$N, " | ", C$N, " | ", O$N, " |\n",
  "| Age non-missing (n) | ", P$age_n, " | ", C$age_n, " | ", O$age_n, " |\n",
  "| Mean age \u00B1 SD (years) | ", P$mean_sd, " | ", C$mean_sd, " | ", O$mean_sd, " |\n",
  "| Median (IQR) age (years) | ", P$med_iqr, " | ", C$med_iqr, " | ", O$med_iqr, " |\n",
  "| Min / Max age (years) | ", P$minmax, " | ", C$minmax, " | ", O$minmax, " |\n",
  "| Female / Male (%) | ", P$fm, " | ", C$fm, " | ", O$fm, " |\n",
  "| Sex missing, n (%) | ", P$sex_miss, " | ", C$sex_miss, " | ", O$sex_miss, " |\n"
))

for (r in genes) {
  cat(paste0(
    "| ", r$marker, " samples at T0 (n) | ", r$P$t0, " | ", r$C$t0, " | ", r$O$t0, " |\n",
    "| ", r$marker, " samples at T1 (n) | ", r$P$t1, " | ", r$C$t1, " | ", r$O$t1, " |\n",
    "| ", r$marker, " samples at T2 (n) | ", r$P$t2, " | ", r$C$t2, " | ", r$O$t2, " |\n",
    "| ", r$marker, " missing T0, n (%) | ", r$P$m0, " | ", r$C$m0, " | ", r$O$m0, " |\n",
    "| ", r$marker, " missing T1, n (%) | ", r$P$m1, " | ", r$C$m1, " | ", r$O$m1, " |\n",
    "| ", r$marker, " missing T2, n (%) | ", r$P$m2, " | ", r$C$m2, " | ", r$O$m2, " |\n"
  ))
}

```

## Summary Statistics be

### 

```{r}
tab <- data.frame(
  Marker = c("BDNF", "Tau", "PSD95"),
  Mean = c(
    mean(bdnf_mean$Mean, na.rm = TRUE),
    mean(tau_mean$Mean, na.rm = TRUE),
    mean(psd95_mean$Mean, na.rm = TRUE)
  ),
  SD = c(
    sd(bdnf_mean$Mean, na.rm = TRUE),
    sd(tau_mean$Mean, na.rm = TRUE),
    sd(psd95_mean$Mean, na.rm = TRUE)
  )
)

tab$`Mean ± SD` <- sprintf("%.3f ± %.3f", tab$Mean, tab$SD)
tab <- tab[, c("Marker", "Mean ± SD")]

# Print as markdown
cat("| Marker | Mean ± SD |\n")
cat("|--------|-----------|\n")
apply(tab, 1, function(r) cat(sprintf("| %s | %s |\n", r[1], r[2])))
```

## Plots

### Interaction Plots by Gene (mean across CpGs)

```{r}
#| label: fig-interactions-no-group
#| fig-cap: "Methylation levels (mean ± SE) across time points collapsed across groups for BDNF, Tau, and PSD95."
#| fig-width: 8
#| fig-height: 10

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

make_nogroup_plot <- function(df_wide, title,
                              ylab = "Mean methylation (mean across CpGs)") {

  df_long <- df_wide %>%
    pivot_longer(cols = c(T0, T1, T2),
                 names_to = "Zeitpunkt",
                 values_to = "Mean") %>%
    mutate(Zeitpunkt = factor(Zeitpunkt, levels = c("T0","T1","T2")))

  sum0 <- df_long %>%
    group_by(Zeitpunkt) %>%
    summarise(
      n = sum(!is.na(Mean)),
      mean = mean(Mean, na.rm = TRUE),
      se   = sd(Mean, na.rm = TRUE) / sqrt(n),
      .groups = "drop"
    )

  ggplot(sum0, aes(x = Zeitpunkt, y = mean, group = 1)) +
    geom_line() +
    geom_point() +
    geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
    theme_classic() +
    labs(title = title, x = "Time Point", y = ylab,
         caption = "Points show mean across participants; error bars are ±SE.")
}

p_bdnf0  <- make_nogroup_plot(bdnf_mean_wide,  title = "BDNF")
p_tau0   <- make_nogroup_plot(tau_mean_wide,   title = "Tau")
p_psd950 <- make_nogroup_plot(psd95_mean_wide, title = "PSD95")

(p_bdnf0 / p_tau0 / p_psd950)
```

```{r}
#| label: fig-interactions-no-group-with-significance
#| fig-cap: "Methylation levels (mean ± SE) across time points collapsed across groups for BDNF, Tau, and PSD95. Brackets show pairwise timepoint contrasts (mixed model: Mean ~ Zeitpunkt + (1|ID))."
#| fig-width: 8
#| fig-height: 10

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(lme4)
library(emmeans)
library(ggsignif)

p_to_star <- function(p){
  ifelse(is.na(p), "NA",
  ifelse(p < 0.001, "***",
  ifelse(p < 0.01,  "**",
  ifelse(p < 0.05,  "*", "ns"))))
}

make_nogroup_plot <- function(df_wide, title,
                              ylab = "Mean methylation (mean across CpGs)") {

  df_long <- df_wide %>%
    pivot_longer(cols = c(T0, T1, T2),
                 names_to = "Zeitpunkt",
                 values_to = "Mean") %>%
    mutate(
      Zeitpunkt = factor(Zeitpunkt, levels = c("T0","T1","T2")),
      ID = factor(ID)
    ) %>%
    filter(!is.na(Mean))

  # summary stats (SE)
  sum0 <- df_long %>%
    group_by(Zeitpunkt) %>%
    summarise(
      n = n(),
      mean = mean(Mean),
      se   = sd(Mean) / sqrt(n),
      .groups = "drop"
    )

  # mixed model for timepoint differences (within-subject)
  fit <- lmer(Mean ~ Zeitpunkt + (1 | ID), data = df_long, REML = TRUE)

  # pairwise contrasts between timepoints
  pw <- as.data.frame(pairs(emmeans(fit, ~ Zeitpunkt))) %>%
    transmute(
      contrast,
      p.value,
      label = p_to_star(p.value)
    ) %>%
    separate(contrast, into = c("x1", "x2"), sep = " - ")

  # bracket y-positions (stacked)
  y_top  <- max(sum0$mean + sum0$se, na.rm = TRUE)
  y_step <- 0.08 * diff(range(sum0$mean, na.rm = TRUE))
  if (!is.finite(y_step) || y_step == 0) y_step <- 0.01

  pw <- pw %>%
    mutate(y_position = y_top + y_step * row_number())

  ggplot(sum0, aes(x = Zeitpunkt, y = mean, group = 1)) +
    geom_line() +
    geom_point() +
    geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
    ggsignif::geom_signif(
      data = pw,
      aes(xmin = x1, xmax = x2, annotations = label, y_position = y_position),
      manual = TRUE,
      inherit.aes = FALSE,
      textsize = 4,
      tip_length = 0.01
    ) +
    theme_classic() +
    labs(
      title = title,
      x = "Time Point",
      y = ylab,
      caption = "Error bars are ±SE. Brackets: pairwise timepoint contrasts from mixed model."
    )
}

p_bdnf0  <- make_nogroup_plot(bdnf_mean_wide,  title = "BDNF")
p_tau0   <- make_nogroup_plot(tau_mean_wide,   title = "Tau")
p_psd950 <- make_nogroup_plot(psd95_mean_wide, title = "PSD95")

(p_bdnf0 / p_tau0 / p_psd950)

```


### Interaction Plots by Gene (mean across CpGs)
```{r}
#| label: fig-interactions-by-gene-mean
#| fig-cap: "Methylation levels (mean ± SE) across time points by group for BDNF, Tau, and PSD95."
#| fig-width: 8
#| fig-height: 12
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

make_interaction_plot <- function(df_wide, title, ylab = "Mean methylation (mean across CpGs)") {

  df_long <- df_wide %>%
    pivot_longer(cols = c(T0, T1, T2),
                 names_to = "Zeitpunkt",
                 values_to = "Mean") %>%
    mutate(
      Zeitpunkt = factor(Zeitpunkt, levels = c("T0", "T1", "T2")),
      Gruppe = factor(Gruppe)
    )

  summary_stats <- df_long %>%
    group_by(Zeitpunkt, Gruppe) %>%
    summarise(
      n = sum(!is.na(Mean)),
      mean = mean(Mean, na.rm = TRUE),
      se = sd(Mean, na.rm = TRUE) / sqrt(n),
      .groups = "drop"
    )

  ggplot(summary_stats, aes(x = Zeitpunkt, y = mean, group = Gruppe, colour = Gruppe)) +
    geom_line() +
    geom_point() +
    geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
    theme_classic() +
    labs(title = title, x = "Time Point", y = ylab, caption = "Points show mean across participants; error bars are ±SE.")
}

# BDNF
p_bdnf <-
make_interaction_plot(
  bdnf_mean_wide,
  title = "Interaction Plot of BDNF Methylation Levels by Group and Time"
)

# TAU
p_tau <-
make_interaction_plot(
  tau_mean_wide,
  title = "Interaction Plot of Tau Methylation Levels by Group and Time"
)

# PSD95
p_psd95 <-
make_interaction_plot(
  psd95_mean_wide,
  title = "Interaction Plot of PSD95 Methylation Levels by Group and Time"
)


(p_bdnf / p_tau / p_psd95) + plot_layout(guides = "collect") &
  theme(legend.position = "right")

```

### Interaction Plots by Gene (mean across CpGs) with significance stars

```{r}
#| label: fig-interactions-by-gene-mean-with-significance
#| fig-cap: "Methylation levels (mean ± 95% CI) across time points by group for BDNF, Tau, and PSD95. Stars indicate Piano vs Control contrasts within each time point from a mixed model (Mean ~ Gruppe * Zeitpunkt + (1|ID))."
#| fig-width: 8
#| fig-height: 12

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(lme4)
library(emmeans)

p_to_star <- function(p){
  ifelse(is.na(p), "NA",
  ifelse(p < 0.001, "***",
  ifelse(p < 0.01,  "**",
  ifelse(p < 0.05,  "*", "ns"))))
}

make_interaction_plot_sig <- function(df_wide, title,
                                      ylab = "Mean methylation (mean across CpGs)",
                                      control_label = "Controll",
                                      piano_label = "Piano") {

  # wide -> long
  df_long <- df_wide %>%
    pivot_longer(cols = c(T0, T1, T2),
                 names_to = "Zeitpunkt",
                 values_to = "Mean") %>%
    mutate(
      Zeitpunkt = factor(Zeitpunkt, levels = c("T0", "T1", "T2")),
      Gruppe = factor(Gruppe, levels = c(control_label, piano_label)),
      ID = factor(ID)
    ) %>%
    filter(!is.na(Mean))

  # mixed model (participant-level)
  fit <- lmer(Mean ~ Gruppe * Zeitpunkt + (1 | ID), data = df_long, REML = TRUE)

  # Piano vs Control at each timepoint
  emm <- emmeans(fit, ~ Gruppe | Zeitpunkt)
  contr <- as.data.frame(contrast(emm, method = list("Piano - Control" = c(-1, 1))))

  pvals <- contr %>%
    transmute(
      Zeitpunkt,
      p.value,
      label = p_to_star(p.value)
    )

  # summary stats for plot (95% CI)
  summary_stats <- df_long %>%
    group_by(Zeitpunkt, Gruppe) %>%
    summarise(
      n = n(),
      mean = mean(Mean),
      sd = sd(Mean),
      se = sd / sqrt(n),
      ci = qt(0.975, df = n - 1) * se,
      .groups = "drop"
    )

  # y-positions for labels (above the top CI per timepoint)
  y_pad <- 0.06 * diff(range(summary_stats$mean, na.rm = TRUE))
  ann <- summary_stats %>%
    group_by(Zeitpunkt) %>%
    summarise(y = max(mean + ci, na.rm = TRUE) + y_pad, .groups = "drop") %>%
    left_join(pvals, by = "Zeitpunkt")

  ggplot(summary_stats, aes(x = Zeitpunkt, y = mean, group = Gruppe, colour = Gruppe)) +
    geom_line() +
    geom_point() +
    geom_errorbar(aes(ymin = mean - ci, ymax = mean + ci), width = 0.2) +
    geom_text(data = ann, aes(x = Zeitpunkt, y = y, label = label),
              inherit.aes = FALSE, vjust = 0, size = 4) +
    theme_classic() +
    labs(
      title = title,
      x = "Time Point",
      y = ylab,
      caption = "Error bars are 95% CI. Stars: Piano vs Control within timepoint from mixed model."
    )
}

# BDNF
p_bdnf <- make_interaction_plot_sig(
  bdnf_mean_wide,
  title = "Interaction Plot of BDNF Methylation Levels by Group and Time"
)

# TAU
p_tau <- make_interaction_plot_sig(
  tau_mean_wide,
  title = "Interaction Plot of Tau Methylation Levels by Group and Time"
)

# PSD95
p_psd95 <- make_interaction_plot_sig(
  psd95_mean_wide,
  title = "Interaction Plot of PSD95 Methylation Levels by Group and Time"
)

(p_bdnf / p_tau / p_psd95) + plot_layout(guides = "collect") &
  theme(legend.position = "right")

```

## Explanation of Methylation Summary Statistics
ChatGPT-help (cf. plots-time-methylation.qmd)

Totally fair to ask again — this is the key conceptual trap in methylation papers, and you’re right at the edge of it. I’ll explain it slowly and cleanly, with your exact data in mind.

The core issue in one sentence

A CpG is not a participant.
If you treat CpGs like independent observations, you inflate N and distort SD/SE.

Everything else follows from that.

What your data actually look like
1️⃣ CpG-level data (long format: BDNF2, TAU2, …)

Each row is:

one CpG site × one person × one time point

Example (simplified):

ID	Time	CpG	methylation
1	T0	m148	0.18
1	T0	m111	0.23
1	T0	m087	0.00
…	…	…	…

So for one person at one time, you might have 20–30 CpGs.

When you do:

mean(BDNF2$methylierung)
sd(BDNF2$methylierung)


you are computing:

“What is the mean/SD of all CpG measurements pooled together?”

This mixes three different sources of variation:

Differences between CpG sites (huge!)

Differences between people

Changes over time + noise

That’s why:

SD is large (~0.09–0.10)

SE becomes artificially tiny if you divide by sqrt(nrow(BDNF2))

Plots look “too confident”

This is not wrong mathematically, but it answers the wrong scientific question.

What your paper actually claims to measure

Your tables, text, and mixed models all say things like:

“BDNF methylation level per gene”

That implies:

one methylation value per person per time point

So you must collapse CpGs first.

2️⃣ Sample-level data (wide format: bdnf_mean_wide, tau_mean_wide)

Here each row is:

one person, with CpGs already averaged

Example:

ID	Group	T0	T1	T2
1	Control	0.075	0.037	NA
2	Piano	0.102	0.060	0.084

Each value (e.g. 0.075) is:

mean methylation across all CpGs for that person at that time

Now when you compute:

mean(bdnf_mean_wide$T0, na.rm=TRUE)
sd(bdnf_mean_wide$T0, na.rm=TRUE)


you are asking:

“How much do people differ in their average BDNF methylation at T0?”

This is exactly what readers expect.

That’s why:

SD ≈ 0.03–0.04

SE ≈ SD / √N_participants

Plots look sensible

Results match your LMM interpretation

Why your interaction plots broke

You originally did this (conceptually):

group_by(Time, Group) %>%
summarise(
  mean = mean(methylation),
  se = sd(methylation) / sqrt(n())
)


But here:

methylation = CpG-level values

n() = number of CpGs × people × time

So SE was:

SD across CpGs ÷ √(CpGs × people)

That is not the SE of the group mean across people.

That’s why BDNF “kind of looked OK” (by coincidence) and TAU didn’t.

The correct mental model (memorize this)
Question	Unit of analysis	Data to use
“What is average gene methylation per person?”	Person	CpGs averaged first
“Do groups differ in gene methylation?”	Person	Sample-level mean
“Does methylation change over time?”	Person	Sample-level mean
“Do CpGs behave differently?”	CpG	CpG-level model

Never mix these levels.

Why your mixed model was actually OK

Your LMM used CpG-level rows, but:

you used random intercepts for person

you interpreted fixed effects at the gene level

That’s defensible if and only if:

you don’t treat CpGs as independent in descriptives

you explain that CpGs are repeated measures within person

Your tables/plots must still reflect person-level summaries.

One sentence you can put in Methods (this resolves everything)

“For descriptive statistics and visualization, methylation values were averaged across all available CpG sites per gene for each participant and time point.”

That sentence legally aligns:

your tables

your plots

your interpretation

your reviewer sanity

Final rule of thumb (print this mentally)

CpGs are repeated measures, not observations.
People are observations.

If you want, next we can:

re-express the LMM more cleanly (random slope for time, CpG as random effect)

decide whether to keep CpG-level modeling or simplify to sample-level LMM

Just tell me which way you want to go.

do this in markdown so i can add it to script



### Boxplots of mean values

```{r}
boxplot(
  list(
    BDNF  = bdnf_mean$Mean,
    Tau   = tau_mean$Mean,
    PSD95 = psd95_mean$Mean
  ),
  xlab = "Marker",
  ylab = "Mean",
  main = "Boxplots of Mean Values"
)

library(ggplot2)

df <- rbind(
  data.frame(Marker = "BDNF",  Mean = bdnf_mean$Mean),
  data.frame(Marker = "Tau",   Mean = tau_mean$Mean),
  data.frame(Marker = "PSD95", Mean = psd95_mean$Mean)
)

ggplot(df, aes(x = Marker, y = Mean)) +
  geom_boxplot() +
  labs(x = "Marker", y = "Mean", title = "Boxplots of Mean Values") +
  theme_classic()

```
```{r}
library(dplyr)

df <- bind_rows(
  bdnf_mean  %>% mutate(Gene = "BDNF"),
  tau_mean   %>% mutate(Gene = "Tau"),
  psd95_mean %>% mutate(Gene = "PSD95")
) %>%
  mutate(
    Gruppe = as.character(Gruppe),
    value  = Mean          # rename raw values to avoid name collision
  )

tab <- df %>%
  group_by(Gene, Gruppe) %>%
  summarise(
    N = sum(is.finite(value)),
    Mean = mean(value, na.rm = TRUE),
    SD   = sd(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(`Mean ± SD` = sprintf("%.3f ± %.3f", Mean, SD)) %>%
  select(Gene, Gruppe, N, `Mean ± SD`) %>%
  arrange(Gene, Gruppe)

# Markdown print (no NULL)
cat("| Gene | Group | N | Mean ± SD |\n")
cat("|------|-------|---:|-----------|\n")
for (i in seq_len(nrow(tab))) {
  cat(sprintf("| %s | %s | %d | %s |\n",
              tab$Gene[i], tab$Gruppe[i], tab$N[i], tab$`Mean ± SD`[i]))
}
```
```{r}
# combine into one long data.frame
df <- rbind(
  data.frame(Gene="BDNF",  Group=bdnf_mean$Gruppe,  value=bdnf_mean$Mean),
  data.frame(Gene="Tau",   Group=tau_mean$Gruppe,   value=tau_mean$Mean),
  data.frame(Gene="PSD95", Group=psd95_mean$Gruppe, value=psd95_mean$Mean)
)

# keep only finite values
df <- df[is.finite(df$value), ]
df$Gene  <- factor(df$Gene,  levels = c("BDNF","Tau","PSD95"))
df$Group <- factor(df$Group)   # e.g., Piano / Controll

# one figure: x = Gene with 2 boxes per gene (split by group), one y-axis
boxplot(
  value ~ Gene + Group,
  data = df,
  xlab = "Gene",
  ylab = "Mean",
  main = "Gene values by group",
  las = 2
)

boxplot(value ~ Gene + Group, data=df,
        col=rep(1:length(levels(df$Group)), times=length(levels(df$Gene))),
        xlab="Gene", ylab="Mean", main="Gene values by group", las=2)

```

```{r}
library(dplyr)
library(ggplot2)

df <- bind_rows(
  bdnf_mean  %>% mutate(Gene = "BDNF"),
  tau_mean   %>% mutate(Gene = "Tau"),
  psd95_mean %>% mutate(Gene = "PSD95")
) %>%
  mutate(
    Gruppe = as.character(Gruppe),
    value  = Mean
  ) %>%
  filter(is.finite(value))

ggplot(df, aes(x = Gene, y = value, fill = Gruppe)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.shape = NA) +
  geom_jitter(aes(color = Gruppe),
              position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
              alpha = 0.4, size = 1.2) +
  labs(x = "Gene", y = "Mean", title = "Gene values by group") +
  theme_classic()

```

### Time mean values
```{r}
mean(bdnf_mean_wide$T0, na.rm = TRUE)
sd(bdnf_mean_wide$T0, na.rm = TRUE)

mean(bdnf_mean_wide$T1, na.rm = TRUE)
sd(bdnf_mean_wide$T1, na.rm = TRUE)

mean(bdnf_mean_wide$T2, na.rm = TRUE)
sd(bdnf_mean_wide$T2, na.rm = TRUE)

mean(tau_mean_wide$T0, na.rm = TRUE)
sd(tau_mean_wide$T0, na.rm = TRUE)

mean(tau_mean_wide$T1, na.rm = TRUE)
sd(tau_mean_wide$T1, na.rm = TRUE)

mean(tau_mean_wide$T2, na.rm = TRUE)
sd(tau_mean_wide$T2, na.rm = TRUE)

mean(psd95_mean_wide$T0, na.rm = TRUE)
sd(psd95_mean_wide$T0, na.rm = TRUE)

mean(psd95_mean_wide$T1, na.rm = TRUE)
sd(psd95_mean_wide$T1, na.rm = TRUE)

mean(psd95_mean_wide$T2, na.rm = TRUE)
sd(psd95_mean_wide$T2, na.rm = TRUE)

summary_table <- data.frame(
  Marker = rep(c("BDNF", "Tau", "PSD95"), each = 3),
  Time = rep(c("T0", "T1", "T2"), times = 3),
  Mean = c(
    mean(bdnf_mean_wide$T0, na.rm = TRUE),
    mean(bdnf_mean_wide$T1, na.rm = TRUE),
    mean(bdnf_mean_wide$T2, na.rm = TRUE),

    mean(tau_mean_wide$T0, na.rm = TRUE),
    mean(tau_mean_wide$T1, na.rm = TRUE),
    mean(tau_mean_wide$T2, na.rm = TRUE),

    mean(psd95_mean_wide$T0, na.rm = TRUE),
    mean(psd95_mean_wide$T1, na.rm = TRUE),
    mean(psd95_mean_wide$T2, na.rm = TRUE)
  ),
  SD = c(
    sd(bdnf_mean_wide$T0, na.rm = TRUE),
    sd(bdnf_mean_wide$T1, na.rm = TRUE),
    sd(bdnf_mean_wide$T2, na.rm = TRUE),

    sd(tau_mean_wide$T0, na.rm = TRUE),
    sd(tau_mean_wide$T1, na.rm = TRUE),
    sd(tau_mean_wide$T2, na.rm = TRUE),

    sd(psd95_mean_wide$T0, na.rm = TRUE),
    sd(psd95_mean_wide$T1, na.rm = TRUE),
    sd(psd95_mean_wide$T2, na.rm = TRUE)
  )
)

summary_table

summary_table$`Mean ± SD` <- sprintf("%.3f ± %.3f",
                                     summary_table$Mean,
                                     summary_table$SD)
summary_table_final <- summary_table[, c("Marker", "Time", "Mean ± SD")]
summary_table_final
```

## Group Differences in Mean Values

```{r}
#| label: group-differences
library(psych)
library(DescTools)

# Calculate the mode for each dataset
Mode(bdnf_mean$Mean, na.rm = TRUE) # Bimodal: two modes at 0 and 0.063
Mode(tau_mean$Mean, na.rm = TRUE) # Mode at 0.066875
Mode(psd95_mean$Mean, na.rm = TRUE) # Bimodal: two modes at 0.02377778 and 0.03177778

Mode(bdnf_median$Median, na.rm = TRUE) # Mode at 0 (92 occurrences)
Mode(tau_median$Median, na.rm = TRUE) # Mode at 0.02 (65 occurrences)
Mode(psd95_median$Median, na.rm = TRUE) # Mode at 0 (317 occurrences)

# Summarize data by group
tapply(bdnf_mean$Mean, bdnf_mean$Gruppe, summary, na.rm = TRUE) # Minimal difference
tapply(bdnf_median$Median, bdnf_median$Gruppe, summary, na.rm = TRUE) # Minimal difference

tapply(tau_mean$Mean, tau_mean$Gruppe, summary, na.rm = TRUE) # No difference
tapply(tau_median$Median, tau_median$Gruppe, summary, na.rm = TRUE) # No difference

tapply(psd95_mean$Mean, psd95_mean$Gruppe, summary, na.rm = TRUE) # No difference
tapply(psd95_median$Median, psd95_median$Gruppe, summary, na.rm = TRUE) # No difference

```

-   Bei BDNF zeigt sich eine leichte Differenz zwischen der Kontrollgruppe (Median: 0.07, Mean: 0.07326) und der Pianogruppe (Median: 0.064, Mean: 0.06598), welche eine niedrigere Metyhlierung (Median: -0.006, Mean: -0.00728) aufwies.
-   Weiterhin zeigen sich für die arithmetischen Mittel von BDNF und PSD95 zwei Modi.
-   Tau war unimodal.

### Boxplots

```{r}
#| label: fig-boxplots-genes
#| fig-cap: "Boxplots of gene expression means by time point and gene."
#| fig-width: 9

library(dplyr)
library(ggplot2)

all_long <- bind_rows(
  bdnf_mean  %>% mutate(Gene = "BDNF"),
  tau_mean   %>% mutate(Gene = "TAU"),
  psd95_mean %>% mutate(Gene = "PSD95")
) %>%
  mutate(
    Zeitpunkt = factor(Zeitpunkt, levels = c("T0","T1","T2")),
    Gene = factor(Gene, levels = c("BDNF","TAU","PSD95")),
    GeneTime = factor(
      paste(Gene, Zeitpunkt, sep = "_"),
      levels = c("BDNF_T0","BDNF_T1","BDNF_T2",
                 "TAU_T0","TAU_T1","TAU_T2",
                 "PSD95_T0","PSD95_T1","PSD95_T2")
    )
  )

x <- library(dplyr)
library(ggplot2)

all_long <- bind_rows(
  bdnf_mean  %>% mutate(Gene = "BDNF"),
  tau_mean   %>% mutate(Gene = "TAU"),
  psd95_mean %>% mutate(Gene = "PSD95")
) %>%
  mutate(
    Zeitpunkt = factor(Zeitpunkt, levels = c("T0","T1","T2")),
    Gene = factor(Gene, levels = c("BDNF","TAU","PSD95")),
    GeneTime = factor(
      paste(Gene, Zeitpunkt, sep = "_"),
      levels = c("BDNF_T0","BDNF_T1","BDNF_T2",
                 "TAU_T0","TAU_T1","TAU_T2",
                 "PSD95_T0","PSD95_T1","PSD95_T2")
    )
  )

x <- ggplot(all_long, aes(x = GeneTime, y = Mean, fill = Gene)) +
  geom_boxplot(na.rm = TRUE, outlier.alpha = 0.6) +
  geom_vline(xintercept = c(3.5, 6.5), linetype = 2) +
  scale_x_discrete(labels = rep(c("T0","T1","T2"), 3)) +
  labs(title = "Mean gene expression by time point", x = "Time (grouped by gene)", y = "Mean", fill = "Gene") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 10),
    plot.title = element_text(face = "bold")
  )

x
```


## Time Differences in Mean Values

```{r}
#| label: time-differences
# Methylierung: PSD95 < Tau < BDNF

tapply(bdnf_mean$Mean, bdnf_mean$Zeitpunkt, mean, na.rm = TRUE) # T0 > T1 < T2 (T0 < T2)
tapply(bdnf_median$Median, bdnf_median$Zeitpunkt, median, na.rm = TRUE) # T0 > T1 < T2 (T0 < T2)

tapply(tau_mean$Mean, tau_mean$Zeitpunkt, mean, na.rm = TRUE) # T0 < T1 = T2
tapply(tau_median$Median, tau_median$Zeitpunkt, median, na.rm = TRUE) # T0 = T1 > T2


# überhaupt keine Differenz
tapply(psd95_mean$Mean, psd95_mean$Zeitpunkt, mean, na.rm = TRUE) # T0 < T1 > T2 (T0 = T2)
tapply(psd95_median$Median, psd95_median$Zeitpunkt, median, na.rm = TRUE) # all = 0 
# noch geringer als Tau und BDNF methyliert
```

-   Die Gene sind allgemein eher gering methyliert (Mean \< 0,08, Median \< 0,04).
-   Dabei ist PSD95 am geringsten und BDNF am höchsten metyhyliert.
-   Bei BDNF zeigt sich, dass sich die Mittelwerte (Arithmetisch und Median) von T0 zu T1 leicht absenken und zu T2 wieder über das Niveau von T0 ansteigen.
-   Bei Tau steigt das arithmetische Mittel der Methylierung leicht von T0 zu T1 und hält das Niveau bei T2.
-   Bei PSD95 verbleibt der Median der Methylierun gleichbleibend bei 0, während das arithmetische Mittel der Methylierung von T0 zu T1 steigt aber bei T2 wieder auf das Ausgangsniveau zurückfällt.

## Important Point and Interval Estimators


```{r}
#| label: descriptives-wide

# Describe the wide format datasets
## BDNF
describe(bdnf_mean_wide)
describe(bdnf_median_wide)

## Tau
describe(tau_mean_wide)
describe(tau_median_wide)

## PSD95
describe(psd95_mean_wide)
describe(psd95_median_wide)
```



## Plotting EDA

### Boxplots
```{r}
eda_bdnf <- cpg
bdnfcols_to_modify <- 5:14

eda_tau <- cpg
taucols_to_modify <- 19:67

eda_psd95 <- cpg
psd95cols_to_modify <- 74:116

# Remove the suffix 'BDNF' from the column names in columns 5 to 14
colnames(eda_bdnf)[bdnfcols_to_modify] <- sub("BDNF$", "", colnames(eda_bdnf)[bdnfcols_to_modify])

# remove "Tau
colnames(eda_tau)[taucols_to_modify] <- sub("Tau$", "", colnames(eda_tau)[taucols_to_modify])

#remove PSD
colnames(eda_psd95)[psd95cols_to_modify] <- sub("PSD95$", "", colnames(eda_psd95)[psd95cols_to_modify])

# View the updated column names
print(colnames(eda_bdnf[c(5:14)]))
print(colnames(eda_tau[c(19:67)]))
print(colnames(eda_psd95[c(72:116)]))

#dev.new(width=14, height=7)
#par(mfrow = c(3, 3))
#layout(matrix(c(0, 1, 0, 2, 2, 2, 3, 3, 3), nrow = 3, byrow = TRUE))
#with()!
```


```{r}
#| label: fig-boxplot-cpg-bdnf

boxplot(eda_bdnf[, c(5:14)], 
        xlab = "CpGs", 
        ylab = "Methylation Level", 
        main = "Boxplot of the BDNF-CpGs")
```


```{r}
#| label: fig-boxplot-cpg-tau

boxplot(eda_tau[, c(19:67)], 
        xlab = "CpGs", 
        ylab = "Methylation Level", 
        main = "Boxplot of the Tau-CpGs")
```


```{r}
#| label: fig-boxplot-cpg-psd95

boxplot(eda_psd95[, c(72:116)], 
        xlab = "CpGs", 
        ylab = "Methylation Level", 
        main = "Boxplot of the PSD95-CpGs")

```

```{r}
#| label: fig-boxplots-cpg-all
#| fig-cap: "Boxplots of CpG methylation levels for BDNF, Tau, and PSD95."
#| fig-width: 10
#| fig-height: 8

op <- par(no.readonly = TRUE)
par(mfrow = c(3, 1), mar = c(6, 4, 3, 1), oma = c(0, 0, 1, 0))

boxplot(eda_bdnf[, 5:14],
        xlab = "CpGs", ylab = "Methylation Level",
        main = "BDNF CpGs",
        las = 2, cex.axis = 0.8)

boxplot(eda_tau[, 19:67],
        xlab = "CpGs", ylab = "Methylation Level",
        main = "Tau CpGs",
        las = 2, cex.axis = 0.6)

boxplot(eda_psd95[, 72:116],
        xlab = "CpGs", ylab = "Methylation Level",
        main = "PSD95 CpGs",
        las = 2, cex.axis = 0.6)

par(op)
```


#### Zoom to methlyated CpGs in Tau and PSD95
siehe "131124_TBM mit Martin.qmd" ab Zeile 1074

### Violinplots
Siehe "Python_violinplots.ipynb"


### Lineplots

```{r}
#| label: helpers-cpg-long
#| include: false

# Helper: convert wide CpG data to long format and keep CpG order

make_long <- function(df, cols) {
levs <- names(df)[cols]
df %>%
pivot_longer(cols = all_of(cols), names_to = "CpG", values_to = "Expression") %>%
mutate(
CpG        = factor(CpG, levels = levs),
Expression = suppressWarnings(as.numeric(Expression))
)
}

# Helper: median + IQR per CpG

plot_median_iqr <- function(long_df, title) {
long_df %>%
group_by(CpG) %>%
summarise(
q25   = quantile(Expression, 0.25, na.rm = TRUE),
median= median(Expression, na.rm = TRUE),
q75   = quantile(Expression, 0.75, na.rm = TRUE),
.groups = "drop"
) %>%
ggplot(aes(CpG)) +
geom_ribbon(aes(ymin = q25, ymax = q75, group = 1), alpha = 0.25) +
geom_line(aes(y = median, group = 1), linewidth = 1) +
geom_point(aes(y = median), size = 2) +
labs(title = title, x = "CpG", y = "Expression") +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Helper: decile profiles (10 typical lines)

plot_deciles <- function(long_df, id_col, title) {
idq <- enquo(id_col)

id_deciles <- long_df %>%
group_by(!!idq) %>%
summarise(id_mean = mean(Expression, na.rm = TRUE), .groups = "drop") %>%
mutate(decile = ntile(id_mean, 10))

decile_cpgl <- long_df %>%
inner_join(id_deciles, by = as_name(idq)) %>%
group_by(decile, CpG) %>%
summarise(median = median(Expression, na.rm = TRUE), .groups = "drop")

ggplot(decile_cpgl,
aes(CpG, median, group = factor(decile), color = factor(decile))) +
geom_line(linewidth = 1) +
geom_point(size = 1.5) +
labs(title = title, x = "CpG", y = "Median expression", color = "Decile") +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# Helper: k-means centroid profiles (~k typical lines)

plot_kmeans <- function(long_df, id_col, k = 8, title = NULL) {
idq <- enquo(id_col)

wide <- long_df %>%
pivot_wider(
id_cols    = !!idq,
names_from = CpG,
values_from= Expression,
values_fn  = ~ median(.x, na.rm = TRUE),
values_fill= NA_real_
)

stopifnot(!any(sapply(wide, is.list)))

mat <- wide %>% select(-!!idq) %>% data.matrix()
for (j in seq_len(ncol(mat))) {
mj <- median(mat[, j], na.rm = TRUE)
mat[is.na(mat[, j]), j] <- mj
}

set.seed(1)
km <- kmeans(scale(mat), centers = k, nstart = 25)

centers <- as.data.frame(km$centers) %>%
rownames_to_column("cluster") %>%
pivot_longer(-cluster, names_to = "CpG", values_to = "Expression") %>%
mutate(
cluster = factor(cluster),
CpG     = factor(CpG, levels = names(wide)[-1])
)

ggplot(centers, aes(CpG, Expression, group = cluster, color = cluster)) +
geom_line(linewidth = 1.1) +
geom_point(size = 1.4) +
labs(
title = ifelse(is.null(title),
"Cluster centroid profiles (scaled)", title),
x = "CpG", y = "Scaled expression", color = "Cluster"
) +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

```

```{r}
#| label: lineplots-cpg


bdnf_cols <- 5:14
tau_cols  <- 19:67
psd_cols  <- 72:116


# Long versions (CpG x person)

bdnf_cpg_long  <- make_long(eda_bdnf,  bdnf_cols)
tau_cpg_long   <- make_long(eda_tau,   tau_cols)
psd95_cpg_long <- make_long(eda_psd95, psd_cols)

# BDNF

plot_median_iqr(bdnf_cpg_long, "BDNF CpGs — median with IQR")
plot_deciles(   bdnf_cpg_long, ID.BDNF,  "BDNF CpGs — ID decile profiles")
plot_kmeans(    bdnf_cpg_long, ID.BDNF,  k = 8,
title = "BDNF CpGs — cluster centroid profiles (scaled)")

# Tau

plot_median_iqr(tau_cpg_long, "Tau CpGs — median with IQR")
plot_deciles(   tau_cpg_long, ID.Tau,   "Tau CpGs — ID decile profiles")
plot_kmeans(    tau_cpg_long, ID.Tau,   k = 8,
title = "Tau CpGs — cluster centroid profiles (scaled)")
# PSD95

plot_median_iqr(psd95_cpg_long, "PSD95 CpGs — median with IQR")
plot_deciles(   psd95_cpg_long, ID.PSD95, "PSD95 CpGs — ID decile profiles")
plot_kmeans(    psd95_cpg_long, ID.PSD95, k = 8,
title = "PSD95 CpGs — cluster centroid profiles (scaled)")

```


### Lineplots by Zeitpunkt

```{r}
#| label: lineplots-zeit

# Column ranges reused

bdnf_cols <- 5:14
tau_cols  <- 19:67
psd_cols  <- 72:116

bdnf_levels <- names(eda_bdnf)[bdnf_cols]
tau_levels  <- names(eda_tau)[tau_cols]
psd_levels  <- names(eda_psd95)[psd_cols]

bdnf_time_long <- eda_bdnf %>%
pivot_longer(all_of(bdnf_cols), names_to = "CpG", values_to = "Expression") %>%
transmute(
Gene       = "BDNF",
ID         = ID.BDNF,
Zeit       = factor(Zeitpunkt.BDNF),
CpG        = factor(CpG, levels = bdnf_levels),
Expression = suppressWarnings(as.numeric(Expression))
)

tau_time_long <- eda_tau %>%
pivot_longer(all_of(tau_cols), names_to = "CpG", values_to = "Expression") %>%
transmute(
Gene       = "Tau",
ID         = ID.Tau,
Zeit       = factor(Zeit.Tau),
CpG        = factor(CpG, levels = tau_levels),
Expression = suppressWarnings(as.numeric(Expression))
)

psd95_time_long <- eda_psd95 %>%
pivot_longer(all_of(psd_cols), names_to = "CpG", values_to = "Expression") %>%
transmute(
Gene       = "PSD95",
ID         = ID.PSD95,
Zeit       = factor(Zeitpunkt.PSD95),
CpG        = factor(CpG, levels = psd_levels),
Expression = suppressWarnings(as.numeric(Expression))
)

all_time_long <- bind_rows(bdnf_time_long, tau_time_long, psd95_time_long)

# A) Zeit summary per gene: median ± IQR

zeit_sum <- all_time_long %>%
group_by(Gene, Zeit, CpG) %>%
summarise(
q25   = quantile(Expression, 0.25, na.rm = TRUE),
median= median(Expression, na.rm = TRUE),
q75   = quantile(Expression, 0.75, na.rm = TRUE),
.groups = "drop"
)

ggplot(zeit_sum, aes(CpG, median, group = Zeit, color = Zeit)) +
geom_line(linewidth = 1) +
geom_point(size = 1.6) +
geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.15, alpha = 0.5) +
facet_wrap(~ Gene, scales = "free_x", ncol = 1) +
labs(title = "CpG profiles by Zeit (median ± IQR)",
x = "CpG", y = "Expression") +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

# B) Spaghetti per ID + LOESS per Zeit

ggplot(all_time_long, aes(CpG, Expression, color = Zeit)) +
geom_line(aes(group = interaction(Gene, Zeit, ID)),
alpha = 0.15, linewidth = 0.4) +
geom_point(alpha = 0.15, size = 0.6) +
geom_smooth(se = FALSE, linewidth = 1.2) +
facet_wrap(~ Gene, scales = "free_x", ncol = 1) +
labs(title = "Changes in CpG expression by Zeit (per gene)",
x = "CpG", y = "Expression") +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## CpG summary table
### BDNF
```{r}
#| label: cpg-summary-bdnf

cpgl_table_bdnf <- bdnf_cpg_long %>%
group_by(CpG) %>%
summarise(
n      = n(),
mean   = mean(Expression, na.rm = TRUE),
sd     = sd(Expression, na.rm = TRUE),
median = median(Expression, na.rm = TRUE),
q25    = quantile(Expression, 0.25, na.rm = TRUE),
q75    = quantile(Expression, 0.75, na.rm = TRUE),
min    = min(Expression, na.rm = TRUE),
max    = max(Expression, na.rm = TRUE),
.groups = "drop"
)

print(cpgl_table_bdnf, n = Inf)

```

### Tau
```{r}
#| label: cpg-summary-tau

cpgl_table_tau <- tau_cpg_long %>%
group_by(CpG) %>%
summarise(
n = n(),
mean = mean(Expression, na.rm = TRUE),
sd = sd(Expression, na.rm = TRUE),
median = median(Expression, na.rm = TRUE),
q25 = quantile(Expression, 0.25, na.rm = TRUE),
q75 = quantile(Expression, 0.75, na.rm = TRUE),
min = min(Expression, na.rm = TRUE),
max = max(Expression, na.rm = TRUE),
.groups = "drop"
)

print(cpgl_table_tau, n = Inf)
```

### PSD95
```{r}
#| label: cpg-summary-psd95

cpgl_table_psd95 <- psd95_cpg_long %>%
group_by(CpG) %>%
summarise(
n      = n(),
mean   = mean(Expression, na.rm = TRUE),
sd     = sd(Expression, na.rm = TRUE),
median = median(Expression, na.rm = TRUE),
q25    = quantile(Expression, 0.25, na.rm = TRUE),
q75    = quantile(Expression, 0.75, na.rm = TRUE),
min    = min(Expression, na.rm = TRUE),
max    = max(Expression, na.rm = TRUE),
.groups = "drop"
)

print(cpgl_table_psd95, n = Inf)

```



## Cross-Tabulations and Chi-Squared Tests
### BDNF

```{r}
#| label: bdnf-wide-clean

bdnf_labels <- read.table(header = TRUE, text = "
name      label
ID        'Participant ID'
Standort  'Site'
Gruppe    'Group'
Sex       'Sex'
Age_start 'Age at baseline (years)'
CogTel    'CogTel total score'
T0        'BDNF at T0'
T1        'BDNF at T1'
T2        'BDNF at T2'
")

bdnf2 <- bdnf_mean_wide %>%
mutate(
ID = as.integer(ID),
Standort = case_when(
Standort %in% c(0, "0") ~ "Site_0",
Standort %in% c(1, "1") ~ "Site_1",
TRUE ~ as.character(Standort)
),
Gruppe = dplyr::recode(Gruppe, "Controll" = "Control"),
Sex    = dplyr::recode(Sex,    "F" = "Female", "M" = "Male"),
across(c(Standort, Gruppe, Sex), ~ fct_infreq(as.factor(.x)))
) %>%
import_labels(bdnf_labels, name_from = "name", label_from = "label") %>%
as_tibble()


```


```{r}
#| label: bdnf-crosstable-group

crosstable(
bdnf2,
c(Age_start, CogTel, T0:T2, Sex),
by = Gruppe,
showNA= "ifany",
total = "both"
) %>%
as_flextable(keep_id = TRUE)

```

```{r}
#| label: bdnf-crosstable-strata

#Long by time, with Time*Group strata

bdnf_time_long <- bdnf2 %>%
pivot_longer(T0, names_to = "Time", values_to = "BDNF") %>%
mutate(
Time = factor(Time, levels = c("T0", "T1", "T2")),
Gruppe = factor(Gruppe),
ID = factor(ID),
Stratum= interaction(Time, Gruppe, drop = TRUE)
)

crosstable(
bdnf_time_long,
c(Age_start, CogTel, Sex, BDNF),
by = Stratum,
showNA= "ifany",
total = "both",
test = TRUE
) %>%
as_flextable()
```

```{r}
#| label: bdnf-sex-chi-square

tab <- with(bdnf2, table(Sex, Gruppe))
tab

chisq.test(tab)                          # chi-square
chi_no_corr <- suppressWarnings(chisq.test(tab, correct = FALSE))
phi <- sqrt(unname(chi_no_corr$statistic) / sum(tab))
phi

prop.test(
x = c(tab["Male", "Control"], tab["Male", "Piano"]),
n = c(colSums(tab)["Control"], colSums(tab)["Piano"])
)

```
### Tau
```{r}
#| label: tau-wide-clean

tau_labels <- read.table(header = TRUE, text = "
name      label
ID        'Participant ID'
Standort  'Site'
Gruppe    'Group'
Sex       'Sex'
Age_start 'Age at baseline (years)'
CogTel    'CogTel total score'
T0        'Tau at T0'
T1        'Tau at T1'
T2        'Tau at T2'
")

tau2 <- tau_mean_wide %>%
mutate(
ID = as.integer(ID),
Standort = case_when(
Standort %in% c(0, "0") ~ "Site_0",
Standort %in% c(1, "1") ~ "Site_1",
TRUE ~ as.character(Standort)
),
Gruppe = dplyr::recode(Gruppe, "Controll" = "Control"),
Sex    = dplyr::recode(Sex,    "F" = "Female", "M" = "Male"),
across(c(Standort, Gruppe, Sex), ~ fct_infreq(as.factor(.x)))
) %>%
import_labels(tau_labels, name_from = "name", label_from = "label") %>%
as_tibble()

```

```{r}
#| label: tau-crosstable-strata

tau_time_long <- tau2 %>%
pivot_longer(T0:T2, names_to = "Time", values_to = "Tau") %>%
mutate(
Time   = factor(Time, levels = c("T0", "T1", "T2")),
Gruppe = factor(Gruppe),
ID     = factor(ID),
Stratum= interaction(Time, Gruppe, drop = TRUE)
)

crosstable(
tau_time_long,
c(Age_start, CogTel, Sex, Tau),
by    = Stratum,
showNA= "ifany",
total = "both",
test  = TRUE
) %>%
as_flextable()

```
### PSD95
```{r}
#| label: psd95-wide-clean

psd95_labels <- read.table(header = TRUE, text = "
name      label
ID        'Participant ID'
Standort  'Site'
Gruppe    'Group'
Sex       'Sex'
Age_start 'Age at baseline (years)'
CogTel    'CogTel total score'
T0        'PSD95 at T0'
T1        'PSD95 at T1'
T2        'PSD95 at T2'
")

psd952 <- psd95_mean_wide %>%
mutate(
ID = as.integer(ID),
Standort = case_when(
Standort %in% c(0, "0") ~ "Site_0",
Standort %in% c(1, "1") ~ "Site_1",
TRUE ~ as.character(Standort)
),
Gruppe = dplyr::recode(Gruppe, "Controll" = "Control"),
Sex    = dplyr::recode(Sex,    "F" = "Female", "M" = "Male"),
across(c(Standort, Gruppe, Sex), ~ fct_infreq(as.factor(.x)))
) %>%
import_labels(psd95_labels, name_from = "name", label_from = "label") %>%
as_tibble()

```

```{r}
#| label: psd95-crosstable-strata

psd95_time_long <- psd952 %>%
pivot_longer(T0:T2, names_to = "Time", values_to = "PSD95") %>%
mutate(
Time   = factor(Time, levels = c("T0", "T1", "T2")),
Gruppe = factor(Gruppe),
ID     = factor(ID),
Stratum= interaction(Time, Gruppe, drop = TRUE)
)

crosstable(
psd95_time_long,
c(Age_start, CogTel, Sex, PSD95),
by    = Stratum,
showNA= "ifany",
total = "both",
test  = TRUE
) %>%
as_flextable()

```