---
title: "Big Despcriptve statistics table"
author: "Jonas J. R. Koberschinski"
---

```{r}
#| echo: true
#| include: false

# Load required libraries
library(tidyverse)
library(tidyr)
library(dplyr)
#library(RColorBrewer) # Commented out unused libraries
#library(lme4)

# Load BDNF data from a text file and display the first few rows
bdnf <- as_tibble(read.table('/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/bdnf.txt', 
                   sep = "\t", 
                   header = TRUE, 
                   row.names = NULL, 
                   quote = ""))
head(bdnf)


# Load Tau data from a text file and display the first few rows
tau <- as_tibble(read.table('/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/tau.txt', 
                   sep = "\t", 
                   header = TRUE, 
                   row.names = NULL, 
                   quote = ""))
head(tau)


# Load PSD95 data from a text file and display the first few rows
psd95 <- as_tibble(read.table('/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/psd95.txt', 
                   sep = "\t", 
                   header = TRUE, 
                   row.names = NULL, 
                   quote = ""))
head(psd95)

# Load CpG data from a text file and display the first few rows
cpg <- as_tibble(read.table(file = '/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/cpg.txt', 
                  sep = "\t", 
                  header = TRUE, 
                  row.names = NULL, 
                  quote = ""))
head(cpg)

```

## Helper functions

```{r}
# Helper: select only columns that actually exist
safe_select <- function(df, cols) {
  cols <- cols[cols %in% names(df)]
  df[, cols, drop = FALSE]
}

# Helper: create a unified smoking variable if either exists
add_smoking <- function(df) {
  if (!("rauchen" %in% names(df)) && ("B14a_rauchen" %in% names(df))) {
    df$rauchen <- df$B14a_rauchen
  }
  df
}

# Baseline covariates you want in Table 1
baseline_covars <- c(
  "Gruppe", "Age_start", "Sex",
  "B09a_groesse", "B09b_gewicht",
  "rauchen", "B14a_rauchen",
  "B15a_alkoholkonsum", "B16_nehmen_sie_medikamente", "B17a_nehmen_sie_drogen",
  "C05_familienstatus", "C06_bildungsgrad"
)

```

## Subset BNDF

```{r}
bdnf <- add_smoking(bdnf)

bdnf_subset <- safe_select(
  bdnf,
  c("ID", "Standort", "Zeitpunkt", "Gruppe", "Sex", "Age_start", "CogTel", baseline_covars)
)

bdnf_methylation <- subset(bdnf, select = c(m148:p020))
bdnf_methylation_mean   <- apply(bdnf_methylation, 1, mean, na.rm = TRUE)
bdnf_methylation_median <- apply(bdnf_methylation, 1, median, na.rm = TRUE)

bdnf_mean <- cbind(bdnf_subset, Mean = bdnf_methylation_mean)
bdnf_median <- cbind(bdnf_subset, Median = bdnf_methylation_median)

# remove problematic IDs
bdnf_mean   <- bdnf_mean[!(bdnf_mean$ID %in% c(192, 193)), ]
bdnf_median <- bdnf_median[!(bdnf_median$ID %in% c(192, 193)), ]
```

## Subset TAU

```{r}
tau <- add_smoking(tau)

tau_subset <- safe_select(
  tau,
  c("ID", "standort_aus_andere_quelle", "Zeit", "Gruppe", "Sex", "Age_start", "CogTel", baseline_covars)
)
colnames(tau_subset)[match(c("standort_aus_andere_quelle","Zeit"), names(tau_subset))] <- c("Standort", "Zeitpunkt")

tau_methylation <- subset(tau, select = c(m331:p27))
tau_methylation_mean   <- apply(tau_methylation, 1, mean, na.rm = TRUE)
tau_methylation_median <- apply(tau_methylation, 1, median, na.rm = TRUE)

tau_mean   <- cbind(tau_subset, Mean = tau_methylation_mean)
tau_median <- cbind(tau_subset, Median = tau_methylation_median)

tau_mean   <- tau_mean[!(tau_mean$ID %in% c(192, 193)), ]
tau_median <- tau_median[!(tau_median$ID %in% c(192, 193)), ]
```

## Subset PSD95

```{r}
psd95 <- add_smoking(psd95)

psd95_subset <- safe_select(
  psd95,
  c("ID", "standort_aus_andere_quelle", "Zeitpunkt", "Gruppe", "Sex", "Age_start", "CogTel", baseline_covars)
)
# rename Standort if present
if ("standort_aus_andere_quelle" %in% names(psd95_subset)) {
  colnames(psd95_subset)[colnames(psd95_subset) == "standort_aus_andere_quelle"] <- "Standort"
}

psd95_methylation <- subset(psd95, select = c(m465:p122))
psd95_methylation_mean   <- apply(psd95_methylation, 1, mean, na.rm = TRUE)
psd95_methylation_median <- apply(psd95_methylation, 1, median, na.rm = TRUE)

psd95_mean   <- cbind(psd95_subset, Mean = psd95_methylation_mean)
psd95_median <- cbind(psd95_subset, Median = psd95_methylation_median)

psd95_mean   <- psd95_mean[!(psd95_mean$ID %in% c(192, 193)), ]
psd95_median <- psd95_median[!(psd95_median$ID %in% c(192, 193)), ]
```

## Check resulting data frames

```{r}
names(bdnf_mean)
names(tau_mean)
names(psd95_mean)

bdnf_mean[1:5, ]
tau_mean[1:5, ]
psd95_mean[1:5, ]
```

```{r}
library(dplyr)

dedup_names_keep_first <- function(df) {
  df[, !duplicated(names(df)), drop = FALSE]
}

bdnf_mean   <- dedup_names_keep_first(bdnf_mean)
tau_mean    <- dedup_names_keep_first(tau_mean)
psd95_mean  <- dedup_names_keep_first(psd95_mean)

bdnf_median <- dedup_names_keep_first(bdnf_median)
tau_median  <- dedup_names_keep_first(tau_median)
psd95_median<- dedup_names_keep_first(psd95_median)
```

```{r}
#| label: big-table-union-demographics-genes
#| results: asis

library(dplyr)

g_piano   <- "Piano"
g_control <- "Controll"

# ---- helpers ----
mean_sd <- function(x) sprintf("%.2f \u00B1 %.1f", mean(x, na.rm=TRUE), sd(x, na.rm=TRUE))
median_iqr <- function(x) {
  q <- quantile(x, c(.25,.5,.75), na.rm=TRUE, names = FALSE)
  sprintf("%.2f (%.2f, %.2f)", q[2], q[1], q[3])
}
min_max <- function(x) sprintf("%.0f / %.0f", min(x, na.rm=TRUE), max(x, na.rm=TRUE))
fmt_missing <- function(x) sprintf("%d (%.1f%%)", sum(is.na(x)), 100 * mean(is.na(x)))
n_nonmiss <- function(x) sum(!is.na(x))

fm_pct <- function(sex_vec) {
  s <- toupper(trimws(sex_vec))
  s[s %in% c("FEMALE")] <- "F"
  s[s %in% c("MALE")]   <- "M"
  tab <- prop.table(table(s)) * 100
  f <- unname(tab["F"]); m <- unname(tab["M"])
  if (is.na(f)) f <- 0
  if (is.na(m)) m <- 0
  sprintf("%.1f / %.1f", f, m)
}

calc_bmi <- function(height_cm, weight_kg) {
  h_m <- height_cm / 100
  bmi <- weight_kg / (h_m^2)
  bmi[is.infinite(bmi)] <- NA
  bmi
}

summ_cont <- function(x) list(
  n     = n_nonmiss(x),
  mean  = mean_sd(x),
  med   = median_iqr(x),
  minmx = min_max(x),
  miss  = fmt_missing(x)
)

# Choose which biomarker summary column to use for "missing samples"
# For *_median data, the column is "Median". (For *_mean it would be "Mean".)
value_col <- "Median"

# ---- UNION master for demographics (unique IDs across all genes) ----
demo_union <- bind_rows(
  bdnf_median  %>% select(ID, Gruppe, Sex, Age_start, B09a_groesse, B09b_gewicht),
  tau_median   %>% select(ID, Gruppe, Sex, Age_start, B09a_groesse, B09b_gewicht),
  psd95_median %>% select(ID, Gruppe, Sex, Age_start, B09a_groesse, B09b_gewicht)
) %>%
  mutate(
    Gruppe = trimws(Gruppe),
    Sex    = trimws(Sex)
  ) %>%
  group_by(ID) %>%
  summarise(
    Gruppe        = first(na.omit(Gruppe)),
    Sex           = first(na.omit(Sex)),
    Age_start     = first(na.omit(Age_start)),
    B09a_groesse  = first(na.omit(B09a_groesse)),
    B09b_gewicht  = first(na.omit(B09b_gewicht)),
    .groups = "drop"
  ) %>%
  mutate(
    BMI = calc_bmi(B09a_groesse, B09b_gewicht)
  )

summ_demo <- function(d) {
  age <- summ_cont(d$Age_start)
  ht  <- summ_cont(d$B09a_groesse)
  wt  <- summ_cont(d$B09b_gewicht)
  bmi <- summ_cont(d$BMI)

  list(
    N = nrow(d),
    age = age,
    ht  = ht,
    wt  = wt,
    bmi = bmi,
    fm = fm_pct(d$Sex),
    sex_miss = fmt_missing(d$Sex)
  )
}

P <- summ_demo(filter(demo_union, Gruppe == g_piano))
C <- summ_demo(filter(demo_union, Gruppe == g_control))
O <- summ_demo(demo_union)

# ---- gene rows: n and missing per timepoint for each marker (LONG format: Zeitpunkt) ----
prep_gene <- function(df) df %>%
  mutate(
    Gruppe    = trimws(Gruppe),
    Zeitpunkt = trimws(as.character(Zeitpunkt))
  )

gene_stats <- function(df, marker, value_col = "Median") {
  df <- prep_gene(df)
  # split by group
  P <- filter(df, Gruppe == g_piano)
  C <- filter(df, Gruppe == g_control)
  O <- df

  # helper: for a given dataset + timepoint, count non-missing and missing for value_col
  one_tp <- function(dat, tp) {
    x <- dat[[value_col]][dat$Zeitpunkt == tp]
    list(n = n_nonmiss(x), miss = fmt_missing(x))
  }

  tps <- c("T0", "T1", "T2")

  list(
    marker = marker,
    P = setNames(lapply(tps, \(tp) one_tp(P, tp)), tps),
    C = setNames(lapply(tps, \(tp) one_tp(C, tp)), tps),
    O = setNames(lapply(tps, \(tp) one_tp(O, tp)), tps)
  )
}

genes <- list(
  gene_stats(bdnf_median,  "BDNF",  value_col),
  gene_stats(tau_median,   "TAU",   value_col),
  gene_stats(psd95_median, "PSD95", value_col)
)

# ---- print unified table ----
cat(paste0(
  "| Characteristic | Piano | Control | Overall |\n",
  "|---|---:|---:|---:|\n",

  "| Participants (N) | ", P$N, " | ", C$N, " | ", O$N, " |\n",

  "| Age non-missing (n) | ", P$age$n, " | ", C$age$n, " | ", O$age$n, " |\n",
  "| Mean age \u00B1 SD (years) | ", P$age$mean, " | ", C$age$mean, " | ", O$age$mean, " |\n",
  "| Median (IQR) age (years) | ", P$age$med, " | ", C$age$med, " | ", O$age$med, " |\n",
  "| Min / Max age (years) | ", P$age$minmx, " | ", C$age$minmx, " | ", O$age$minmx, " |\n",

  "| Height non-missing (n) | ", P$ht$n, " | ", C$ht$n, " | ", O$ht$n, " |\n",
  "| Mean height \u00B1 SD (cm) | ", P$ht$mean, " | ", C$ht$mean, " | ", O$ht$mean, " |\n",
  "| Median (IQR) height (cm) | ", P$ht$med, " | ", C$ht$med, " | ", O$ht$med, " |\n",
  "| Min / Max height (cm) | ", P$ht$minmx, " | ", C$ht$minmx, " | ", O$ht$minmx, " |\n",
  "| Height missing, n (%) | ", P$ht$miss, " | ", C$ht$miss, " | ", O$ht$miss, " |\n",

  "| Weight non-missing (n) | ", P$wt$n, " | ", C$wt$n, " | ", O$wt$n, " |\n",
  "| Mean weight \u00B1 SD (kg) | ", P$wt$mean, " | ", C$wt$mean, " | ", O$wt$mean, " |\n",
  "| Median (IQR) weight (kg) | ", P$wt$med, " | ", C$wt$med, " | ", O$wt$med, " |\n",
  "| Min / Max weight (kg) | ", P$wt$minmx, " | ", C$wt$minmx, " | ", O$wt$minmx, " |\n",
  "| Weight missing, n (%) | ", P$wt$miss, " | ", C$wt$miss, " | ", O$wt$miss, " |\n",

  "| BMI non-missing (n) | ", P$bmi$n, " | ", C$bmi$n, " | ", O$bmi$n, " |\n",
  "| Mean BMI \u00B1 SD (kg/m\u00B2) | ", P$bmi$mean, " | ", C$bmi$mean, " | ", O$bmi$mean, " |\n",
  "| Median (IQR) BMI (kg/m\u00B2) | ", P$bmi$med, " | ", C$bmi$med, " | ", O$bmi$med, " |\n",
  "| Min / Max BMI (kg/m\u00B2) | ", P$bmi$minmx, " | ", C$bmi$minmx, " | ", O$bmi$minmx, " |\n",
  "| BMI missing, n (%) | ", P$bmi$miss, " | ", C$bmi$miss, " | ", O$bmi$miss, " |\n",

  "| Female / Male (%) | ", P$fm, " | ", C$fm, " | ", O$fm, " |\n",
  "| Sex missing, n (%) | ", P$sex_miss, " | ", C$sex_miss, " | ", O$sex_miss, " |\n"
))

for (r in genes) {
  cat(paste0(
    "| ", r$marker, " samples at T0 (n) | ", r$P$T0$n, " | ", r$C$T0$n, " | ", r$O$T0$n, " |\n",
    "| ", r$marker, " samples at T1 (n) | ", r$P$T1$n, " | ", r$C$T1$n, " | ", r$O$T1$n, " |\n",
    "| ", r$marker, " samples at T2 (n) | ", r$P$T2$n, " | ", r$C$T2$n, " | ", r$O$T2$n, " |\n",
    "| ", r$marker, " missing T0, n (%) | ", r$P$T0$miss, " | ", r$C$T0$miss, " | ", r$O$T0$miss, " |\n",
    "| ", r$marker, " missing T1, n (%) | ", r$P$T1$miss, " | ", r$C$T1$miss, " | ", r$O$T1$miss, " |\n",
    "| ", r$marker, " missing T2, n (%) | ", r$P$T2$miss, " | ", r$C$T2$miss, " | ", r$O$T2$miss, " |\n"
  ))
}


```

## Baseline characteristics table (mean ± SD)

Cave! ID 141 is missing because they have no T0 record in any of the three gene datasets (bdnf\_*, tau\_*, psd95\_\*). So:

Your union table (all IDs across all timepoints) = 144

Your NEJM baseline_df filtered to Zeitpunkt == "T0" = 143

The “missing one” is exactly ID 141.

That’s not a bug — it’s a definition issue: “baseline” = T0 excludes people without T0.

### Mean

```{r}
#| label: tbl-nejm
#| tbl-cap: "Baseline characteristics (Mean [SD])"
#| message: false
#| warning: false

library(dplyr)
library(forcats)
library(gtsummary)
library(gt)

# ---- helpers ----
first_nonmiss <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[1]
}

calc_bmi <- function(height_cm, weight_kg) {
  h_m <- height_cm / 100
  bmi <- weight_kg / (h_m^2)
  bmi[is.infinite(bmi)] <- NA
  bmi
}

# ---- UNION baseline dataset (T0, one row per ID) ----
baseline_df <- bind_rows(
  bdnf_mean  %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad"))),
  tau_mean   %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad"))),
  psd95_mean %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad")))
) %>%
  mutate(
    Zeitpunkt = trimws(as.character(Zeitpunkt)),
    Gruppe    = trimws(Gruppe),
    Sex       = trimws(Sex)
  ) %>%
  filter(Zeitpunkt == "T0") %>%
  group_by(ID) %>%
  summarise(across(everything(), first_nonmiss), .groups = "drop") %>%
  mutate(
    BMI = calc_bmi(B09a_groesse, B09b_gewicht),
    rauchen2 = coalesce(rauchen, B14a_rauchen),
    Site = case_when(
      Standort == 0 ~ "Geneva",
      Standort == 1 ~ "Hannover",
      TRUE ~ as.character(Standort)
    )
  ) %>%
  # NEJM: keep only participants with known group assignment
  filter(!is.na(Gruppe)) %>%
  mutate(
    Gruppe = factor(Gruppe, levels = c("Piano", "Controll")),
    Site   = factor(Site, levels = c("Hannover", "Geneva")),  # display order

    # Sex -> Female/Male
    Sex = factor(Sex),
    Sex = fct_recode(Sex,
      "Female" = "F", "Female" = "female", "Female" = "FEMALE",
      "Male"   = "M", "Male"   = "male",   "Male"   = "MALE"
    ),

    rauchen2 = factor(rauchen2),
    B15a_alkoholkonsum = factor(B15a_alkoholkonsum),
    B16_nehmen_sie_medikamente = factor(B16_nehmen_sie_medikamente),
    B17a_nehmen_sie_drogen = factor(B17a_nehmen_sie_drogen),
    C05_familienstatus = factor(C05_familienstatus),
    C06_bildungsgrad = factor(C06_bildungsgrad)
  )

# ---- variables shown (Site placed after Sex) ----
table1_vars <- c(
  "Age_start", "Sex", "Site",
  "B09a_groesse", "B09b_gewicht", "BMI",
  "rauchen2",
  "B15a_alkoholkonsum",
  "B16_nehmen_sie_medikamente",
  "B17a_nehmen_sie_drogen",
  "C05_familienstatus",
  "C06_bildungsgrad",
  "CogTel"
)

t1_mean <- baseline_df %>%
  select(Gruppe, all_of(table1_vars)) %>%
  tbl_summary(
    by = Gruppe,
    label = list(
      Age_start ~ "Age, years",
      Sex ~ "Sex",
      Site ~ "Study site",
      B09a_groesse ~ "Height, cm",
      B09b_gewicht ~ "Weight, kg",
      BMI ~ "Body-mass index, kg/m\u00B2",
      rauchen2 ~ "Smoking status",
      B15a_alkoholkonsum ~ "Alcohol use",
      B16_nehmen_sie_medikamente ~ "Prescription medication use",
      B17a_nehmen_sie_drogen ~ "Illicit drug use",
      C05_familienstatus ~ "Family status",
      C06_bildungsgrad ~ "Educational attainment",
      CogTel ~ "Cognitive score (CogTel)"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    missing = "no"
  ) %>%
  add_overall(last = TRUE) %>%
  add_p() %>%
  bold_labels() %>%
  modify_header(label ~ "**Characteristic**") %>%
  modify_spanning_header(all_stat_cols() ~ "**Group**") %>%
  #modify_caption("**Table 1a. Baseline Characteristics (Mean [SD])**")

t1_mean %>% as_gt()

```

```{r}
#| label: tbl-demo-nejm-mean
#| tbl-cap: "Baseline characteristics (Mean [SD])"
#| message: false
#| warning: false

library(dplyr)
library(gtsummary)
library(gt)

# t1_mean is your gtsummary object (from tbl_summary()...)
gt_mean <- t1_mean %>%
  as_gt() %>%
  tab_options(
    table.width = pct(100),          # <-- makes it responsive / full width
    table.layout = "fixed",          # <-- prevents weird resizing
    data_row.padding = px(2),
    heading.padding = px(2)
  ) %>%
  cols_width(
    everything() ~ pct(20),          # simple equal widths (adjust if needed)
    1 ~ pct(40)                      # Characteristic column wider
  ) %>%
  tab_style(
    style = cell_text(whitespace = "normal"),  # <-- enables wrapping
    locations = cells_body(columns = everything())
  ) %>%
  tab_style(
    style = cell_text(whitespace = "normal"),
    locations = cells_column_labels(everything())
  ) %>%
  tab_source_note(
    md("Values are mean (SD) for continuous variables and no. (%) for categorical variables. \
P values are from the Wilcoxon rank-sum test for continuous variables and Pearson’s chi-squared test or Fisher’s exact test for categorical variables, as appropriate. \
Baseline characteristics were taken from the earliest available assessment (T0 preferred); one participant had no T0 sample (ID 141).")
  )

gt_mean

```

```{r}
#| label: save-tbl-demo-nejm-mean
#| message: false
#| warning: false

gt_tbl <- t1_meanu %>% as_gt()
gt::gtsave(gt_tbl, "_rendered/tbl-demo-nejm-mean.html")
```

## mit ID 144 who had no T0 record in any gene dataset

```{r}
#| label: baseline-df-earliest
#| message: false
#| warning: false

library(dplyr)

first_nonmiss <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[1]
}

calc_bmi <- function(height_cm, weight_kg) {
  h_m <- height_cm / 100
  bmi <- weight_kg / (h_m^2)
  bmi[is.infinite(bmi)] <- NA
  bmi
}

tp_order <- c("T0","T1","T2")

baseline_df <- bind_rows(
  bdnf_mean  %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad"))),
  tau_mean   %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad"))),
  psd95_mean %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad")))
) %>%
  mutate(
    Zeitpunkt = trimws(as.character(Zeitpunkt)),
    Zeitpunkt = factor(Zeitpunkt, levels = tp_order, ordered = TRUE),
    Gruppe    = trimws(Gruppe),
    Sex       = trimws(Sex)
  ) %>%
  arrange(ID, Zeitpunkt) %>%
  group_by(ID) %>%
  slice(1) %>%                 # earliest available row per ID (T0 preferred)
  ungroup() %>%
  group_by(ID) %>%
  summarise(across(everything(), first_nonmiss), .groups = "drop") %>%
  mutate(
    BMI = calc_bmi(B09a_groesse, B09b_gewicht),
    rauchen2 = coalesce(rauchen, B14a_rauchen),
    Site = case_when(
      Standort == 0 ~ "Geneva",
      Standort == 1 ~ "Hannover",
      TRUE ~ as.character(Standort)
    )
  ) %>%
  filter(!is.na(Gruppe))

# sanity check
nrow(baseline_df)
table(baseline_df$Zeitpunkt, useNA = "ifany")  # should show 1 person at T1 (ID 141)
```

### Median

```{r}
#| label: tbl1-nejm-median
#| results: asis
#| message: false
#| warning: false

library(dplyr)
library(forcats)
library(gtsummary)
library(gt)

# baseline_df is rebuilt here so the chunk is standalone.
first_nonmiss <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[1]
}
calc_bmi <- function(height_cm, weight_kg) {
  h_m <- height_cm / 100
  bmi <- weight_kg / (h_m^2)
  bmi[is.infinite(bmi)] <- NA
  bmi
}

baseline_df <- bind_rows(
  bdnf_mean  %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad"))),
  tau_mean   %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad"))),
  psd95_mean %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad")))
) %>%
  mutate(
    Zeitpunkt = trimws(as.character(Zeitpunkt)),
    Gruppe    = trimws(Gruppe),
    Sex       = trimws(Sex)
  ) %>%
  filter(Zeitpunkt == "T0") %>%
  group_by(ID) %>%
  summarise(across(everything(), first_nonmiss), .groups = "drop") %>%
  mutate(
    BMI = calc_bmi(B09a_groesse, B09b_gewicht),
    rauchen2 = coalesce(rauchen, B14a_rauchen),
    Site = case_when(
      Standort == 0 ~ "Geneva",
      Standort == 1 ~ "Hannover",
      TRUE ~ as.character(Standort)
    )
  ) %>%
  filter(!is.na(Gruppe)) %>%
  mutate(
    Gruppe = factor(Gruppe, levels = c("Piano", "Controll")),
    Site   = factor(Site, levels = c("Hannover", "Geneva")),
    Sex = factor(Sex),
    Sex = fct_recode(Sex,
      "Female" = "F", "Female" = "female", "Female" = "FEMALE",
      "Male"   = "M", "Male"   = "male",   "Male"   = "MALE"
    ),
    rauchen2 = factor(rauchen2),
    B15a_alkoholkonsum = factor(B15a_alkoholkonsum),
    B16_nehmen_sie_medikamente = factor(B16_nehmen_sie_medikamente),
    B17a_nehmen_sie_drogen = factor(B17a_nehmen_sie_drogen),
    C05_familienstatus = factor(C05_familienstatus),
    C06_bildungsgrad = factor(C06_bildungsgrad)
  )

table1_vars <- c(
  "Age_start", "Sex", "Site",
  "B09a_groesse", "B09b_gewicht", "BMI",
  "rauchen2",
  "B15a_alkoholkonsum",
  "B16_nehmen_sie_medikamente",
  "B17a_nehmen_sie_drogen",
  "C05_familienstatus",
  "C06_bildungsgrad",
  "CogTel"
)

t1_median <- baseline_df %>%
  select(Gruppe, all_of(table1_vars)) %>%
  tbl_summary(
    by = Gruppe,
    label = list(
      Age_start ~ "Age, years",
      Sex ~ "Sex",
      Site ~ "Study site",
      B09a_groesse ~ "Height, cm",
      B09b_gewicht ~ "Weight, kg",
      BMI ~ "Body-mass index, kg/m\u00B2",
      rauchen2 ~ "Smoking status",
      B15a_alkoholkonsum ~ "Alcohol use",
      B16_nehmen_sie_medikamente ~ "Prescription medication use",
      B17a_nehmen_sie_drogen ~ "Illicit drug use",
      C05_familienstatus ~ "Family status",
      C06_bildungsgrad ~ "Educational attainment",
      CogTel ~ "Cognitive score (CogTel)"
    ),
    statistic = list(
      all_continuous() ~ "{median} [{p25}, {p75}]",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    missing = "no"
  ) %>%
  add_overall(last = TRUE) %>%
  add_p(test = all_continuous() ~ "wilcox.test") %>%  # typical with medians
  bold_labels() %>%
  modify_header(label ~ "**Characteristic**") %>%
  modify_spanning_header(all_stat_cols() ~ "**Group**") %>%
  modify_caption("**Table 1b. Baseline Characteristics (Median [IQR])**")

t1_median %>% as_gt()

```

## Markdown-Nejm

### mean

```{r}
#| label: tbl1-nejm-mean-md
#| results: asis
#| message: false
#| warning: false

library(dplyr)

# ---- helpers ----
first_nonmiss <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) NA else x[1]
}

calc_bmi <- function(height_cm, weight_kg) {
  h_m <- height_cm / 100
  bmi <- weight_kg / (h_m^2)
  bmi[is.infinite(bmi)] <- NA
  bmi
}

fmt_mean_sd <- function(x, digits = 2) {
  m <- mean(x, na.rm = TRUE); s <- sd(x, na.rm = TRUE)
  if (is.nan(m) || is.nan(s)) return("")
  sprintf(paste0("%.",digits,"f (%.",digits,"f)"), m, s)
}

fmt_n_pct_levels <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("")
  tab <- table(x)
  n <- sum(tab)
  paste(sprintf("%s: %d (%.1f%%)", names(tab), as.integer(tab), 100*as.integer(tab)/n), collapse="; ")
}

fmt_p <- function(p) {
  if (is.na(p)) return("")
  if (p < 0.001) "<0.001" else sprintf("%.3f", p)
}

p_ttest <- function(x, g) {
  ok <- !is.na(x) & !is.na(g)
  if (sum(ok) < 3) return(NA_real_)
  tryCatch(t.test(x[ok] ~ g[ok])$p.value, error = function(e) NA_real_)
}

p_wilcox <- function(x, g) {
  ok <- !is.na(x) & !is.na(g)
  if (sum(ok) < 3) return(NA_real_)
  tryCatch(wilcox.test(x[ok] ~ g[ok], exact = FALSE)$p.value, error = function(e) NA_real_)
}

p_cat <- function(x, g) {
  ok <- !is.na(x) & !is.na(g)
  if (sum(ok) < 3) return(NA_real_)
  tab <- table(x[ok], g[ok])
  # Fisher if any expected small / sparse
  if (any(tab < 5)) {
    tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
  } else {
    tryCatch(chisq.test(tab, correct = FALSE)$p.value, error = function(e) NA_real_)
  }
}

tp_order <- c("T0","T1","T2")

# ---- UNION baseline dataset (earliest available per ID) ----
baseline_df <- bind_rows(
  bdnf_mean  %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad"))),
  tau_mean   %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad"))),
  psd95_mean %>% select(any_of(c("ID","Standort","Zeitpunkt","Gruppe","Sex","Age_start","CogTel",
                                "B09a_groesse","B09b_gewicht","rauchen","B14a_rauchen",
                                "B15a_alkoholkonsum","B16_nehmen_sie_medikamente","B17a_nehmen_sie_drogen",
                                "C05_familienstatus","C06_bildungsgrad")))
) %>%
  mutate(
    Zeitpunkt = trimws(as.character(Zeitpunkt)),
    Zeitpunkt = factor(Zeitpunkt, levels = tp_order, ordered = TRUE),
    Gruppe    = trimws(Gruppe),
    Sex       = trimws(Sex)
  ) %>%
  arrange(ID, Zeitpunkt) %>%
  group_by(ID) %>%
  slice(1) %>%
  ungroup() %>%
  group_by(ID) %>%
  summarise(across(everything(), first_nonmiss), .groups = "drop") %>%
  mutate(
    BMI = calc_bmi(B09a_groesse, B09b_gewicht),
    rauchen2 = coalesce(rauchen, B14a_rauchen),
    Site = case_when(
      Standort == 0 ~ "Geneva",
      Standort == 1 ~ "Hannover",
      TRUE ~ as.character(Standort)
    ),
    Sex = toupper(Sex),
    Sex = ifelse(Sex %in% c("F","FEMALE"), "Female",
                 ifelse(Sex %in% c("M","MALE"), "Male", Sex))
  ) %>%
  filter(!is.na(Gruppe)) %>%
  mutate(
    Gruppe = factor(Gruppe, levels = c("Piano","Controll")),
    Site   = factor(Site, levels = c("Hannover","Geneva"))
  )

# ---- build markdown table ----
P <- baseline_df %>% filter(Gruppe == "Piano")
C <- baseline_df %>% filter(Gruppe == "Controll")

rows <- tibble::tibble(
  Characteristic = character(),
  Piano = character(),
  Control = character(),
  Overall = character(),
  `P value` = character()
)

add_row_cont <- function(label, var, digits = 2) {
  x <- baseline_df[[var]]
  g <- baseline_df$Gruppe
  tibble::tibble(
    Characteristic = label,
    Piano   = fmt_mean_sd(P[[var]], digits),
    Control = fmt_mean_sd(C[[var]], digits),
    Overall = fmt_mean_sd(x, digits),
    `P value` = fmt_p(p_ttest(x, g))
  )
}

add_row_cat <- function(label, var) {
  x <- baseline_df[[var]]
  g <- baseline_df$Gruppe
  tibble::tibble(
    Characteristic = label,
    Piano   = fmt_n_pct_levels(P[[var]]),
    Control = fmt_n_pct_levels(C[[var]]),
    Overall = fmt_n_pct_levels(x),
    `P value` = fmt_p(p_cat(x, g))
  )
}

rows <- dplyr::bind_rows(
  tibble::tibble(
    Characteristic = "Participants (N)",
    Piano = as.character(nrow(P)),
    Control = as.character(nrow(C)),
    Overall = as.character(nrow(baseline_df)),
    `P value` = ""
  ),
  add_row_cont("Age, years — mean (SD)", "Age_start", 2),
  add_row_cat("Sex — no. (%)", "Sex"),
  add_row_cat("Study site — no. (%)", "Site"),
  add_row_cont("Height, cm — mean (SD)", "B09a_groesse", 2),
  add_row_cont("Weight, kg — mean (SD)", "B09b_gewicht", 2),
  add_row_cont("Body-mass index, kg/m\u00B2 — mean (SD)", "BMI", 2),
  add_row_cat("Smoking status — no. (%)", "rauchen2"),
  add_row_cat("Alcohol use — no. (%)", "B15a_alkoholkonsum"),
  add_row_cat("Prescription medication use — no. (%)", "B16_nehmen_sie_medikamente"),
  add_row_cat("Illicit drug use — no. (%)", "B17a_nehmen_sie_drogen"),
  add_row_cat("Family status — no. (%)", "C05_familienstatus"),
  add_row_cat("Educational attainment — no. (%)", "C06_bildungsgrad"),
  add_row_cont("Cognitive score (CogTel) — mean (SD)", "CogTel", 2)
)

# Print as markdown
cat("| Characteristic | Piano | Control | Overall | P value |\n")
cat("|---|---:|---:|---:|---:|\n")
apply(rows, 1, function(r) {
  cat("| ", paste(r, collapse=" | "), " |\n", sep="")
})

cat("\n*Values are mean (SD) for continuous variables and no. (%) for categorical variables. P values are from Student’s t test for continuous variables and Pearson’s chi-squared test or Fisher’s exact test for categorical variables, as appropriate. Baseline characteristics were taken from the earliest available assessment (T0 preferred); one participant had no T0 sample (ID 141).*\n")

```

### median

```{r}
#| label: tbl1-nejm-median-md
#| results: asis
#| message: false
#| warning: false

library(dplyr)

# Reuse baseline_df from the previous chunk if it exists; otherwise rebuild quickly.
if (!exists("baseline_df")) stop("Run tbl1-nejm-mean-md first, or copy the baseline_df builder here.")

fmt_median_iqr <- function(x, digits = 2) {
  q <- stats::quantile(x, probs = c(.25,.5,.75), na.rm = TRUE, names = FALSE)
  if (any(is.nan(q))) return("")
  sprintf(paste0("%.",digits,"f [%.",digits,"f, %.",digits,"f]"), q[2], q[1], q[3])
}

fmt_n_pct_levels <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return("")
  tab <- table(x)
  n <- sum(tab)
  paste(sprintf("%s: %d (%.1f%%)", names(tab), as.integer(tab), 100*as.integer(tab)/n), collapse="; ")
}

fmt_p <- function(p) {
  if (is.na(p)) return("")
  if (p < 0.001) "<0.001" else sprintf("%.3f", p)
}

p_wilcox <- function(x, g) {
  ok <- !is.na(x) & !is.na(g)
  if (sum(ok) < 3) return(NA_real_)
  tryCatch(wilcox.test(x[ok] ~ g[ok], exact = FALSE)$p.value, error = function(e) NA_real_)
}

p_cat <- function(x, g) {
  ok <- !is.na(x) & !is.na(g)
  if (sum(ok) < 3) return(NA_real_)
  tab <- table(x[ok], g[ok])
  if (any(tab < 5)) {
    tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
  } else {
    tryCatch(chisq.test(tab, correct = FALSE)$p.value, error = function(e) NA_real_)
  }
}

P <- baseline_df %>% filter(Gruppe == "Piano")
C <- baseline_df %>% filter(Gruppe == "Controll")

add_row_cont <- function(label, var, digits = 2) {
  x <- baseline_df[[var]]
  g <- baseline_df$Gruppe
  tibble::tibble(
    Characteristic = label,
    Piano   = fmt_median_iqr(P[[var]], digits),
    Control = fmt_median_iqr(C[[var]], digits),
    Overall = fmt_median_iqr(x, digits),
    `P value` = fmt_p(p_wilcox(x, g))
  )
}

add_row_cat <- function(label, var) {
  x <- baseline_df[[var]]
  g <- baseline_df$Gruppe
  tibble::tibble(
    Characteristic = label,
    Piano   = fmt_n_pct_levels(P[[var]]),
    Control = fmt_n_pct_levels(C[[var]]),
    Overall = fmt_n_pct_levels(x),
    `P value` = fmt_p(p_cat(x, g))
  )
}

rows <- dplyr::bind_rows(
  tibble::tibble(
    Characteristic = "Participants (N)",
    Piano = as.character(nrow(P)),
    Control = as.character(nrow(C)),
    Overall = as.character(nrow(baseline_df)),
    `P value` = ""
  ),
  add_row_cont("Age, years — median [IQR]", "Age_start", 2),
  add_row_cat("Sex — no. (%)", "Sex"),
  add_row_cat("Study site — no. (%)", "Site"),
  add_row_cont("Height, cm — median [IQR]", "B09a_groesse", 2),
  add_row_cont("Weight, kg — median [IQR]", "B09b_gewicht", 2),
  add_row_cont("Body-mass index, kg/m\u00B2 — median [IQR]", "BMI", 2),
  add_row_cat("Smoking status — no. (%)", "rauchen2"),
  add_row_cat("Alcohol use — no. (%)", "B15a_alkoholkonsum"),
  add_row_cat("Prescription medication use — no. (%)", "B16_nehmen_sie_medikamente"),
  add_row_cat("Illicit drug use — no. (%)", "B17a_nehmen_sie_drogen"),
  add_row_cat("Family status — no. (%)", "C05_familienstatus"),
  add_row_cat("Educational attainment — no. (%)", "C06_bildungsgrad"),
  add_row_cont("Cognitive score (CogTel) — median [IQR]", "CogTel", 2)
)

cat("| Characteristic | Piano | Control | Overall | P value |\n")
cat("|---|---:|---:|---:|---:|\n")
apply(rows, 1, function(r) {
  cat("| ", paste(r, collapse=" | "), " |\n", sep="")
})

cat("\n*Values are median [IQR] for continuous variables and no. (%) for categorical variables. P values are from the Wilcoxon rank-sum test for continuous variables and Pearson’s chi-squared test or Fisher’s exact test for categorical variables, as appropriate. Baseline characteristics were taken from the earliest available assessment (T0 preferred); one participant had no T0 sample (ID 141).*\n")

```
```{r}
#| label: tbl1-nejm-median-knitr
#| tbl-cap: "Table 1. Baseline characteristics (median [IQR])"
#| results: asis
#| message: false
#| warning: false

library(dplyr)
library(knitr)

# ---- helpers ----
dash <- "\u2013"  # en dash

fmt_med_iqr <- function(x, digits = 1) {
  q <- quantile(x, c(.25,.5,.75), na.rm = TRUE, names = FALSE)
  if (any(is.nan(q))) return("")
  sprintf(paste0("%.",digits,"f (%.",digits,"f",dash,"%.",digits,"f)"), q[2], q[1], q[3])
}

fmt_n_pct <- function(n, denom) {
  if (is.na(n) || is.na(denom) || denom == 0) return("")
  sprintf("%d (%.1f)", n, 100*n/denom)
}

fmt_p <- function(p) {
  if (is.na(p)) return("")
  if (p < 0.001) "<0.001" else sprintf("%.3f", p)
}

p_wilcox <- function(x, g) {
  ok <- !is.na(x) & !is.na(g)
  if (sum(ok) < 3) return(NA_real_)
  tryCatch(wilcox.test(x[ok] ~ g[ok], exact = FALSE)$p.value, error = function(e) NA_real_)
}

p_cat <- function(x, g) {
  ok <- !is.na(x) & !is.na(g)
  if (sum(ok) < 3) return(NA_real_)
  tab <- table(x[ok], g[ok])
  # NEJM-style: Fisher when sparse / small expected counts
  if (any(tab < 5)) {
    tryCatch(fisher.test(tab)$p.value, error = function(e) NA_real_)
  } else {
    tryCatch(chisq.test(tab, correct = FALSE)$p.value, error = function(e) NA_real_)
  }
}

# ---- ensure Gruppe levels and site names ----
baseline_df <- baseline_df %>%
  mutate(
    Gruppe = factor(trimws(Gruppe), levels = c("Piano","Controll")),
    Site   = factor(Site, levels = c("Hannover","Geneva")),
    Sex    = factor(Sex, levels = c("Female","Male"))
  )

P <- baseline_df %>% filter(Gruppe == "Piano")
C <- baseline_df %>% filter(Gruppe == "Controll")

nP <- nrow(P); nC <- nrow(C); nO <- nrow(baseline_df)

# ---- recode categorical variables to NEJM-friendly labels ----
# Adjust these label maps if your coding differs.
baseline_df <- baseline_df %>%
  mutate(
    Smoking = case_when(
      rauchen2 %in% c(1, "1", "Yes", "yes", "TRUE", TRUE) ~ "Current smoker",
      rauchen2 %in% c(0, "0", "No", "no", "FALSE", FALSE) ~ "Not current smoker",
      TRUE ~ as.character(rauchen2)
    ),

    Alcohol = dplyr::recode(as.character(B15a_alkoholkonsum),
      `0` = "None",
      `1` = "Occasional",
      `2` = "Regular",
      .default = as.character(B15a_alkoholkonsum)
    ),

    Meds = dplyr::recode(as.character(B16_nehmen_sie_medikamente),
      `0` = "No",
      `1` = "Yes",
      .default = as.character(B16_nehmen_sie_medikamente)
    ),

    Illicit = dplyr::recode(as.character(B17a_nehmen_sie_drogen),
      `0` = "No",
      `1` = "Yes",
      .default = as.character(B17a_nehmen_sie_drogen)
    ),

    Family = dplyr::recode(as.character(C05_familienstatus),
      `1` = "Single",
      `2` = "Married/Partnered",
      `3` = "Divorced",
      `4` = "Widowed",
      .default = as.character(C05_familienstatus)
    ),

    Education = dplyr::recode(as.character(C06_bildungsgrad),
      `1` = "Primary",
      `2` = "Secondary",
      `3` = "Vocational",
      `4` = "Upper secondary",
      `5` = "University",
      `6` = "Postgraduate",
      .default = as.character(C06_bildungsgrad)
    )
  )

# helper: make a block (header row + subrows)
block_cat <- function(df, var, header_label, level_order = NULL, show_levels = NULL) {
  x <- df[[var]]
  g <- df$Gruppe
  p <- fmt_p(p_cat(x, g))

  # levels to show
  levs <- sort(unique(na.omit(as.character(x))))
  if (!is.null(level_order)) levs <- intersect(level_order, levs)
  if (!is.null(show_levels)) levs <- intersect(show_levels, levs)

  # denominators per group for percents (non-missing)
  denomP <- sum(!is.na(P[[var]]))
  denomC <- sum(!is.na(C[[var]]))
  denomO <- sum(!is.na(df[[var]]))

  header <- tibble::tibble(
    Characteristic = paste0("**", header_label, " — no. (%)**"),
    Piano   = "",
    Control = "",
    Overall = "",
    `P value` = p
  )

  subs <- lapply(levs, function(L) {
    nP <- sum(as.character(P[[var]]) == L, na.rm = TRUE)
    nC <- sum(as.character(C[[var]]) == L, na.rm = TRUE)
    nO <- sum(as.character(df[[var]]) == L, na.rm = TRUE)
    tibble::tibble(
      Characteristic = paste0("&nbsp;&nbsp;", L),
      Piano   = fmt_n_pct(nP, denomP),
      Control = fmt_n_pct(nC, denomC),
      Overall = fmt_n_pct(nO, denomO),
      `P value` = ""
    )
  }) |> dplyr::bind_rows()

  bind_rows(header, subs)
}

block_cont <- function(df, var, label, digits = 1) {
  p <- fmt_p(p_wilcox(df[[var]], df$Gruppe))
  tibble::tibble(
    Characteristic = paste0("**", label, "**"),
    Piano   = fmt_med_iqr(P[[var]], digits),
    Control = fmt_med_iqr(C[[var]], digits),
    Overall = fmt_med_iqr(df[[var]], digits),
    `P value` = p
  )
}

# ---- assemble NEJM-style table ----
tbl <- bind_rows(
  tibble::tibble(
    Characteristic = "**Participants**",
    Piano   = sprintf("%d", nP),
    Control = sprintf("%d", nC),
    Overall = sprintf("%d", nO),
    `P value` = ""
  ),

  block_cont(baseline_df, "Age_start", "Age — yr", digits = 1),

  block_cat(baseline_df, "Sex", "Sex", level_order = c("Female","Male")),

  block_cat(baseline_df, "Site", "Study site", level_order = c("Hannover","Geneva")),

  block_cont(baseline_df, "B09a_groesse", "Height — cm", digits = 0),
  block_cont(baseline_df, "B09b_gewicht", "Weight — kg", digits = 0),
  block_cont(baseline_df, "BMI", paste0("Body-mass index — kg/m\u00B2"), digits = 1),

  # Smoking: NEJM usually shows only "Current smoker" row
  block_cat(baseline_df, "Smoking", "Smoking status", show_levels = "Current smoker"),

  block_cat(baseline_df, "Alcohol", "Alcohol use",
            level_order = c("None","Occasional","Regular")),

  # Binary: show only "Yes" row
  block_cat(baseline_df, "Meds", "Prescription medication use", show_levels = "Yes"),
  block_cat(baseline_df, "Illicit", "Illicit drug use", show_levels = "Yes"),

  block_cat(baseline_df, "Family", "Family status",
            level_order = c("Single","Married/Partnered","Divorced","Widowed")),

  block_cat(baseline_df, "Education", "Educational attainment",
            level_order = c("Primary","Secondary","Vocational","Upper secondary","University","Postgraduate")),

  block_cont(baseline_df, "CogTel", "Cognitive score (CogTel)", digits = 1)
)

# Header includes group Ns (NEJM style)
colnames(tbl) <- c(
  "Characteristic",
  sprintf("Piano (N=%d)", nP),
  sprintf("Control (N=%d)", nC),
  sprintf("Overall (N=%d)", nO),
  "P value"
)

kable(tbl, format = "pipe", align = c("l","r","r","r","r"))

cat("\n\n*Values are median (interquartile range) for continuous variables and no. (%) for categorical variables. P values were calculated with the Wilcoxon rank-sum test for continuous variables and Pearson’s chi-square test or Fisher’s exact test, as appropriate, for categorical variables. Percentages may not total 100 because of rounding.*\n")

```

