---
title: "Exploratory Data Analysis (from excel sheets)"
author: "Jonas J.R. Koberschinski"
date: "29-01-2026"
output: html
---

```{r}
#| label: setup
#| include: false
# Global setup: libraries & options

library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(forcats)
library(rlang)

library(psych)
library(psychTools)
library(DescTools)

library(crosstable)
library(flextable)

# Global options
options(dplyr.summarise.inform = FALSE)
```


## Data import

## Data import

```{r}
#| echo: false      # Hide the code
#| include: false   # Exclude the output
#| warning: false   # Suppress warnings
#| message: false   # Suppress messages#

## Set working directory

# setwd("~/Work/Promotionsdatananalyse mit Excel und Martin/Rohdaten/")
# print(getwd())

## Load necessary packages
library(haven)
library(tidyverse)
library(dplyr)
library(readxl)
library(writexl)


BDNF <- read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/BDNF_Neuroplastizität_Alter_CpGs_ohne_m11_und_p36_mit_verhalten_vollständig_umstrukturiert.xlsx")

TAU <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/TAU_CpGs_mit_verhalten_ohne_p61_p59_p34_p29_umstruk.xlsx"))

PSD95 <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/PSD95_CpGs_mit_verhalten_größer_umstrukturiert.xlsx"))
```

## Drop NA
```{r}
#| echo: true
#| include: false

# BDNF
# Remove rows with NA values in the 'methylierung' column
BDNF2 <- BDNF %>% drop_na(methylierung)

# Summarize the number of NA values in the 'methylierung' column (commented out)
# sum(is.na(BDNF$methylierung))

# without NA
# summary(BDNF2)



# TAu

# Remove rows with NA values in the 'methylierung' column
TAU2 <- TAU %>% drop_na(methylation)

# Summarize the number of NA values in the 'methylierung' column (commented out)
# sum(is.na(TAU$methylierung))

# without NA
# summary(TAU2)


# PSD95
# Remove rows with NA values in the 'methylierung' column
PSD952 <- PSD95 %>% drop_na(methylation)

# Summarize the number of NA values in the 'methylierung' column (commented out)
# sum(is.na(BDNF$methylierung))

# without NA
# summary(PSD952)
```
## Remove problematic IDs
```{r}
# Remove problematic IDs (apply consistently)
bad_ids <- c(192, 193)

BDNF2  <- BDNF2  %>% filter(!(ID %in% bad_ids))
TAU2   <- TAU2   %>% filter(!(ID %in% bad_ids))
PSD952 <- PSD952 %>% filter(!(ID %in% bad_ids))

```



```{r}
library(dplyr)
library(tidyr)

BDNF_wide <- BDNF2 %>%
  group_by(ID, Zeitpunkt) %>%
  summarise(
    methylierung = mean(methylierung, na.rm = TRUE),  # oder median(...)
    .groups = "drop"
  ) %>%
  pivot_wider(
    id_cols = ID,
    names_from = Zeitpunkt,
    values_from = methylierung
  )
```


Asked ChatGPT
what is differnt between this restructurinkgn for BDNF_wide
{r}
library(dplyr)
library(tidyr)

BDNF_wide <- BDNF2 %>%
  group_by(ID, Zeitpunkt) %>%
  summarise(
    methylierung = mean(methylierung, na.rm = TRUE),  # oder median(...)
    .groups = "drop"
  ) %>%
  pivot_wider(
    id_cols = ID,
    names_from = Zeitpunkt,
    values_from = methylierung
  )
  
  
  
and that for summary_stats_bdnf in plots-time-methylation.qmd
{r}
#| label: fig-bdnf-timegroup-interaction-wrong
#| fig-cap: "BDNF methylation levels (mean ± SE) by group across time points."
library(dplyr)
library(ggplot2)

# Calculate summary statistics for BDNF
summary_stats_bdnf <- BDNF2 %>%
  group_by(Zeitpunkt, Group) %>%
  summarise(
    mean_methylierung_bdnf = mean(methylierung, na.rm = TRUE),
    se_methylierung_bdnf = sd(methylierung, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Interaction plot with error bars for BDNF
interaction_plot_bdnf <- ggplot(summary_stats_bdnf, aes(x = Zeitpunkt, y = mean_methylierung_bdnf, group = Group, colour = Group)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = mean_methylierung_bdnf - se_methylierung_bdnf,
      ymax = mean_methylierung_bdnf + se_methylierung_bdnf
    ),
    width = 0.2
  ) +
  theme_classic() +
  labs(
    title = "Interaction Plot of BDNF Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_bdnf)


They do two different kinds of “restructuring” (and answer two different questions).

1) BDNF_wide: make one row per subject (ID) with separate columns for T0/T1/T2
BDNF_wide <- BDNF2 %>%
  group_by(ID, Zeitpunkt) %>%
  summarise(methylierung = mean(methylierung, na.rm = TRUE), .groups="drop") %>%
  pivot_wider(id_cols = ID, names_from = Zeitpunkt, values_from = methylierung)

What it’s doing

Grouping level: ID × Zeitpunkt

Meaning: For each person and each time point, you collapse all repeated methylation values (coming from multiple CpGs / plates / replicates) into one number (here: the mean).

Then: you reshape to wide, so you get columns like T0, T1, T2.

Output structure

Rows: IDs

Columns: T0, T1, T2 (plus ID)

Use case: models that expect repeated measures in columns, computing change scores (T2–T0), etc.

Important consequence: This ignores Group entirely, and it averages within-person across whatever duplicates exist.

2) summary_stats_bdnf: compute group-level means + SE for plotting
summary_stats_bdnf <- BDNF2 %>%
  group_by(Zeitpunkt, Group) %>%
  summarise(
    mean_methylierung_bdnf = mean(methylierung, na.rm = TRUE),
    se_methylierung_bdnf = sd(methylierung, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

What it’s doing

Grouping level: Zeitpunkt × Group

Meaning: At each time point, for each group (Control/…), compute:

the mean methylation across all rows in that group-time subset

the standard error, using n() = number of rows in that subset

No widening—it stays in long summary form for ggplot.

Output structure

Rows: (time point × group) combinations (e.g., T0-Control, T0-Intervention, T1-Control, …)

Columns: mean_methylierung_bdnf, se_methylierung_bdnf

Big statistical difference

Because BDNF2 still contains multiple rows per person (multiple CpGs etc.), this summary treats those rows as independent observations:

mean() is effectively a CpG-weighted (or replicate-weighted) mean.

se = sd/sqrt(n()) uses n = number of rows, not number of subjects.

So the SE can look too small and the plot can be misleading if you intended subject-level inference.