---
title: "Exploratory Data Analysis"
author: "Jonas J.R. Koberschinski"
date: "14-11-2024"
output: html
---

## Data import
```{r}
#| echo: true
#| include: false

# load mean&median data
# List of file names (without paths and extension)
file_names <- c(
  "bdnf_mean.txt",
  "bdnf_median.txt",
  
  "tau_mean.txt",
  "tau_median.txt",
  
  "psd95_mean.txt",
  "psd95_median.txt"
)

# Folder where the files are located
subfolder <- "/home/kobi/Work/TBM-R-Analysis/tables/mean&median_tables"

# Loop through the file names, read each table, and assign it to a variable
for (file_name in file_names) {
  # Construct the full file path by combining the subfolder and file name
  file_path <- file.path(subfolder, file_name)
  
  # Remove the .txt extension to get the base name for the variable
  var_name <- tools::file_path_sans_ext(file_name)
  
  # Read the table and assign it to a variable with the same name
  assign(var_name, read.table(file_path, sep = "\t", header = TRUE, row.names = NULL, quote = ""))
}


# load cpg-data
cpg <- read.table(file = '/home/kobi/Work/TBM-R-Analysis/tables/Ausgangstabellen/cpg.txt', sep = "\t", header = TRUE, row.names = NULL, quote = "")


# load wide data
# List of file names (without paths and extension)
file_names_wide <- c(
  "bdnf_mean_wide.txt",
  "bdnf_median_wide.txt",
  
  "tau_mean_wide.txt",
  "tau_median_wide.txt",
  
  "psd95_mean_wide.txt",
  "psd95_median_wide.txt"
)

# Folder where the files are located
subfolder_wide <- "/home/kobi/Work/TBM-R-Analysis/tables/wide_format/"

# Loop through the file names, read each table, and assign it to a variable
for (file_name in file_names_wide) {
  # Construct the full file path by combining the subfolder and file name
  file_path <- file.path(subfolder_wide, file_name)
  
  # Remove the .txt extension to get the base name for the variable
  var_name <- tools::file_path_sans_ext(file_name)
  
  # Read the table and assign it to a variable with the same name
  assign(var_name, read.table(file_path, sep = "\t", header = TRUE, row.names = NULL, quote = ""))
}

```

# EDA (Exploratory Data Analysis)

## Group Differences in Mean Values

```{r}
library(psych)
library(DescTools)

# Calculate the mode for each dataset
Mode(bdnf_mean$Mean, na.rm = TRUE) # Bimodal: two modes at 0 and 0.063
Mode(tau_mean$Mean, na.rm = TRUE) # Mode at 0.066875
Mode(psd95_mean$Mean, na.rm = TRUE) # Bimodal: two modes at 0.02377778 and 0.03177778

Mode(bdnf_median$Median, na.rm = TRUE) # Mode at 0 (92 occurrences)
Mode(tau_median$Median, na.rm = TRUE) # Mode at 0.02 (65 occurrences)
Mode(psd95_median$Median, na.rm = TRUE) # Mode at 0 (317 occurrences)

# Summarize data by group
tapply(bdnf_mean$Mean, bdnf_mean$Gruppe, summary, na.rm = TRUE) # Minimal difference
tapply(bdnf_median$Median, bdnf_median$Gruppe, summary, na.rm = TRUE) # Minimal difference

tapply(tau_mean$Mean, tau_mean$Gruppe, summary, na.rm = TRUE) # No difference
tapply(tau_median$Median, tau_median$Gruppe, summary, na.rm = TRUE) # No difference

tapply(psd95_mean$Mean, psd95_mean$Gruppe, summary, na.rm = TRUE) # No difference
tapply(psd95_median$Median, psd95_median$Gruppe, summary, na.rm = TRUE) # No difference

```

-   Bei BDNF zeigt sich eine leichte Differenz zwischen der Kontrollgruppe (Median: 0.07, Mean: 0.07326) und der Pianogruppe (Median: 0.064, Mean: 0.06598), welche eine niedrigere Metyhlierung (Median: -0.006, Mean: -0.00728) aufwies.
-   Weiterhin zeigen sich für die arithmetischen Mittel von BDNF und PSD95 zwei Modi.
-   Tau war unimodal.

## Time Differences in Mean Values

```{r}
# Methylierung: PSD95 < Tau < BDNF

tapply(bdnf_mean$Mean, bdnf_mean$Zeitpunkt, mean, na.rm = TRUE) # T0 > T1 < T2 (T0 < T2)
tapply(bdnf_median$Median, bdnf_median$Zeitpunkt, median, na.rm = TRUE) # T0 > T1 < T2 (T0 < T2)

tapply(tau_mean$Mean, tau_mean$Zeitpunkt, mean, na.rm = TRUE) # T0 < T1 = T2
tapply(tau_median$Median, tau_median$Zeitpunkt, median, na.rm = TRUE) # T0 = T1 > T2


# überhaupt keine Differenz
tapply(psd95_mean$Mean, psd95_mean$Zeitpunkt, mean, na.rm = TRUE) # T0 < T1 > T2 (T0 = T2)
tapply(psd95_median$Median, psd95_median$Zeitpunkt, median, na.rm = TRUE) # all = 0 
# noch geringer als Tau und BDNF methyliert
```

-   Die Gene sind allgemein eher gering methyliert (Mean \< 0,08, Median \< 0,04).
-   Dabei ist PSD95 am geringsten und BDNF am höchsten metyhyliert.
-   Bei BDNF zeigt sich, dass sich die Mittelwerte (Arithmetisch und Median) von T0 zu T1 leicht absenken und zu T2 wieder über das Niveau von T0 ansteigen.
-   Bei Tau steigt das arithmetische Mittel der Methylierung leicht von T0 zu T1 und hält das Niveau bei T2.
-   Bei PSD95 verbleibt der Median der Methylierun gleichbleibend bei 0, während das arithmetische Mittel der Methylierung von T0 zu T1 steigt aber bei T2 wieder auf das Ausgangsniveau zurückfällt.

## Important Point and Interval Estimators


```{r}
library(psych)
library(psychTools)

# Describe the wide format datasets
## BDNF
describe(bdnf_mean_wide)
describe(bdnf_median_wide)

## Tau
describe(tau_mean_wide)
describe(tau_median_wide)

## PSD95
describe(psd95_mean_wide)
describe(psd95_median_wide)
```

## Cross-Tabulations and Chi-Squared Tests
### BDNF

```{r}
library(dplyr)
library(crosstable)
library(forcats)
library(tibble)

# 1) Variable labels (edit text as you see fit)
bdnf_labels <- read.table(header = TRUE, text = "
name      label
ID        'Participant ID'
Standort  'Site'
Gruppe    'Group'
Sex       'Sex'
Age_start 'Age at baseline (years)'
CogTel    'CogTel total score'
T0        'BDNF at T0'
T1        'BDNF at T1'
T2        'BDNF at T2'
")

# 2) Recode, factor, and attach labels
bdnf2 <- bdnf_mean_wide %>%
  mutate(
    ID = as.integer(ID),

    # Tidy up site coding (adjust labels to your real site names if known)
    Standort = case_when(
      Standort %in% c(0, "0") ~ "Site_0",
      Standort %in% c(1, "1") ~ "Site_1",
      TRUE ~ as.character(Standort)
    ),

    # Fix the group typo and keep as readable factor
    Gruppe = recode(Gruppe, "Controll" = "Control"),

    # Expand sex codes
    Sex = recode(Sex, "F" = "Female", "M" = "Male"),

    # Make categorical variables factors (ordered by frequency)
    across(c(Standort, Gruppe, Sex), ~ fct_infreq(as.factor(.x)))
  ) %>%
  import_labels(bdnf_labels, name_from = "name", label_from = "label") %>%
  as_tibble()

# Optional: quick crosstable like in your mtcars example
# crosstable(bdnf2, by = Gruppe, cols = c(Sex, Age_start, CogTel, T0:T2), total = "both")

```


```{r}
# Crosstable of numeric + categorical vars by Group, keeping participant ID
crosstable(
  bdnf2,
  c(Age_start, CogTel, T0:T2, Sex),
  by = Gruppe,
  showNA = "ifany",
  total = "both"
) %>%
  as_flextable(keep_id = TRUE)


```

```{r}
# 1) Build the 2x2 table
tab <- with(bdnf2, table(Sex, Gruppe))
tab
#        Gruppe
# Sex     Control Piano
# Female       46    37
# Male         31    30

# 2) Chi-squared test of independence (adds Yates correction by default)
chisq.test(tab)

# If any expected cell < 5, prefer Fisher's exact:
# fisher.test(tab)

# 3) Effect size (phi for 2x2) = sqrt(chi2 / N)
chi <- suppressWarnings(chisq.test(tab, correct = FALSE))
phi <- sqrt(unname(chi$statistic) / sum(tab))
phi

# 4) Two-sample proportion test for "male share" as an alternative view
prop.test(x = c(tab["Male","Control"], tab["Male","Piano"]),
          n = c(colSums(tab)["Control"], colSums(tab)["Piano"]))
```

```{r}
bdnf_long2 <- bdnf_long %>%
  mutate(Stratum = interaction(Time, drop = TRUE))  # e.g., T0.Control, T0.Piano, ...

crosstable(
  bdnf_long,
  c(Age_start, CogTel, Sex, Gruppe, BDNF),
  by = Stratum,
  showNA = "ifany",
  total = "both",
  test = TRUE # adds global test across strata
) %>%
  as_flextable()

```
### Tau
```{r}
# 1) Variable labels (edit text as you see fit)
tau_labels <- read.table(header = TRUE, text = "
name      label
ID        'Participant ID'
Standort  'Site'
Gruppe    'Group'
Sex       'Sex'
Age_start 'Age at baseline (years)'
CogTel    'CogTel total score'
T0        'Tau at T0'
T1        'Tau at T1'
T2        'Tau at T2'
")

# 2) Recode, factor, and attach labels
tau2 <- tau_mean_wide %>%
  mutate(
    ID = as.integer(ID),

    # Tidy up site coding (adjust labels to your real site names if known)
    Standort = case_when(
      Standort %in% c(0, "0") ~ "Site_0",
      Standort %in% c(1, "1") ~ "Site_1",
      TRUE ~ as.character(Standort)
    ),

    # Fix the group typo and keep as readable factor
    Gruppe = recode(Gruppe, "Controll" = "Control"),

    # Expand sex codes
    Sex = recode(Sex, "F" = "Female", "M" = "Male"),

    # Make categorical variables factors (ordered by frequency)
    across(c(Standort, Gruppe, Sex), ~ fct_infreq(as.factor(.x)))
  ) %>%
  import_labels(tau_labels, name_from = "name", label_from = "label") %>%
  as_tibble()

```

```{r}

tau_long <- tau2 %>%
  pivot_longer(T0:T2, names_to = "Time", values_to = "TAU") %>%
  mutate(Time = factor(Time, levels = c("T0","T1","T2")),
         Gruppe = factor(Gruppe),
         ID = factor(ID))
tau_long2 <- tau_long %>%
  mutate(Stratum = interaction(Time, Gruppe, drop = TRUE))  # e.g., T0.Control, T0.Piano, ...

crosstable(
  tau_long2,
  c(Age_start, CogTel, Sex, TAU),
  by = Stratum,
  showNA = "ifany",
  total = "both",
  test = TRUE # adds global test across strata
) %>%
  as_flextable()

```
### PSD95
```{r}
# 1) Variable labels (edit text as you see fit)
psd95_labels <- read.table(header = TRUE, text = "
name      label
ID        'Participant ID'
Standort  'Site'
Gruppe    'Group'
Sex       'Sex'
Age_start 'Age at baseline (years)'
CogTel    'CogTel total score'
T0        'PSD95 at T0'
T1        'PSD95 at T1'
T2        'PSD95 at T2'
")

# 2) Recode, factor, and attach labels
psd952 <- psd95_mean_wide %>%
  mutate(
    ID = as.integer(ID),

    # Tidy up site coding (adjust labels to your real site names if known)
    Standort = case_when(
      Standort %in% c(0, "0") ~ "Site_0",
      Standort %in% c(1, "1") ~ "Site_1",
      TRUE ~ as.character(Standort)
    ),

    # Fix the group typo and keep as readable factor
    Gruppe = recode(Gruppe, "Controll" = "Control"),

    # Expand sex codes
    Sex = recode(Sex, "F" = "Female", "M" = "Male"),

    # Make categorical variables factors (ordered by frequency)
    across(c(Standort, Gruppe, Sex), ~ fct_infreq(as.factor(.x)))
  ) %>%
  import_labels(tau_labels, name_from = "name", label_from = "label") %>%
  as_tibble()

```

```{r}

psd95_long <- psd952 %>%
  pivot_longer(T0:T2, names_to = "Time", values_to = "PSD95") %>%
  mutate(Time = factor(Time, levels = c("T0","T1","T2")),
         Gruppe = factor(Gruppe),
         ID = factor(ID))
psd95_long2 <- psd95_long %>%
  mutate(Stratum = interaction(Time, Gruppe, drop = TRUE))  # e.g., T0.Control, T0.Piano, ...

crosstable(
  psd95_long2,
  c(Age_start, CogTel, Sex, PSD95),
  by = Stratum,
  showNA = "ifany",
  total = "both",
  test = TRUE # adds global test across strata
) %>%
  as_flextable()

```

## Normality Tests





## Plotting EDA

### Boxplots
```{r}
eda_bdnf <- cpg
bdnfcols_to_modify <- 5:14

eda_tau <- cpg
taucols_to_modify <- 19:67

eda_psd95 <- cpg
psd95cols_to_modify <- 74:116

# Remove the suffix 'BDNF' from the column names in columns 5 to 14
colnames(eda_bdnf)[bdnfcols_to_modify] <- sub("BDNF$", "", colnames(eda_bdnf)[bdnfcols_to_modify])

# remove "Tau
colnames(eda_tau)[taucols_to_modify] <- sub("Tau$", "", colnames(eda_tau)[taucols_to_modify])

#remove PSD
colnames(eda_psd95)[psd95cols_to_modify] <- sub("PSD95$", "", colnames(eda_psd95)[psd95cols_to_modify])

# View the updated column names
print(colnames(eda_bdnf[c(5:14)]))
print(colnames(eda_tau[c(19:67)]))
print(colnames(eda_psd95[c(72:116)]))

#dev.new(width=14, height=7)
#par(mfrow = c(3, 3))
#layout(matrix(c(0, 1, 0, 2, 2, 2, 3, 3, 3), nrow = 3, byrow = TRUE))
#with()!
boxplot(eda_bdnf[, c(5:14)], 
        xlab = "CpGs", 
        ylab = "Methylation Level", 
        main = "Boxplot of the BDNF-CpGs")

boxplot(eda_tau[, c(19:67)], 
        xlab = "CpGs", 
        ylab = "Methylation Level", 
        main = "Boxplot of the Tau-CpGs")

boxplot(eda_psd95[, c(72:116)], 
        xlab = "CpGs", 
        ylab = "Methylation Level", 
        main = "Boxplot of the PSD95-CpGs")

```
#### Zoom to methlyated CpGs in Tau and PSD95
siehe "131124_TBM mit Martin.qmd" ab Zeile 1074

### Violinplots
Siehe "Python_violinplots.ipynb"


### Lineplots

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(rlang)
library(tibble)

# ==== column ranges (yours) ====
bdnf_cols <- 5:14
tau_cols  <- 19:67
psd_cols  <- 72:116

# ==== helper: build long df & keep CpG order ====
make_long <- function(df, cols) {
  levs <- names(df)[cols]
  df %>%
    pivot_longer(cols = all_of(cols), names_to = "CpG", values_to = "Expression") %>%
    mutate(CpG = factor(CpG, levels = levs),
           Expression = suppressWarnings(as.numeric(Expression)))
}

# ==== helper: median + IQR ====
plot_median_iqr <- function(long_df, title) {
  long_df %>%
    group_by(CpG) %>%
    summarise(
      q25 = quantile(Expression, 0.25, na.rm = TRUE),
      median = median(Expression, na.rm = TRUE),
      q75 = quantile(Expression, 0.75, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    ggplot(aes(CpG)) +
    geom_ribbon(aes(ymin = q25, ymax = q75, group = 1), alpha = 0.25) +
    geom_line(aes(y = median, group = 1), linewidth = 1) +
    geom_point(aes(y = median), size = 2) +
    labs(title = title, x = "CpG", y = "Expression") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# ==== helper: decile profiles (10 typical lines) ====
plot_deciles <- function(long_df, id_col, title) {
  idq <- enquo(id_col)

  id_deciles <- long_df %>%
    group_by(!!idq) %>%
    summarise(id_mean = mean(Expression, na.rm = TRUE), .groups = "drop") %>%
    mutate(decile = ntile(id_mean, 10))

  decile_cpgl <- long_df %>%
    inner_join(id_deciles, by = as_name(idq)) %>%
    group_by(decile, CpG) %>%
    summarise(median = median(Expression, na.rm = TRUE), .groups = "drop")

  ggplot(decile_cpgl,
         aes(CpG, median, group = factor(decile), color = factor(decile))) +
    geom_line(linewidth = 1) + geom_point(size = 1.5) +
    labs(title = title, x = "CpG", y = "Median expression", color = "Decile") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# ==== helper: k-means centroid profiles (~k typical lines) ====
plot_kmeans <- function(long_df, id_col, k = 8, title = NULL) {
  idq <- enquo(id_col)

  wide <- long_df %>%
    pivot_wider(
      id_cols    = !!idq,
      names_from = CpG,
      values_from= Expression,
      values_fn  = ~ median(.x, na.rm = TRUE),
      values_fill= NA_real_
    )
  stopifnot(!any(sapply(wide, is.list)))

  mat <- wide %>% select(-!!idq) %>% data.matrix()
  for (j in seq_len(ncol(mat))) {
    mj <- median(mat[, j], na.rm = TRUE)
    mat[is.na(mat[, j]), j] <- mj
  }

  set.seed(1)
  km <- kmeans(scale(mat), centers = k, nstart = 25)

  centers <- as.data.frame(km$centers) %>%
    rownames_to_column("cluster") %>%
    pivot_longer(-cluster, names_to = "CpG", values_to = "Expression") %>%
    mutate(cluster = factor(cluster),
           CpG = factor(CpG, levels = names(wide)[-1]))

  ggplot(centers, aes(CpG, Expression, group = cluster, color = cluster)) +
    geom_line(linewidth = 1.1) + geom_point(size = 1.4) +
    labs(title = ifelse(is.null(title), "Cluster centroid profiles (scaled)", title),
         x = "CpG", y = "Scaled expression", color = "Cluster") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}

# ==== build long data for all three panels ====
bdnf_long  <- make_long(eda_bdnf,   bdnf_cols)
tau_long   <- make_long(eda_tau,    tau_cols)
psd95_long <- make_long(eda_psd95,  psd_cols)

# ==== BDNF plots ====
plot_median_iqr(bdnf_long,  "BDNF CpGs — median with IQR")
plot_deciles(   bdnf_long,  ID.BDNF,  "BDNF CpGs — ID decile profiles")
plot_kmeans(    bdnf_long,  ID.BDNF,  k = 8, title = "BDNF CpGs — cluster centroid profiles (scaled)")

# ==== Tau plots ====
plot_median_iqr(tau_long,   "Tau CpGs — median with IQR")
plot_deciles(   tau_long,   ID.Tau,   "Tau CpGs — ID decile profiles")
plot_kmeans(    tau_long,   ID.Tau,   k = 8, title = "Tau CpGs — cluster centroid profiles (scaled)")

# ==== PSD95 plots ====
plot_median_iqr(psd95_long, "PSD95 CpGs — median with IQR")
plot_deciles(   psd95_long, ID.PSD95, "PSD95 CpGs — ID decile profiles")
plot_kmeans(    psd95_long, ID.PSD95, k = 8, title = "PSD95 CpGs — cluster centroid profiles (scaled)")

```


### Lineplots by Zeitpunkt

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Column ranges you already use
bdnf_cols <- 5:14
tau_cols  <- 19:67
psd_cols  <- 72:116

# --- 1) Build a unified long dataset (keeps CpG order for each gene) ---
bdnf_levels <- names(eda_bdnf)[bdnf_cols]
tau_levels  <- names(eda_tau)[tau_cols]
psd_levels  <- names(eda_psd95)[psd_cols]

bdnf_long <- eda_bdnf %>%
  pivot_longer(all_of(bdnf_cols), names_to = "CpG", values_to = "Expression") %>%
  transmute(Gene = "BDNF",
            ID   = ID.BDNF,
            Zeit = factor(Zeitpunkt.BDNF),
            CpG  = factor(CpG, levels = bdnf_levels),
            Expression = suppressWarnings(as.numeric(Expression)))

tau_long <- eda_tau %>%
  pivot_longer(all_of(tau_cols), names_to = "CpG", values_to = "Expression") %>%
  transmute(Gene = "Tau",
            ID   = ID.Tau,
            Zeit = factor(Zeit.Tau),
            CpG  = factor(CpG, levels = tau_levels),
            Expression = suppressWarnings(as.numeric(Expression)))

psd95_long <- eda_psd95 %>%
  pivot_longer(all_of(psd_cols), names_to = "CpG", values_to = "Expression") %>%
  transmute(Gene = "PSD95",
            ID   = ID.PSD95,
            Zeit = factor(Zeitpunkt.PSD95),
            CpG  = factor(CpG, levels = psd_levels),
            Expression = suppressWarnings(as.numeric(Expression)))

all_long <- bind_rows(bdnf_long, tau_long, psd95_long)

# --- A) Zeit summary per gene: median ± IQR (faceted by Gene) ---
zeit_sum <- all_long %>%
  group_by(Gene, Zeit, CpG) %>%
  summarise(
    q25   = quantile(Expression, 0.25, na.rm = TRUE),
    median= median(Expression,         na.rm = TRUE),
    q75   = quantile(Expression, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(zeit_sum, aes(CpG, median, group = Zeit, color = Zeit)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.15, alpha = 0.5) +
  facet_wrap(~ Gene, scales = "free_x", ncol = 1) +
  labs(title = "CpG profiles by Zeit (median ± IQR)",
       x = "CpG", y = "Expression") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# --- B) Spaghetti per ID (faint) + LOESS per Zeit (bold), faceted by Gene ---
ggplot(all_long, aes(CpG, Expression, color = Zeit)) +
  geom_line(aes(group = interaction(Gene, Zeit, ID)), alpha = 0.15, linewidth = 0.4) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_smooth(se = FALSE, linewidth = 1.2) +   # one smooth per Zeit (because color=Zeit)
  facet_wrap(~ Gene, scales = "free_x", ncol = 1) +
  labs(title = "Changes in CpG expression by Zeit (per gene)",
       x = "CpG", y = "Expression") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## CpG summary table
### BDNF
```{r}
library(dplyr)
# Summary table for CpGs
cpgl_table_bdnf <- eda_bdnf_long %>%
  group_by(CpG) %>%
  summarise(
    n      = n(),
    mean   = mean(Expression, na.rm = TRUE),
    sd     = sd(Expression, na.rm = TRUE),
    median = median(Expression, na.rm = TRUE),
    q25    = quantile(Expression, 0.25, na.rm = TRUE),
    q75    = quantile(Expression, 0.75, na.rm = TRUE),
    min    = min(Expression, na.rm = TRUE),
    max    = max(Expression, na.rm = TRUE),
    .groups = "drop"
  )
print(cpgl_table_bdnf, n = Inf)
```

### Tau
```{r}
library(dplyr)
# Summary table for CpGs
cpgl_table_tau <- eda_tau_long %>%
  group_by(CpG) %>%
  summarise(
    n      = n(),
    mean   = mean(Expression, na.rm = TRUE),
    sd     = sd(Expression, na.rm = TRUE),
    median = median(Expression, na.rm = TRUE),
    q25    = quantile(Expression, 0.25, na.rm = TRUE),
    q75    = quantile(Expression, 0.75, na.rm = TRUE),
    min    = min(Expression, na.rm = TRUE),
    max    = max(Expression, na.rm = TRUE),
    .groups = "drop"
  )
print(cpgl_table_tau, n = Inf)
```

### PSD95
```{r}
library(dplyr)
# Summary table for CpGs
cpgl_table_psd95 <- eda_psd95_long %>%
  group_by(CpG) %>%
  summarise(
    n      = n(),
    mean   = mean(Expression, na.rm = TRUE),
    sd     = sd(Expression, na.rm = TRUE),
    median = median(Expression, na.rm = TRUE),
    q25    = quantile(Expression, 0.25, na.rm = TRUE),
    q75    = quantile(Expression, 0.75, na.rm = TRUE),
    min    = min(Expression, na.rm = TRUE),
    max    = max(Expression, na.rm = TRUE),
    .groups = "drop"
  )
print(cpgl_table_psd95, n = Inf)
```