---
title: "Linear models"
author: "Jonas J.R. Koberschinski"
date: "14-11-2024"
---
```{r}
#| include: false
# List of file names (without paths and extension)
file_names <- c(
  "bdnf_mean_T2T0.txt",
  "bdnf_mean_T1T0.txt",
  "bdnf_mean_T2T1.txt",
  
  "bdnf_median_T2T0.txt",
  "bdnf_median_T1T0.txt",
  "bdnf_median_T2T1.txt",
  
  "tau_mean_T2T0.txt",
  "tau_mean_T1T0.txt",
  "tau_mean_T2T1.txt",
  
  "tau_median_T2T0.txt",
  "tau_median_T1T0.txt",
  "tau_median_T2T1.txt",
  
  "psd95_mean_T2T0.txt",
  "psd95_mean_T1T0.txt",
  "psd95_mean_T2T1.txt",
  
  "psd95_median_T2T0.txt",
  "psd95_median_T1T0.txt",
  "psd95_median_T2T1.txt"
)

# Folder where the files are located
subfolder <- "/Users/kobi/Desktop/Promotionsdatananalyse mit Excel und Martin/Promotion (Quarto Manuscript)/tables/timepoint_tables"

# Loop through the file names, read each table, and assign it to a variable
for (file_name in file_names) {
  # Construct the full file path by combining the subfolder and file name
  file_path <- file.path(subfolder, file_name)
  
  # Remove the .txt extension to get the base name for the variable
  var_name <- tools::file_path_sans_ext(file_name)
  
  # Read the table and assign it to a variable with the same name
  assign(var_name, read.table(file_path, sep = "\t", header = TRUE, row.names = NULL, quote = ""))
}

# Example: access the 'bdnf_mean_T2T0' table
head(bdnf_mean_T2T0)

```

## LM: Regression analysis (Bivariate/multivariate)

### BDNF

#### T2-T0

##### All variables and step()

```{r}
# lm mean
lm_bdnf_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_mean_T2T0)

summary(lm_bdnf_mean_T2T0)
anova(lm_bdnf_mean_T2T0)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T2T0)
step(lm_bdnf_mean_T2T0) # call: lm(formula = T2_T0 ~ T0, data = bdnf_mean_T2T0)

# lm median
lm_bdnf_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_median_T2T0)

summary(lm_bdnf_median_T2T0)
anova(lm_bdnf_median_T2T0)
plot(lm_bdnf_median_T2T0)
step(lm_bdnf_median_T2T0) # call: lm(formula = T2_T0 ~ T0, data = bdnf_median_T2T0)
```

##### AIC model recommendation

```{r}
#mean
bdnf_aic_mean_T2T0 <- lm(formula = T2_T0 ~ T0, data = bdnf_mean_T2T0)
summary(bdnf_aic_mean_T2T0)
anova(bdnf_aic_mean_T2T0)
par(mfrow = c(2,2))
plot(bdnf_aic_mean_T2T0)

#median
bdnf_aic_median_T2T0 <- lm(formula = T2_T0 ~ T0, data = bdnf_median_T2T0)
summary(bdnf_aic_median_T2T0)
anova(bdnf_aic_median_T2T0)
plot(bdnf_aic_median_T2T0)
```

#### T1 - T0

##### All variables and step()

```{r}
# mean
lm_bdnf_mean_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_mean_T1T0)

summary(lm_bdnf_mean_T1T0)
anova(lm_bdnf_mean_T1T0)
step(lm_bdnf_mean_T1T0)

# median
lm_bdnf_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_median_T1T0)

summary(lm_bdnf_median_T1T0)
anova(lm_bdnf_median_T1T0)
step(lm_bdnf_median_T1T0)
```

##### AIC model recommendation

```{r}
#mean
bdnf_aic_mean_T1T0 <- lm(formula = T1_T0 ~ Gruppe + T0, data = bdnf_mean_T1T0)
summary(bdnf_aic_mean_T1T0)
anova(bdnf_aic_mean_T1T0)
par(mfrow = c(2,2))
plot(bdnf_aic_mean_T1T0)

#median
bdnf_aic_median_T1T0 <- lm(formula = T1_T0 ~ Gruppe + CogTel + T0, data = bdnf_median_T1T0)
summary(bdnf_aic_median_T1T0)
anova(bdnf_aic_median_T1T0)
plot(bdnf_aic_median_T1T0)
```

#### T2 - T1

##### All variables and step()

```{r}
# with T1
# mean
lm_bdnf_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = bdnf_mean_T2T1)

summary(lm_bdnf_mean_T2T1)
anova(lm_bdnf_mean_T2T1)
step(lm_bdnf_mean_T2T1)

# median

lm_bdnf_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = bdnf_median_T2T1)

summary(lm_bdnf_median_T2T1)
anova(lm_bdnf_median_T2T1)
step(lm_bdnf_median_T2T1)
```

##### AIC model recommendation

```{r}
#mean
bdnf_aic_mean_T2T1 <- lm(formula = T2_T1 ~ T1, data = bdnf_mean_T2T1)
summary(bdnf_aic_mean_T2T1)
anova(bdnf_aic_mean_T2T1)
par(mfrow = c(2,2))
plot(bdnf_aic_mean_T2T1)

#median
bdnf_aic_median_T2T1 <- lm(formula = T2_T1 ~ T1, data = bdnf_median_T2T1)
summary(bdnf_aic_median_T2T1)
anova(bdnf_aic_median_T2T1)
plot(bdnf_aic_median_T2T1)
```

#### Plot models

```{r}
#?termplot

library(visreg)

#T2-T0
visreg(bdnf_aic_mean_T2T0, type="conditional", gg = TRUE)
visreg(bdnf_aic_median_T2T0, gg = TRUE)

#T1-T0
visreg(bdnf_aic_mean_T1T0, "Gruppe", gg = TRUE)
visreg(bdnf_aic_median_T1T0, "CogTel", by = "Gruppe", gg = TRUE)

#T2-T1
visreg(bdnf_aic_mean_T2T1, gg = TRUE)
visreg(bdnf_aic_median_T2T1, gg = TRUE)
```

### Tau

#### T2-T0

##### All variables and step()

```{r}
# lm mean
lm_tau_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_mean_T2T0)

summary(lm_tau_mean_T2T0)
anova(lm_tau_mean_T2T0)

par(mfrow = c(2,2))
plot(lm_tau_mean_T2T0)
step(lm_tau_mean_T2T0)

# lm median
lm_tau_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_median_T2T0)

summary(lm_tau_median_T2T0)
anova(lm_tau_median_T2T0)
plot(lm_tau_median_T2T0)
step(lm_tau_median_T2T0)

```

##### AIC model recommendation

```{r}
# mean
tau_aic_mean_T2T0 <- lm(formula = T2_T0 ~ T0, data = tau_mean_T2T0)
summary(tau_aic_mean_T2T0)
anova(tau_aic_mean_T2T0)
par(mfrow = c(2,2))
plot(tau_aic_mean_T2T0)

#median
tau_aic_median_T2T0 <- lm(formula = T2_T0 ~ CogTel + T0, data = tau_median_T2T0)
summary(tau_aic_median_T2T0)
anova(tau_aic_median_T2T0)
plot(tau_aic_median_T2T0)
```

#### T1 - T0

##### All variables and step()

```{r}
# mean
lm_tau_mean_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_mean_T1T0)

summary(lm_tau_mean_T1T0)
anova(lm_tau_mean_T1T0)
step(lm_tau_mean_T1T0)

# median

lm_tau_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_median_T1T0)

summary(lm_tau_median_T1T0)
anova(lm_tau_median_T1T0)
step(lm_tau_median_T1T0)
```

##### AIC model recommendation

```{r}
#mean
tau_aic_mean_T1T0 <- lm(formula = T1_T0 ~ Sex + Age_start + T0, data = tau_mean_T1T0)
summary(tau_aic_mean_T1T0 )
anova(tau_aic_mean_T1T0 )
par(mfrow = c(2,2))
plot(tau_aic_mean_T1T0 )

#median
tau_aic_median_T1T0 <- lm(formula = T1_T0 ~ Standort + T0, data = tau_median_T1T0)
summary(tau_aic_median_T1T0)
anova(tau_aic_median_T1T0)
plot(tau_aic_median_T1T0)
```

#### T2 - T1

##### All variables and step()

```{r}
# with T1
# mean
lm_tau_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = tau_mean_T2T1)

summary(lm_tau_mean_T2T1)
anova(lm_tau_mean_T2T1)
step(lm_tau_mean_T2T1)

# median

lm_tau_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = tau_median_T2T1)

summary(lm_tau_median_T2T1)
anova(lm_tau_median_T2T1)
step(lm_tau_median_T2T1)
```

##### AIC model recommendation

```{r}
#mean
tau_aic_mean_T2T1 <- lm(formula = T2_T1 ~ T1, data = tau_mean_T2T1)
summary(tau_aic_mean_T2T1)
anova(tau_aic_mean_T2T1)
par(mfrow = c(2,2))
plot(tau_aic_mean_T2T1)

#median
tau_aic_median_T2T1 <- lm(formula = T2_T1 ~ Standort + CogTel + T1, data = tau_median_T2T1)
summary(tau_aic_median_T2T1)
anova(tau_aic_median_T2T1)
plot(tau_aic_median_T2T1)
```

#### Plot models

```{r}
#T2-T0
visreg(tau_aic_mean_T2T0, gg = T)
visreg(tau_aic_median_T2T0, gg = T)

#T1-T0
visreg(tau_aic_mean_T1T0, gg = F)
visreg(tau_aic_median_T1T0, gg = F)

#T2-T1
visreg(tau_aic_mean_T2T1, gg = F)
visreg(tau_aic_median_T2T1, gg = F)
```

### PSD95

#### T2-T0

##### All variables and step()

```{r}
# lm mean
lm_psd95_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_mean_T2T0)

summary(lm_psd95_mean_T2T0)
anova(lm_psd95_mean_T2T0)

par(mfrow = c(2,2))
plot(lm_psd95_mean_T2T0)
step(lm_psd95_mean_T2T0)

# lm median
lm_psd95_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_median_T2T0)

summary(lm_psd95_median_T2T0)
anova(lm_psd95_median_T2T0)
plot(lm_psd95_median_T2T0)
step(lm_psd95_median_T2T0) # call: lm(formula = T2_T0 ~ T0, data = bdnf_median_T2T0)
```

##### AIC model recommendation

```{r}
#mean
psd95_aic_mean_T2T0 <- lm(formula = T2_T0 ~ T0, data = psd95_mean_T2T0)
summary(psd95_aic_mean_T2T0)
anova(psd95_aic_mean_T2T0)
par(mfrow = c(2,2))
plot(psd95_aic_mean_T2T0)

#median
psd95_aic_median_T2T0 <- lm(formula = T2_T0 ~ Sex + T0, data = psd95_median_T2T0)
summary(psd95_aic_median_T2T0)
anova(psd95_aic_median_T2T0)
plot(psd95_aic_median_T2T0)
```

#### T1 - T0

##### All variables and step()

```{r}
# mean
lm_psd95_mean_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_mean_T1T0)

summary(lm_psd95_mean_T1T0)
anova(lm_psd95_mean_T1T0)
step(lm_psd95_mean_T1T0)

# median
lm_psd95_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_median_T1T0)

summary(lm_psd95_median_T1T0)
anova(lm_psd95_median_T1T0)
step(lm_psd95_median_T1T0)
```

##### AIC model recommendation

```{r}
#mean
psd95_aic_mean_T1T0 <- lm(formula = T1_T0 ~ Standort + Sex + T0, data = psd95_mean_T1T0)
summary(psd95_aic_mean_T1T0)
anova(psd95_aic_mean_T1T0)
par(mfrow = c(2,2))
plot(psd95_aic_mean_T1T0)

#median
psd95_aic_median_T1T0 <- lm(formula = T1_T0 ~ Standort + Sex + CogTel + T0, data = psd95_median_T1T0)
summary(psd95_aic_median_T1T0)
anova(psd95_aic_median_T1T0)
plot(psd95_aic_median_T1T0)
```

#### T2 - T1

##### All variables and step()

```{r}
# with T1
# mean
lm_psd95_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = psd95_mean_T2T1)

summary(lm_psd95_mean_T2T1)
anova(lm_psd95_mean_T2T1)
step(lm_psd95_mean_T2T1)

# median

lm_psd95_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = psd95_median_T2T1)

summary(lm_psd95_median_T2T1)
anova(lm_psd95_median_T2T1)
step(lm_psd95_median_T2T1)
```

##### AIC model recommendation

```{r}
#mean
psd95_aic_mean_T2T1 <- lm(formula = T2_T1 ~ Gruppe + T1, data = psd95_mean_T2T1)
summary(psd95_aic_mean_T2T1)
anova(psd95_aic_mean_T2T1)
par(mfrow = c(2,2))
plot(psd95_aic_mean_T2T1)

#median
psd95_aic_median_T2T1 <- lm(formula = T2_T1 ~ Sex + T1, data = psd95_median_T2T1)
summary(psd95_aic_median_T2T1)
anova(psd95_aic_median_T2T1)
plot(psd95_aic_median_T2T1)
```

#### Plot models

```{r}
#T2-T0
visreg(psd95_aic_mean_T2T0, gg = T)
visreg(psd95_aic_median_T2T0, gg = T)

#T1-T0
visreg(psd95_aic_mean_T1T0, gg = F)
visreg(psd95_aic_median_T1T0, gg = F)

#T2-T1
visreg(psd95_aic_mean_T2T1, gg = F)
visreg(psd95_aic_median_T2T1, gg = F)
```

