---
title: "Linear_models_short_101224"
author: "Jonas J.R. Koberschinski"
date: "14-11-2024"
output: html
---

## Data import
```{r}
#| echo: true
#| include: false

# List of file names (without paths and extension)
file_names <- c(
  "bdnf_mean_T2T0.txt",
  "bdnf_mean_T1T0.txt",
  "bdnf_mean_T2T1.txt",
  
  "bdnf_median_T2T0.txt",
  "bdnf_median_T1T0.txt",
  "bdnf_median_T2T1.txt",
  
  "tau_mean_T2T0.txt",
  "tau_mean_T1T0.txt",
  "tau_mean_T2T1.txt",
  
  "tau_median_T2T0.txt",
  "tau_median_T1T0.txt",
  "tau_median_T2T1.txt",
  
  "psd95_mean_T2T0.txt",
  "psd95_mean_T1T0.txt",
  "psd95_mean_T2T1.txt",
  
  "psd95_median_T2T0.txt",
  "psd95_median_T1T0.txt",
  "psd95_median_T2T1.txt"
)

# Folder where the files are located
subfolder <- "/Users/kobi/Desktop/Promotionsdatananalyse mit Excel und Martin/Promotion (Quarto Manuscript)/tables/timepoint_tables"

# Loop through the file names, read each table, and assign it to a variable
for (file_name in file_names) {
  # Construct the full file path by combining the subfolder and file name
  file_path <- file.path(subfolder, file_name)
  
  # Remove the .txt extension to get the base name for the variable
  var_name <- tools::file_path_sans_ext(file_name)
  
  # Read the table and assign it to a variable with the same name
  assign(var_name, read.table(file_path, sep = "\t", header = TRUE, row.names = NULL, quote = ""))
}

# Example: access the 'bdnf_mean_T2T0' table
head(bdnf_mean_T2T0)

```

## LM: Regression analysis (Bivariate/multivariate)

### BDNF

#### T2-T0

##### mean
```{r}
lm_bdnf_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_mean_T2T0)

summary(lm_bdnf_mean_T2T0)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T2T0)
```

##### median

```{r}
lm_bdnf_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_median_T2T0)

summary(lm_bdnf_median_T2T0)

par(mfrow = c(2,2))
plot(lm_bdnf_median_T2T0)
```


#### T1 - T0
##### mean

```{r}
lm_bdnf_mean_T1T0 <- lm(T1_T0 ~ as.factor(Standort) + as.factor(Gruppe) + as.factor(Sex) + Age_start + CogTel + T0, 
                  data = bdnf_mean_T1T0)

summary(lm_bdnf_mean_T1T0)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T1T0)
```

Gruppendifferenz!

##### median

Gruppendifferenz bei mean nicht mehr vorhanden beim Median
```{r}
lm_bdnf_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_median_T1T0)

summary(lm_bdnf_median_T1T0)

par(mfrow = c(2,2))
plot(lm_bdnf_median_T1T0)
```


#### T2 - T1
##### mean

```{r}
lm_bdnf_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = bdnf_mean_T2T1)

summary(lm_bdnf_mean_T2T1)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T2T1)
```

##### median

```{r}
lm_bdnf_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = bdnf_median_T2T1)

summary(lm_bdnf_median_T2T1)

par(mfrow = c(2,2))
plot(lm_bdnf_median_T2T1)
```


### Tau

#### T2-T0

##### mean
```{r}
lm_tau_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_mean_T2T0)

summary(lm_tau_mean_T2T0)


par(mfrow = c(2,2))
plot(lm_tau_mean_T2T0)
```

##### median

```{r}
lm_tau_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_median_T2T0)

summary(lm_tau_median_T2T0)

par(mfrow = c(2,2))
plot(lm_tau_median_T2T0)
```


#### T1 - T0

##### mean
```{r}
lm_tau_mean_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_mean_T1T0)

summary(lm_tau_mean_T1T0)

par(mfrow = c(2,2))
plot(lm_tau_mean_T1T0)
```

##### median

```{r}
lm_tau_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_median_T1T0)

summary(lm_tau_median_T1T0)

par(mfrow = c(2,2))
plot(lm_tau_median_T1T0)
```
Standortdifferenz im Median?

#### T2 - T1
##### mean

```{r}
lm_tau_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = tau_mean_T2T1)

summary(lm_tau_mean_T2T1)

par(mfrow = c(2,2))
plot(lm_tau_mean_T2T1)
```

##### median

```{r}
lm_tau_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = tau_median_T2T1)

summary(lm_tau_median_T2T1)

par(mfrow = c(2,2))
plot(lm_tau_median_T2T1)
```

### PSD95

#### T2-T0

##### mean

```{r}
lm_psd95_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_mean_T2T0)

summary(lm_psd95_mean_T2T0)

par(mfrow = c(2,2))
plot(lm_psd95_mean_T2T0)
```

##### median

```{r}
lm_psd95_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_median_T2T0)

summary(lm_psd95_median_T2T0)

par(mfrow = c(2,2))
plot(lm_psd95_median_T2T0)
```

#### T1 - T0

##### mean

```{r}
lm_psd95_mean_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_mean_T1T0)

summary(lm_psd95_mean_T1T0)
```
Standortdifferenz im Mean?
##### median

```{r}
lm_psd95_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_median_T1T0)

summary(lm_psd95_median_T1T0)
```
Keine Standortdifferenz im Mean? Dafür Geschlechtsdifferenz?!

#### T2 - T1
##### mean
```{r}
lm_psd95_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = psd95_mean_T2T1)

summary(lm_psd95_mean_T2T1)

par(mfrow = c(2,2))
plot(lm_psd95_mean_T2T1)

```

##### median

```{r}
lm_psd95_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = psd95_median_T2T1)

summary(lm_psd95_median_T2T1)

par(mfrow = c(2,2))
plot(lm_psd95_median_T2T1)
```





