---
title: "Linear Models (Short Version)"
author: "Jonas J.R. Koberschinski"
date: "21-11-2024"
last_modified: "2026-02-02"
output: html
---

## Data import
```{r}
#| echo: true
#| include: false

# List of file names (without paths and extension)
file_names <- c(
  "bdnf_mean_T2T0.txt",
  "bdnf_mean_T1T0.txt",
  "bdnf_mean_T2T1.txt",
  
  "bdnf_median_T2T0.txt",
  "bdnf_median_T1T0.txt",
  "bdnf_median_T2T1.txt",
  
  "tau_mean_T2T0.txt",
  "tau_mean_T1T0.txt",
  "tau_mean_T2T1.txt",
  
  "tau_median_T2T0.txt",
  "tau_median_T1T0.txt",
  "tau_median_T2T1.txt",
  
  "psd95_mean_T2T0.txt",
  "psd95_mean_T1T0.txt",
  "psd95_mean_T2T1.txt",
  
  "psd95_median_T2T0.txt",
  "psd95_median_T1T0.txt",
  "psd95_median_T2T1.txt"
)

# Folder where the files are located
subfolder <- "/home/kobi/Work/TBM-R-Analysis/tables/timepoint_tables"

# Loop through the file names, read each table, and assign it to a variable
for (file_name in file_names) {
  # Construct the full file path by combining the subfolder and file name
  file_path <- file.path(subfolder, file_name)
  
  # Remove the .txt extension to get the base name for the variable
  var_name <- tools::file_path_sans_ext(file_name)
  
  # Read the table and assign it to a variable with the same name
  assign(var_name, read.table(file_path, sep = "\t", header = TRUE, row.names = NULL, quote = ""))
}

# Example: access the 'bdnf_mean_T2T0' table
head(bdnf_mean_T2T0)

```

## LM: Regression analysis (Bivariate/multivariate)

With dependent variable: Change in biomarker levels between timepoints (T2-T0, T1-T0, T2-T1)
Without stepwise selection of predictors through step().

### BDNF

#### T2-T0
##### 
```{r}
y <- bdnf_mean_T2T0$T2_T0
n <- length(y)

par(mfrow = c(1,2))
hist(y, breaks = 30, main = "T2_T0: Histogram", xlab = "T2_T0", freq = FALSE)
lines(density(y), lwd = 2)
qqnorm(y, main = "QQ plot: T2_T0"); qqline(y, col = "red")
par(mfrow = c(1,1))

shapiro.test(y)
```

##### mean
```{r}
lm_bdnf_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_mean_T2T0)

summary(lm_bdnf_mean_T2T0)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T2T0)
res <- residuals(lm_bdnf_mean_T2T0)

shapiro.test(res) 
```

##### median

```{r}
lm_bdnf_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_median_T2T0)

summary(lm_bdnf_median_T2T0)

par(mfrow = c(2,2))
plot(lm_bdnf_median_T2T0)
```


#### T1 - T0
##### mean

```{r}
lm_bdnf_mean_T1T0 <- lm(T1_T0 ~ as.factor(Standort) + as.factor(Gruppe) + as.factor(Sex) + Age_start + CogTel + T0, 
                  data = bdnf_mean_T1T0)

summary(lm_bdnf_mean_T1T0)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T1T0)
```

Gruppendifferenz!

##### median

Gruppendifferenz bei mean nicht mehr vorhanden beim Median
```{r}
lm_bdnf_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_median_T1T0)

summary(lm_bdnf_median_T1T0)

par(mfrow = c(2,2))
plot(lm_bdnf_median_T1T0)
```


#### T2 - T1
##### mean

```{r}
lm_bdnf_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = bdnf_mean_T2T1)

summary(lm_bdnf_mean_T2T1)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T2T1)
```

##### median

```{r}
lm_bdnf_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = bdnf_median_T2T1)

summary(lm_bdnf_median_T2T1)

par(mfrow = c(2,2))
plot(lm_bdnf_median_T2T1)
```


### Tau

#### T2-T0

##### mean
```{r}
lm_tau_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_mean_T2T0)

summary(lm_tau_mean_T2T0)


par(mfrow = c(2,2))
plot(lm_tau_mean_T2T0)
```

##### median

```{r}
lm_tau_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_median_T2T0)

summary(lm_tau_median_T2T0)

par(mfrow = c(2,2))
plot(lm_tau_median_T2T0)
```


#### T1 - T0

##### mean
```{r}
lm_tau_mean_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_mean_T1T0)

summary(lm_tau_mean_T1T0)

par(mfrow = c(2,2))
plot(lm_tau_mean_T1T0)
```

##### median

```{r}
lm_tau_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_median_T1T0)

summary(lm_tau_median_T1T0)

par(mfrow = c(2,2))
plot(lm_tau_median_T1T0)
```
Standortdifferenz im Median?

#### T2 - T1
##### mean

```{r}
lm_tau_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = tau_mean_T2T1)

summary(lm_tau_mean_T2T1)

par(mfrow = c(2,2))
plot(lm_tau_mean_T2T1)
```

##### median

```{r}
lm_tau_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = tau_median_T2T1)

summary(lm_tau_median_T2T1)

par(mfrow = c(2,2))
plot(lm_tau_median_T2T1)
```

### PSD95

#### T2-T0

##### mean

```{r}
lm_psd95_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_mean_T2T0)

summary(lm_psd95_mean_T2T0)

par(mfrow = c(2,2))
plot(lm_psd95_mean_T2T0)
```

##### median

```{r}
lm_psd95_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_median_T2T0)

summary(lm_psd95_median_T2T0)

par(mfrow = c(2,2))
plot(lm_psd95_median_T2T0)
```

#### T1 - T0

##### mean

```{r}
lm_psd95_mean_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_mean_T1T0)

summary(lm_psd95_mean_T1T0)
```
Standortdifferenz im Mean?
##### median

```{r}
lm_psd95_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_median_T1T0)

summary(lm_psd95_median_T1T0)
```
Keine Standortdifferenz im Mean? Dafür Geschlechtsdifferenz?!

#### T2 - T1
##### mean
```{r}
lm_psd95_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = psd95_mean_T2T1)

summary(lm_psd95_mean_T2T1)

par(mfrow = c(2,2))
plot(lm_psd95_mean_T2T1)

```

##### median

```{r}
lm_psd95_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = psd95_median_T2T1)

summary(lm_psd95_median_T2T1)

par(mfrow = c(2,2))
plot(lm_psd95_median_T2T1)
```

```{r}
#| echo: true
#| message: false
#| warning: false

library(report)

# Collect ONLY the mean models
mean_models <- list(
  bdnf_T2T0  = lm_bdnf_mean_T2T0,
  bdnf_T1T0  = lm_bdnf_mean_T1T0,
  bdnf_T2T1  = lm_bdnf_mean_T2T1,
  
  tau_T2T0   = lm_tau_mean_T2T0,
  tau_T1T0   = lm_tau_mean_T1T0,
  tau_T2T1   = lm_tau_mean_T2T1,
  
  psd95_T2T0 = lm_psd95_mean_T2T0,
  psd95_T1T0 = lm_psd95_mean_T1T0,
  psd95_T2T1 = lm_psd95_mean_T2T1
)

# ---- 1) Narrative reports (most common use) ----
# Print each report under a header
for (nm in names(mean_models)) {
  cat("\n\n", "## Mean model:", nm, "\n\n")
  cat(as.character(report(mean_models[[nm]])))
  cat("\n")
}

# ---- 2) Optional: Tables produced by {report} ----
# If you want model tables as well (one table per model), see:
# for (nm in names(mean_models)) {
#   cat("\n\n", "### Table:", nm, "\n\n")
#   print(report_table(mean_models[[nm]]))
# }

```

```{r}

library(report)

all_reports <- lapply(names(mean_models), function(nm) {
  paste0("## Mean model: ", nm, "\n\n", as.character(report(mean_models[[nm]])))
})

cat(paste(all_reports, collapse = "\n\n"))
```

```{r}
# Check the actual coefficient table
summary(lm_bdnf_mean_T1T0)$coefficients

# Confirm which columns are numeric and their ranges
str(bdnf_mean_T1T0[, c("Age_start", "CogTel", "T0", "T1_T0")])
summary(bdnf_mean_T1T0[, c("Age_start", "CogTel", "T0", "T1_T0")])

# Check correlations to see what "should" be the strong negative predictor
cor(bdnf_mean_T1T0[, c("T1_T0", "T0", "CogTel")], use = "pairwise.complete.obs")

```

```{r}
#| echo: true
#| message: false
#| warning: false

library(broom)
library(dplyr)
library(ggplot2)

# Put your mean models in a named list (as before)
mean_models <- list(
  bdnf_T2T0  = lm_bdnf_mean_T2T0,
  bdnf_T1T0  = lm_bdnf_mean_T1T0,
  bdnf_T2T1  = lm_bdnf_mean_T2T1,
  tau_T2T0   = lm_tau_mean_T2T0,
  tau_T1T0   = lm_tau_mean_T1T0,
  tau_T2T1   = lm_tau_mean_T2T1,
  psd95_T2T0 = lm_psd95_mean_T2T0,
  psd95_T1T0 = lm_psd95_mean_T1T0,
  psd95_T2T1 = lm_psd95_mean_T2T1
)

coef_df <- bind_rows(lapply(names(mean_models), function(nm) {
  tidy(mean_models[[nm]], conf.int = TRUE) |>
    mutate(model = nm)
}))

# Helper columns: biomarker and contrast from the model name
coef_df <- coef_df |>
  mutate(
    biomarker = sub("_.*", "", model),
    contrast  = sub("^[^_]+_", "", model)
  )

# Group term can appear as "GruppePiano" or "as.factor(Gruppe)Piano"
group_terms <- c("GruppePiano", "as.factor(Gruppe)Piano")

group_df <- coef_df |>
  filter(term %in% group_terms) |>
  mutate(term_pretty = "Group: Piano vs Control")
```


```{r}
#| label: fig-lm-coef-plots
#| fig-cap: "Coefficient plots …"
#| echo: false
#| message: false
#| warning: false

lm_coef_plots <- ggplot(group_df, aes(x = contrast, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_pointrange() +
  facet_wrap(~ biomarker, scales = "free_y") +
  labs(
    title = "Group effect (Piano vs Control) across time contrasts",
    x = "Time contrast",
    y = "Coefficient estimate (95% CI)"
  )

lm_coef_plots
```

```{r}
baseline_terms <- c("T0", "T1")

baseline_df <- coef_df |>
  filter(term %in% baseline_terms) |>
  mutate(term_pretty = paste0("Baseline: ", term))

ggplot(baseline_df, aes(x = contrast, y = estimate, ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_pointrange() +
  facet_grid(biomarker ~ term_pretty, scales = "free_y") +
  labs(
    title = "Baseline (T0/T1) effect across time contrasts",
    x = "Time contrast",
    y = "Coefficient estimate (95% CI)"
  )

```

