---
title: "Linear models (PostPCA)"
author: "Jonas J.R. Koberschinski"
date: "20-10-2025"
lastmod: "25-10-2025"
output: html
---

## Data import

```{r}
#| echo: true
#| include: false

# List of file names (without paths and extension)
file_names <- c(
  "bdnf_mean_T2T0.txt",
  "bdnf_mean_T1T0.txt",
  "bdnf_mean_T2T1.txt",
  
  "bdnf_median_T2T0.txt",
  "bdnf_median_T1T0.txt",
  "bdnf_median_T2T1.txt",
  
  "tau_mean_T2T0.txt",
  "tau_mean_T1T0.txt",
  "tau_mean_T2T1.txt",
  
  "tau_median_T2T0.txt",
  "tau_median_T1T0.txt",
  "tau_median_T2T1.txt",
  
  "psd95_mean_T2T0.txt",
  "psd95_mean_T1T0.txt",
  "psd95_mean_T2T1.txt",
  
  "psd95_median_T2T0.txt",
  "psd95_median_T1T0.txt",
  "psd95_median_T2T1.txt"
)

# Folder where the files are located
subfolder <- "/home/kobi/Work/TBM-R-Analysis/tables/timepoint_tables"

# Loop through the file names, read each table, and assign it to a variable
for (file_name in file_names) {
  # Construct the full file path by combining the subfolder and file name
  file_path <- file.path(subfolder, file_name)
  
  # Remove the .txt extension to get the base name for the variable
  var_name <- tools::file_path_sans_ext(file_name)
  
  # Read the table and assign it to a variable with the same name
  assign(var_name, read.table(file_path, sep = "\t", header = TRUE, row.names = NULL, quote = ""))
}

# Example: access the 'bdnf_mean_T2T0' table
head(bdnf_mean_T2T0)

```

## LM: Regression analysis (Bivariate/multivariate)

### BDNF

#### T2-T0

##### mean

```{r}
lm_bdnf_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_mean_T2T0)

summary(lm_bdnf_mean_T2T0)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T2T0)
```

##### median

```{r}
lm_bdnf_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_median_T2T0)

summary(lm_bdnf_median_T2T0)

par(mfrow = c(2,2))
plot(lm_bdnf_median_T2T0)
```

#### T1 - T0

##### mean

```{r}
lm_bdnf_mean_T1T0 <- lm(T1_T0 ~ as.factor(Standort) + as.factor(Gruppe) + as.factor(Sex) + Age_start + CogTel + T0, 
                  data = bdnf_mean_T1T0)

summary(lm_bdnf_mean_T1T0)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T1T0)
```

Gruppendifferenz!

##### median

Gruppendifferenz bei mean nicht mehr vorhanden beim Median

```{r}
lm_bdnf_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = bdnf_median_T1T0)

summary(lm_bdnf_median_T1T0)

par(mfrow = c(2,2))
plot(lm_bdnf_median_T1T0)
```

#### T2 - T1

##### mean

```{r}
lm_bdnf_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = bdnf_mean_T2T1)

summary(lm_bdnf_mean_T2T1)

par(mfrow = c(2,2))
plot(lm_bdnf_mean_T2T1)
```

##### median

```{r}
lm_bdnf_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = bdnf_median_T2T1)

summary(lm_bdnf_median_T2T1)

par(mfrow = c(2,2))
plot(lm_bdnf_median_T2T1)
```

### Tau

#### T2-T0

##### mean

```{r}
lm_tau_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_mean_T2T0)

summary(lm_tau_mean_T2T0)


par(mfrow = c(2,2))
plot(lm_tau_mean_T2T0)
```

##### median

```{r}
lm_tau_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_median_T2T0)

summary(lm_tau_median_T2T0)

par(mfrow = c(2,2))
plot(lm_tau_median_T2T0)
```

#### T1 - T0

##### mean

```{r}
lm_tau_mean_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_mean_T1T0)

summary(lm_tau_mean_T1T0)

par(mfrow = c(2,2))
plot(lm_tau_mean_T1T0)
```

##### median

```{r}
lm_tau_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = tau_median_T1T0)

summary(lm_tau_median_T1T0)

par(mfrow = c(2,2))
plot(lm_tau_median_T1T0)
```

Standortdifferenz im Median?

#### T2 - T1

##### mean

```{r}
lm_tau_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = tau_mean_T2T1)

summary(lm_tau_mean_T2T1)

par(mfrow = c(2,2))
plot(lm_tau_mean_T2T1)
```

##### median

```{r}
lm_tau_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = tau_median_T2T1)

summary(lm_tau_median_T2T1)

par(mfrow = c(2,2))
plot(lm_tau_median_T2T1)
```

### PSD95

#### T2-T0

##### mean

```{r}
lm_psd95_mean_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_mean_T2T0)

summary(lm_psd95_mean_T2T0)

par(mfrow = c(2,2))
plot(lm_psd95_mean_T2T0)
```

##### median

```{r}
lm_psd95_median_T2T0 <- lm(T2_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_median_T2T0)

summary(lm_psd95_median_T2T0)

par(mfrow = c(2,2))
plot(lm_psd95_median_T2T0)
```

#### T1 - T0

##### mean

```{r}
lm_psd95_mean_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_mean_T1T0)

summary(lm_psd95_mean_T1T0)
```

Standortdifferenz im Mean? \##### median

```{r}
lm_psd95_median_T1T0 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start + CogTel + T0, 
                  data = psd95_median_T1T0)

summary(lm_psd95_median_T1T0)
```

Keine Standortdifferenz im Mean? Dafür Geschlechtsdifferenz?!

#### T2 - T1

##### mean

```{r}
lm_psd95_mean_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = psd95_mean_T2T1)

summary(lm_psd95_mean_T2T1)

par(mfrow = c(2,2))
plot(lm_psd95_mean_T2T1)

```

##### median

```{r}
lm_psd95_median_T2T1 <- lm(T2_T1 ~ Standort + Gruppe + Sex + Age_start + CogTel + T1, 
                  data = psd95_median_T2T1)

summary(lm_psd95_median_T2T1)

par(mfrow = c(2,2))
plot(lm_psd95_median_T2T1)
```


```{r}
#| echo: true
#| message: false
#| warning: false

library(report)
library(broom)
library(dplyr)
library(ggplot2)

# --------------------------
# 1) Collect models
# --------------------------

mean_models <- list(
  bdnf_T2T0  = lm_bdnf_mean_T2T0,
  bdnf_T1T0  = lm_bdnf_mean_T1T0,
  bdnf_T2T1  = lm_bdnf_mean_T2T1,

  tau_T2T0   = lm_tau_mean_T2T0,
  tau_T1T0   = lm_tau_mean_T1T0,
  tau_T2T1   = lm_tau_mean_T2T1,

  psd95_T2T0 = lm_psd95_mean_T2T0,
  psd95_T1T0 = lm_psd95_mean_T1T0,
  psd95_T2T1 = lm_psd95_mean_T2T1
)

median_models <- list(
  bdnf_T2T0  = lm_bdnf_median_T2T0,
  bdnf_T1T0  = lm_bdnf_median_T1T0,
  bdnf_T2T1  = lm_bdnf_median_T2T1,

  tau_T2T0   = lm_tau_median_T2T0,
  tau_T1T0   = lm_tau_median_T1T0,
  tau_T2T1   = lm_tau_median_T2T1,

  psd95_T2T0 = lm_psd95_median_T2T0,
  psd95_T1T0 = lm_psd95_median_T1T0,
  psd95_T2T1 = lm_psd95_median_T2T1
)

# --------------------------
# 2) Narrative reports
# --------------------------

cat("\n\n# Report: Mean models\n\n")
for (nm in names(mean_models)) {
  cat("\n\n", "## Mean model:", nm, "\n\n")
  cat(as.character(report(mean_models[[nm]])))
  cat("\n")
}

cat("\n\n# Report: Median models\n\n")
for (nm in names(median_models)) {
  cat("\n\n", "## Median model:", nm, "\n\n")
  cat(as.character(report(median_models[[nm]])))
  cat("\n")
}

# --------------------------
# 3) Helper: tidy all models into one df
# --------------------------

tidy_models <- function(model_list, stat_label = c("mean", "median")) {
  stat_label <- match.arg(stat_label)
  bind_rows(lapply(names(model_list), function(nm) {
    broom::tidy(model_list[[nm]], conf.int = TRUE) |>
      mutate(
        model = nm,
        stat  = stat_label,
        biomarker = sub("_.*", "", nm),
        contrast  = sub("^[^_]+_", "", nm)
      )
  }))
}

coef_df <- bind_rows(
  tidy_models(mean_models, "mean"),
  tidy_models(median_models, "median")
)

# --------------------------
# 4) Plot A: Group effect across contrasts
# --------------------------
# Group term can appear in different encodings depending on factor handling
group_terms <- c("GruppePiano", "as.factor(Gruppe)Piano")

group_df <- coef_df |>
  filter(term %in% group_terms)

if (nrow(group_df) > 0) {
  p_group <- ggplot(group_df, aes(x = contrast, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_pointrange() +
    facet_grid(stat ~ biomarker, scales = "free_y") +
    labs(
      title = "Group effect (Piano vs Control) across time contrasts",
      x = "Time contrast",
      y = "Coefficient estimate (95% CI)"
    )
  print(p_group)
} else {
  message("No group terms found in model coefficients (check variable coding of Gruppe).")
}

# --------------------------
# 5) Plot B: Baseline effect across contrasts (T0/T1)
# --------------------------
baseline_terms <- c("T0", "T1")

baseline_df <- coef_df |>
  filter(term %in% baseline_terms) |>
  mutate(baseline = term)

if (nrow(baseline_df) > 0) {
  p_base <- ggplot(baseline_df, aes(x = contrast, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_pointrange() +
    facet_grid(stat + baseline ~ biomarker, scales = "free_y") +
    labs(
      title = "Baseline effect (T0 / T1) across time contrasts",
      x = "Time contrast",
      y = "Coefficient estimate (95% CI)"
    )
  print(p_base)
} else {
  message("No baseline terms found (expected terms T0 or T1).")
}

```




```{r}
#| echo: true
#| message: false
#| warning: false

library(report)
library(broom)
library(dplyr)
library(ggplot2)

# --------------------------
# 1) Collect MEAN models only
# --------------------------

mean_models <- list(
  bdnf_T2T0  = lm_bdnf_mean_T2T0,
  bdnf_T1T0  = lm_bdnf_mean_T1T0,
  bdnf_T2T1  = lm_bdnf_mean_T2T1,

  tau_T2T0   = lm_tau_mean_T2T0,
  tau_T1T0   = lm_tau_mean_T1T0,
  tau_T2T1   = lm_tau_mean_T2T1,

  psd95_T2T0 = lm_psd95_mean_T2T0,
  psd95_T1T0 = lm_psd95_mean_T1T0,
  psd95_T2T1 = lm_psd95_mean_T2T1
)

# --------------------------
# 2) Narrative reports (MEAN only)
# --------------------------

cat("\n\n# Report: Mean models\n\n")
for (nm in names(mean_models)) {
  cat("\n\n", "## Mean model:", nm, "\n\n")
  cat(as.character(report(mean_models[[nm]])))
  cat("\n")
}

# --------------------------
# 3) Helper: tidy MEAN models into one df
# --------------------------

coef_df <- bind_rows(lapply(names(mean_models), function(nm) {
  broom::tidy(mean_models[[nm]], conf.int = TRUE) |>
    mutate(
      model = nm,
      biomarker = sub("_.*", "", nm),
      contrast  = sub("^[^_]+_", "", nm)
    )
}))
```


```{r}
#| label: fig-lm-group-effect-filtered
#| fig-cap: "Group effect (Piano vs Control) across time contrasts (mean models). Points are coefficients; bars are 95% CIs."
# --------------------------
# 4) Plot A: Group effect across contrasts (MEAN only)
# --------------------------

# Group term can appear in different encodings depending on factor handling
group_terms <- c("GruppePiano", "as.factor(Gruppe)Piano")

group_df <- coef_df |>
  filter(term %in% group_terms)

if (nrow(group_df) > 0) {
  p_group <- ggplot(group_df, aes(x = contrast, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_pointrange() +
    facet_wrap(~ biomarker, scales = "free_y") +
    labs(
      title = "Group effect (Piano vs Control) across time contrasts (Mean models)",
      x = "Time contrast",
      y = "Coefficient estimate (95% CI)"
    )
  print(p_group)
} else {
  message("No group terms found in model coefficients (check variable coding of Gruppe).")
}
```


```{r}
# --------------------------
# 5) Plot B: Baseline effect across contrasts (T0/T1) (MEAN only)
# --------------------------

baseline_terms <- c("T0", "T1")

baseline_df <- coef_df |>
  filter(term %in% baseline_terms) |>
  mutate(baseline = term)

if (nrow(baseline_df) > 0) {
  p_base <- ggplot(baseline_df, aes(x = contrast, y = estimate, ymin = conf.low, ymax = conf.high)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_pointrange() +
    facet_grid(baseline ~ biomarker, scales = "free_y") +
    labs(
      title = "Baseline effect (T0 / T1) across time contrasts (Mean models)",
      x = "Time contrast",
      y = "Coefficient estimate (95% CI)"
    )
  print(p_base)
} else {
  message("No baseline terms found (expected terms T0 or T1).")
}

```



```{r}

modelsummary(
  meanmodels,
  statistic = c("({std.error})", "p={p.value}"),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  output = "markdown"
)
```


```{r}
#| echo: true
#| message: false
#| warning: false

library(modelsummary)

# 1) Rename models for readable column headers (optional but nice)
names(mean_models) <- c(
  "BDNF T2–T0", "BDNF T1–T0", "BDNF T2–T1",
  "Tau T2–T0",  "Tau T1–T0",  "Tau T2–T1",
  "PSD95 T2–T0","PSD95 T1–T0","PSD95 T2–T1"
)

# 2) Keep only Group and baseline timepoint terms
# (handles both "GruppePiano" and "as.factor(Gruppe)Piano")
coef_keep <- c(
  "GruppePiano"             = "Group: Piano vs Control",
  "as.factor(Gruppe)Piano"  = "Group: Piano vs Control",
  "T0"                      = "Baseline (T0 or T1)",
  "T1"                      = "Baseline (T0 or T1)"
)

modelsummary(
  mean_models,
  coef_map = coef_keep,
  estimate = "{estimate} ({std.error})",
  statistic = "p.value",
  fmt = 3,
  stars = TRUE,
  gof_omit = "IC|Log|Adj|RMSE|AIC|BIC|Deviance|Std.Errors"
)

```





## Summary of results
```{r}
library(report)
report(lm_bdnf_mean_T1T0)
#report(lm_bdnf_median_T1T0)
report(lm_tau_mean_T1T0)
#report(lm_tau_median_T1T0)
report(lm_psd95_mean_T1T0)
#report(lm_psd95_median_T1T0)
summary(CogTel)
```

```{r}
report(lm_bdnf_mean_T2T0)
#report(lm_bdnf_median_T2T0)
report(lm_tau_mean_T2T0)
#report(lm_tau_median_T2T0)
report(lm_psd95_mean_T2T0)
#report(lm_psd95_median_T2T0)
```

```{r}
library(dplyr)
library(report)
library(parameters)
library(modelsummary)

# --- helper: center continuous covariates for interpretable intercepts & stability ---
prep_change <- function(df) {
  df %>%
    mutate(
      Standort = as.factor(Standort),
      Gruppe   = as.factor(Gruppe),
      Sex      = as.factor(Sex),
      Age_start = as.numeric(Age_start),
      CogTel    = as.numeric(CogTel),
      T0        = as.numeric(T0),
      T1_T0     = as.numeric(T1_T0),
      Age_start_c = scale(Age_start, scale = FALSE),
      CogTel_c    = scale(CogTel, scale = FALSE),
      T0_c        = scale(T0, scale = FALSE)
    )
}

bdnf_dat  <- prep_change(bdnf_mean_T1T0)
tau_dat   <- prep_change(tau_mean_T1T0)
psd95_dat <- prep_change(psd95_mean_T1T0)

lm_bdnf  <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start_c + CogTel_c + T0_c, data = bdnf_dat)
lm_tau   <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start_c + CogTel_c + T0_c, data = tau_dat)
lm_psd95 <- lm(T1_T0 ~ Standort + Gruppe + Sex + Age_start_c + CogTel_c + T0_c, data = psd95_dat)

# --- 1) REPORT text blocks (like you showed, but now with centered covariates)
cat(report(lm_bdnf), "\n\n")
cat(report(lm_tau), "\n\n")
cat(report(lm_psd95), "\n\n")

# --- 2) One combined regression table (publication friendly)
models <- list(
  "BDNF (T1-T0)"  = lm_bdnf,
  "TAU (T1-T0)"   = lm_tau,
  "PSD95 (T1-T0)" = lm_psd95
)

modelsummary(
  models,
  statistic = c("({std.error})", "p={p.value}"),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  output = "markdown"
)

```


