---
title: "Plots: Time + Metyhlation"
author: "Jonas J.R. Koberschinski"
date: "09-12-2024"
last_modified: "2026-01-29"
output: html
---

## Data import

```{r}
#| echo: false      # Hide the code
#| include: false   # Exclude the output
#| warning: false   # Suppress warnings
#| message: false   # Suppress messages#

## Set working directory

# setwd("~/Work/Promotionsdatananalyse mit Excel und Martin/Rohdaten/")
# print(getwd())

## Load necessary packages
library(haven)
library(tidyverse)
library(dplyr)
library(readxl)
library(writexl)


BDNF <- read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/BDNF_Neuroplastizität_Alter_CpGs_ohne_m11_und_p36_mit_verhalten_vollständig_umstrukturiert.xlsx")

TAU <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/TAU_CpGs_mit_verhalten_ohne_p61_p59_p34_p29_umstruk.xlsx"))

PSD95 <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/PSD95_CpGs_mit_verhalten_größer_umstrukturiert.xlsx"))
```

## Drop NA
```{r}
#| echo: true
#| include: false

# BDNF
# Remove rows with NA values in the 'methylierung' column
BDNF2 <- BDNF %>% drop_na(methylierung)

# Summarize the number of NA values in the 'methylierung' column (commented out)
# sum(is.na(BDNF$methylierung))

# without NA
# summary(BDNF2)



# TAu

# Remove rows with NA values in the 'methylierung' column
TAU2 <- TAU %>% drop_na(methylation)

# Summarize the number of NA values in the 'methylierung' column (commented out)
# sum(is.na(TAU$methylierung))

# without NA
# summary(TAU2)


# PSD95
# Remove rows with NA values in the 'methylierung' column
PSD952 <- PSD95 %>% drop_na(methylation)

# Summarize the number of NA values in the 'methylierung' column (commented out)
# sum(is.na(BDNF$methylierung))

# without NA
# summary(PSD952)
```
## Remove problematic IDs
```{r}
# Remove problematic IDs (apply consistently)
bad_ids <- c(192, 193)

BDNF2  <- BDNF2  %>% filter(!(ID %in% bad_ids))
TAU2   <- TAU2   %>% filter(!(ID %in% bad_ids))
PSD952 <- PSD952 %>% filter(!(ID %in% bad_ids))

```

## Descriptive statistics table
```{r}

# Calculate mean and SD for each gene
tab <- data.frame(
  Marker = c("BDNF", "Tau", "PSD95"),
  Mean = c(
    mean(BDNF2$methylierung, na.rm = TRUE),
    mean(TAU2$methylation, na.rm = TRUE),
    mean(PSD952$methylation, na.rm = TRUE)
  ),
  SD = c(
    sd(BDNF2$methylierung, na.rm = TRUE),
    sd(TAU2$methylation, na.rm = TRUE),
    sd(PSD952$methylation, na.rm = TRUE)
  )
)

tab$`Mean ± SD` <- sprintf("%.3f ± %.3f", tab$Mean, tab$SD)
tab <- tab[, c("Marker", "Mean ± SD")]

# Print as markdown
cat("| Marker | Mean ± SD |\n")
cat("|--------|-----------|\n")
apply(tab, 1, function(r) cat(sprintf("| %s | %s |\n", r[1], r[2])))
```


## Interaction Plots

ChatGPT help:

The core issue in one sentence

A CpG is not a participant.
If you treat CpGs like independent observations, you inflate N and distort SD/SE.

Everything else follows from that.

What your data actually look like
1️⃣ CpG-level data (long format: BDNF2, TAU2, …)

Each row is:

one CpG site × one person × one time point

Example (simplified):

ID	Time	CpG	methylation
1	T0	m148	0.18
1	T0	m111	0.23
1	T0	m087	0.00
…	…	…	…

So for one person at one time, you might have 20–30 CpGs.

When you do:

mean(BDNF2$methylierung)
sd(BDNF2$methylierung)


you are computing:

“What is the mean/SD of all CpG measurements pooled together?”

This mixes three different sources of variation:

Differences between CpG sites (huge!)

Differences between people

Changes over time + noise

That’s why:

SD is large (~0.09–0.10)

SE becomes artificially tiny if you divide by sqrt(nrow(BDNF2))

Plots look “too confident”

This is not wrong mathematically, but it answers the wrong scientific question.

What your paper actually claims to measure

Your tables, text, and mixed models all say things like:

“BDNF methylation level per gene”

That implies:

one methylation value per person per time point

So you must collapse CpGs first.

2️⃣ Sample-level data (wide format: bdnf_mean_wide, tau_mean_wide)

Here each row is:

one person, with CpGs already averaged

Example:

ID	Group	T0	T1	T2
1	Control	0.075	0.037	NA
2	Piano	0.102	0.060	0.084

Each value (e.g. 0.075) is:

mean methylation across all CpGs for that person at that time

Now when you compute:

mean(bdnf_mean_wide$T0, na.rm=TRUE)
sd(bdnf_mean_wide$T0, na.rm=TRUE)


you are asking:

“How much do people differ in their average BDNF methylation at T0?”

This is exactly what readers expect.

That’s why:

SD ≈ 0.03–0.04

SE ≈ SD / √N_participants

Plots look sensible

Results match your LMM interpretation

Why your interaction plots broke

You originally did this (conceptually):

group_by(Time, Group) %>%
summarise(
  mean = mean(methylation),
  se = sd(methylation) / sqrt(n())
)


But here:

methylation = CpG-level values

n() = number of CpGs × people × time

So SE was:

SD across CpGs ÷ √(CpGs × people)

That is not the SE of the group mean across people.

That’s why BDNF “kind of looked OK” (by coincidence) and TAU didn’t.

The correct mental model (memorize this)
Question	Unit of analysis	Data to use
“What is average gene methylation per person?”	Person	CpGs averaged first
“Do groups differ in gene methylation?”	Person	Sample-level mean
“Does methylation change over time?”	Person	Sample-level mean
“Do CpGs behave differently?”	CpG	CpG-level model

Never mix these levels.

Why your mixed model was actually OK

Your LMM used CpG-level rows, but:

you used random intercepts for person

you interpreted fixed effects at the gene level

That’s defensible if and only if:

you don’t treat CpGs as independent in descriptives

you explain that CpGs are repeated measures within person

Your tables/plots must still reflect person-level summaries.

One sentence you can put in Methods (this resolves everything)

“For descriptive statistics and visualization, methylation values were averaged across all available CpG sites per gene for each participant and time point.”

That sentence legally aligns:

your tables

your plots

your interpretation

your reviewer sanity

Final rule of thumb (print this mentally)

CpGs are repeated measures, not observations.
People are observations.

If you want, next we can:

re-express the LMM more cleanly (random slope for time, CpG as random effect)

decide whether to keep CpG-level modeling or simplify to sample-level LMM

Just tell me which way you want to go.

do this in markdown so i can add it to script


### BDNF

```{r}
# Interaction plot showing methylation levels by group and time
interaction_plot_bdnf <- ggplot(BDNF2, aes(x = Zeitpunkt, y = methylierung, group = Group, colour = Group)) +
  geom_line(stat = "summary", fun = mean) +
  geom_point(stat = "summary", fun = mean) +
  theme_classic() +
  labs(
    title = "Interaction Plot of BDNF Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_bdnf)
```

```{r}
#| label: fig-bdnf-timegroup-interaction-wrong
#| fig-cap: "BDNF methylation levels (mean ± SE) by group across time points."
library(dplyr)
library(ggplot2)

# Calculate summary statistics for BDNF
summary_stats_bdnf <- BDNF2 %>%
  group_by(Zeitpunkt, Group) %>%
  summarise(
    mean_methylierung_bdnf = mean(methylierung, na.rm = TRUE),
    se_methylierung_bdnf = sd(methylierung, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Interaction plot with error bars for BDNF
interaction_plot_bdnf <- ggplot(summary_stats_bdnf, aes(x = Zeitpunkt, y = mean_methylierung_bdnf, group = Group, colour = Group)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = mean_methylierung_bdnf - se_methylierung_bdnf,
      ymax = mean_methylierung_bdnf + se_methylierung_bdnf
    ),
    width = 0.2
  ) +
  theme_classic() +
  labs(
    title = "Interaction Plot of BDNF Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_bdnf)
```

```{r}
#| label: fig-bdnf-timegroup-interaction-cpg-mean-wrong
#| fig-cap: "BDNF methylation levels (mean ± SE) by group across time points, averaged across CpGs."
library(dplyr)
library(ggplot2)

# 1) per-participant per-timepoint mean across CpGs
bdnf_samplemeans <- BDNF2 %>%
  group_by(ID, Zeitpunkt, Group) %>%
  summarise(
    sample_mean = mean(methylierung, na.rm = TRUE),
    .groups = "drop"
  )

# 2) group/time summary across participants
df_plot_bdnf <- bdnf_samplemeans %>%
  group_by(Zeitpunkt, Group) %>%
  summarise(
    n_participants = n_distinct(ID),
    mean = mean(sample_mean, na.rm = TRUE),
    sd = sd(sample_mean, na.rm = TRUE),
    se = sd / sqrt(n_participants),
    .groups = "drop"
  ) %>%
  arrange(Group, Zeitpunkt)

# look at the data used for plotting
print(df_plot_bdnf)
# or if you want a nicer view in RStudio:
# View(df_plot_bdnf)

# plot
ggplot(df_plot_bdnf, aes(x = Zeitpunkt, y = mean, group = Group, colour = Group)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(
    title = "Interaction Plot of BDNF Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean methylation (across CpGs)",
    caption = "Points show mean across participants; error bars are ±SE."
  )

```

```{r}
#interaction_plot_smoking_bdnf <- ggplot(summary_stats_bdnf, aes(x = Zeitpunkt, y = mean_methylierung_bdnf, group = rauchen, colour = rauchen)) +
#  geom_line() +
#  geom_point() +
#  geom_errorbar(
#    aes(
#      ymin = mean_methylierung_bdnf - se_methylierung_bdnf,
#      ymax = mean_methylierung_bdnf + se_methylierung_bdnf
#    ),
#    width = 0.2
#  ) +
#  theme_classic() +
#  labs(
#    title = "Interaction Plot of BDNF Methylation Levels by Group and Time",
#    x = "Time Point",
#    y = "Mean Methylation Levels"
#  )
#print(interaction_plot_smoking_bdnf)

```

```{r}
boxplot(methylierung ~ Group, data = BDNF2,
        main = "Boxplot of BDNF Methylation Levels by Group and Time",
        xlab = "Group and Time Point",
        ylab = "Methylation Levels",
        col = c("lightblue", "lightgreen"))
```
```{r}
BDNF2$time_group <- interaction(BDNF2$Zeitpunkt, BDNF2$Group, sep = " - ", drop = TRUE)

boxplot(methylierung ~ time_group, data = BDNF2,
        main = "BDNF methylation by time and group",
        xlab = "Time point - Group",
        ylab = "Methylation level",
        outline = FALSE,
        las = 2,
        col = rep(c("lightblue","lightgreen"), times = nlevels(BDNF2$Zeitpunkt)))


```

### Tau

```{r}
# Interaction plot showing methylation levels by group and time
interaction_plot_tau <- ggplot(TAU2, aes(x = Zeitpunkt, y = methylation, group = Gruppe, colour = Gruppe)) +
  geom_line(stat = "summary", fun = mean) +
  geom_point(stat = "summary", fun = mean) +
  theme_classic() +
  labs(
    title = "Interaction Plot of Tau Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_tau)
```

```{r}
#| label: fig-tau-timegroup-interaction-wrong
#| fig-cap: "Tau methylation levels (mean ± SE) by group across time points."

library(dplyr)
library(ggplot2)

# Calculate summary statistics
summary_stats_tau <- TAU2 %>%
  group_by(Zeitpunkt, Gruppe) %>%
  summarise(
    mean_methylation_tau = mean(methylation, na.rm = TRUE),
    se_methylation_tau = sd(methylation, na.rm = TRUE) / sqrt(n()),
  )

# Interaction plot with error bars
interaction_plot_tau_error <- ggplot(summary_stats_tau, aes(x = as.factor(Zeitpunkt), y = mean_methylation_tau, group = Gruppe, colour = Gruppe)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = mean_methylation_tau - se_methylation_tau,
      ymax = mean_methylation_tau + se_methylation_tau
    ),
    width = 0.2
  ) +
  theme_classic() +
  labs(
    title = "Interaction Plot of Tau Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_tau_error)
```

```{r}
#| label: fig-tau-timegroup-interaction-cpg-mean-wrong
#| fig-cap: "Tau methylation levels (mean ± SE) by group across time points."

library(dplyr)
library(ggplot2)

# 1) per-participant per-timepoint mean across CpGs
tau_samplemeans <- TAU2 %>%
  group_by(ID, Zeit, Gruppe) %>%
  summarise(
    sample_mean = mean(methylation, na.rm = TRUE),
    .groups = "drop"
  )

# 2) group/time summary across participants
summary_stats_tau <- tau_samplemeans %>%
  group_by(Zeit, Gruppe) %>%
  summarise(
    n_participants = n(),
    mean_methylation_tau = mean(sample_mean, na.rm = TRUE),
    se_methylation_tau = sd(sample_mean, na.rm = TRUE) / sqrt(n_participants),
    .groups = "drop"
  )

# 3) plot
ggplot(summary_stats_tau,
       aes(x = as.factor(Zeit), y = mean_methylation_tau, group = Gruppe, colour = Gruppe)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = mean_methylation_tau - se_methylation_tau,
      ymax = mean_methylation_tau + se_methylation_tau
    ),
    width = 0.2
  ) +
  theme_classic() +
  labs(
    title = "Interaction Plot of Tau Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean methylation (across CpGs)",
    caption = "Points show mean across participants; error bars are ±SE."
  )

```


### PSD95

```{r}
# Interaction plot showing methylation levels by group and time
interaction_plot_psd95 <- ggplot(PSD952, aes(x = Zeitpunkt, y = methylation, group = Gruppe, colour = Gruppe)) +
  geom_line(stat = "summary", fun = mean) +
  geom_point(stat = "summary", fun = mean) +
  theme_classic() +
  labs(
    title = "Interaction Plot of PSD95 Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_psd95)
```

```{r}
#| label: fig-psd95-timegroup-interaction-wrong
#| fig-cap: "PSD95 methylation levels (mean ± SE) by group across time points."
library(dplyr)
library(ggplot2)

# Calculate summary statistics for PSD95
summary_stats_psd95 <- PSD952 %>%
  group_by(Zeitpunkt, Gruppe) %>%
  summarise(
    mean_methylation_psd95 = mean(methylation, na.rm = TRUE),
    se_methylation_psd95 = sd(methylation, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Interaction plot with error bars for PSD95
interaction_plot_psd95 <- ggplot(summary_stats_psd95, aes(x = Zeitpunkt, y = mean_methylation_psd95, group = Gruppe, colour = Gruppe)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = mean_methylation_psd95 - se_methylation_psd95,
      ymax = mean_methylation_psd95 + se_methylation_psd95
    ),
    width = 0.2
  ) +
  theme_classic() +
  labs(
    title = "Interaction Plot of PSD95 Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean Methylation Levels"
  )

print(interaction_plot_psd95)
```
```{r}
psd95_samplemeans <- PSD952 %>%
  group_by(ID, Zeitpunkt, Gruppe) %>%
  summarise(
    sample_mean = mean(methylation, na.rm = TRUE),
    .groups = "drop"
  )
summary_stats_psd95 <- psd95_samplemeans %>%
  group_by(Zeitpunkt, Gruppe) %>%
  summarise(
    n_participants = n(),
    mean_methylation_psd95 = mean(sample_mean, na.rm = TRUE),
    se_methylation_psd95 = sd(sample_mean, na.rm = TRUE) / sqrt(n_participants),
    .groups = "drop"
  )



ggplot(summary_stats_psd95,
       aes(x = Zeitpunkt,
           y = mean_methylation_psd95,
           group = Gruppe,
           colour = Gruppe)) +
  geom_line() +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = mean_methylation_psd95 - se_methylation_psd95,
      ymax = mean_methylation_psd95 + se_methylation_psd95
    ),
    width = 0.2
  ) +
  theme_classic() +
  labs(
    title = "Interaction Plot of PSD95 Methylation Levels by Group and Time",
    x = "Time Point",
    y = "Mean methylation (averaged across CpGs)",
    caption = "Points show mean across participants; error bars represent ±SE."
  )

```


## Plots of all genes for publication

### group differences

```{r}
#| label: fig-interactions-by-gene-cpg-mean-wrong
#| fig-cap: "Methylation levels (mean ± SE) across time points by group for BDNF, Tau, and PSD95."
#| fig-width: 8
#| fig-height: 10

library(dplyr)
library(ggplot2)
library(patchwork)

# ---- BDNF ----
sum_bdnf <- BDNF2 %>%
  group_by(Zeitpunkt, Group) %>%
  summarise(
    mean = mean(methylierung, na.rm = TRUE),
    se   = sd(methylierung, na.rm = TRUE) / sqrt(sum(!is.na(methylierung))),
    .groups = "drop"
  )

p_bdnf <- ggplot(sum_bdnf, aes(Zeitpunkt, mean, group = Group, colour = Group)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "BDNF", x = "Time Point", y = "Mean methylation (±SE)")

# ---- Tau ----
sum_tau <- TAU2 %>%
  group_by(Zeit, Gruppe) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_tau <- ggplot(sum_tau, aes(Zeit, mean, group = Gruppe, colour = Gruppe)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "Tau", x = "Time Point", y = "Mean methylation (±SE)")

# ---- PSD95 ----
sum_psd95 <- PSD952 %>%
  group_by(Zeitpunkt, Gruppe) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_psd95 <- ggplot(sum_psd95, aes(Zeitpunkt, mean, group = Gruppe, colour = Gruppe)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "PSD95", x = "Time Point", y = "Mean methylation (±SE)")

(p_bdnf / p_tau / p_psd95) + plot_layout(guides = "collect") &
  theme(legend.position = "right")

```

### Over Time

```{r}
#| label: fig-interactions-no-group
#| fig-cap: "Methylation levels (mean ± SE) across time points collapsed across groups for BDNF, Tau, and PSD95."
#| fig-width: 8
#| fig-height: 10

library(dplyr)
library(ggplot2)
library(patchwork)

# BDNF (no group)
sum_bdnf0 <- BDNF2 %>%
  group_by(Zeitpunkt) %>%
  summarise(
    mean = mean(methylierung, na.rm = TRUE),
    se   = sd(methylierung, na.rm = TRUE) / sqrt(sum(!is.na(methylierung))),
    .groups = "drop"
  )

p_bdnf0 <- ggplot(sum_bdnf0, aes(Zeitpunkt, mean, group = 1)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "BDNF", x = "Time Point", y = "Mean methylation (±SE)")

# Tau (no group)
sum_tau0 <- TAU2 %>%
  group_by(Zeit) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_tau0 <- ggplot(sum_tau0, aes(Zeit, mean, group = 1)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "Tau", x = "Time Point", y = "Mean methylation (±SE)")

# PSD95 (no group)
sum_psd950 <- PSD952 %>%
  group_by(Zeitpunkt) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_psd950 <- ggplot(sum_psd950, aes(Zeitpunkt, mean, group = 1)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "PSD95", x = "Time Point", y = "Mean methylation (±SE)")

(p_bdnf0 / p_tau0 / p_psd950)
```



## Filtered after PCA

### Before elbow

```{r}

keep_psd95 <- c("p95", "p60", "p23")
keep_tau   <- c("m331", "m337", "m310", "m315", "m323")
keep_bdnf  <- c("m148", "m066", "m111", "m058")

PSD952_f <- PSD952 %>% filter(CpG_Position %in% keep_psd95)
TAU2_f   <- TAU2   %>% filter(CpG_position %in% keep_tau)
BDNF2_f  <- BDNF2  %>% filter(CpG_position %in% keep_bdnf)
```

```{r}
#| label: fig-interactions-by-filtered-gene
#| fig-cap: "Methylation levels (mean ± SE) across time points by group for selected high-variance CpGs (BDNF: m148, m066, m111, m058; Tau: m331, m337, m310, m315, m323; PSD95: p95, p60, p23)."
#| fig-width: 8
#| fig-height: 10

library(dplyr)
library(ggplot2)
library(patchwork)

# ---- BDNF ----
sum_bdnf <- BDNF2_f %>%
  group_by(Zeitpunkt, Group) %>%
  summarise(
    mean = mean(methylierung, na.rm = TRUE),
    se   = sd(methylierung, na.rm = TRUE) / sqrt(sum(!is.na(methylierung))),
    .groups = "drop"
  )

p_bdnf <- ggplot(sum_bdnf, aes(Zeitpunkt, mean, group = Group, colour = Group)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "BDNF (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")

# ---- Tau ----
sum_tau <- TAU2_f %>%
  group_by(Zeit, Gruppe) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_tau <- ggplot(sum_tau, aes(Zeit, mean, group = Gruppe, colour = Gruppe)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "Tau (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")

# ---- PSD95 ----
sum_psd95 <- PSD952_f %>%
  group_by(Zeitpunkt, Gruppe) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_psd95 <- ggplot(sum_psd95, aes(Zeitpunkt, mean, group = Gruppe, colour = Gruppe)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "PSD95 (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")

(p_bdnf / p_tau / p_psd95) + plot_layout(guides = "collect") &
  theme(legend.position = "right")

```




```{r}
#| label: fig-interactions-no-group-filtered-genes
#| fig-cap: "Methylation levels (mean ± SE) across time points collapsed across groups for selected high-variance CpGs."
#| fig-width: 8
#| fig-height: 10

library(dplyr)
library(ggplot2)
library(patchwork)

# BDNF (no group)
sum_bdnf0 <- BDNF2_f %>%
  group_by(Zeitpunkt) %>%
  summarise(
    mean = mean(methylierung, na.rm = TRUE),
    se   = sd(methylierung, na.rm = TRUE) / sqrt(sum(!is.na(methylierung))),
    .groups = "drop"
  )

p_bdnf0 <- ggplot(sum_bdnf0, aes(Zeitpunkt, mean, group = 1)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "BDNF (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")

# Tau (no group)
sum_tau0 <- TAU2_f %>%
  group_by(Zeit) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_tau0 <- ggplot(sum_tau0, aes(Zeit, mean, group = 1)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "Tau (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")

# PSD95 (no group)
sum_psd950 <- PSD952_f %>%
  group_by(Zeitpunkt) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_psd950 <- ggplot(sum_psd950, aes(Zeitpunkt, mean, group = 1)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "PSD95 (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")

(p_bdnf0 / p_tau0 / p_psd950)
```


### After elbow

```{r}

keep_tau2   <- c("m331", "m337", "m310", "m315", "m323", "m306")
keep_bdnf2 <- c("m148", "m066", "m111", "m058", "m039")

TAU2_f   <- TAU2   %>% filter(CpG_position %in% keep_tau2)
BDNF2_f  <- BDNF2  %>% filter(CpG_position %in% keep_bdnf2)
```

```{r}
#| label: fig-interactions-more-filtered-genes
#| fig-cap: "Methylation levels (mean ± SE) across time points by group for selected high-variance CpGs (BDNF: m148, m066, m111, m058; Tau: m331, m337, m310, m315, m323; PSD95: p95, p60, p23)."
#| fig-width: 8
#| fig-height: 9 

library(dplyr)
library(ggplot2)
library(patchwork)

# ---- BDNF ----
sum_bdnf <- BDNF2_f %>%
  group_by(Zeitpunkt, Group) %>%
  summarise(
    mean = mean(methylierung, na.rm = TRUE),
    se   = sd(methylierung, na.rm = TRUE) / sqrt(sum(!is.na(methylierung))),
    .groups = "drop"
  )

p_bdnf <- ggplot(sum_bdnf, aes(Zeitpunkt, mean, group = Group, colour = Group)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "BDNF (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")

# ---- Tau ----
sum_tau <- TAU2_f %>%
  group_by(Zeit, Gruppe) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_tau <- ggplot(sum_tau, aes(Zeit, mean, group = Gruppe, colour = Gruppe)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "Tau (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")


(p_bdnf / p_tau)
```


```{r}
#| label: fig-interactions-no-group-more-filtered-genes
#| fig-cap: "Methylation levels (mean ± SE) across time points collapsed across groups for selected high-variance CpGs."
#| fig-width: 8
#| fig-height: 10

library(dplyr)
library(ggplot2)
library(patchwork)

# BDNF (no group)
sum_bdnf0 <- BDNF2_f %>%
  group_by(Zeitpunkt) %>%
  summarise(
    mean = mean(methylierung, na.rm = TRUE),
    se   = sd(methylierung, na.rm = TRUE) / sqrt(sum(!is.na(methylierung))),
    .groups = "drop"
  )

p_bdnf0 <- ggplot(sum_bdnf0, aes(Zeitpunkt, mean, group = 1)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "BDNF (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")

# Tau (no group)
sum_tau0 <- TAU2_f %>%
  group_by(Zeit) %>%
  summarise(
    mean = mean(methylation, na.rm = TRUE),
    se   = sd(methylation, na.rm = TRUE) / sqrt(sum(!is.na(methylation))),
    .groups = "drop"
  )

p_tau0 <- ggplot(sum_tau0, aes(Zeit, mean, group = 1)) +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  theme_classic() +
  labs(title = "Tau (filtered CpGs)", x = "Time Point", y = "Mean methylation (±SE)")

(p_bdnf0 / p_tau0)
```
