---
title: "Zero Inflated Beta Regression for PSD95 Methylation Data"
author: "Jonas J. R. Koberschinski"
date: "25-10-2025"
lastmod: "25-10-2025"
output: html
---

# Preparation
## Load libraries

```{r}
library(tidyverse) # For data manipulation and visualization
library(dplyr) # For data manipulation
library(readxl) # For reading Excel files
library(glmmTMB) # For generalized linear mixed models
library(performance) # For model performance assessment
library(DHARMa) # For residual diagnostics
library(ggplot2) # For data visualization
```

## Data import

```{r}
psd95 <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/PSD95_CpGs_mit_verhalten_größer_umstrukturiert.xlsx"))

head(psd95)
```

## Subset Data and Drop NA

```{r}
# Create a subset of the data with relevant columns and drop rows with NA values
psd95_subset <- psd95 %>%
  select(
    ID,
    standort_aus_andere_quelle,
    Zeitpunkt,
    Gruppe,
    Sex,
    Age_start,
    CogTel,
    CpG_Position,
    Psycholgical,
    rauchen,
    methylation
  ) %>%
  mutate(
    ID = as.factor(ID),
    standort_aus_andere_quelle = as.factor(standort_aus_andere_quelle),
    Zeitpunkt = as.factor(Zeitpunkt),
    Gruppe = as.factor(Gruppe),
    Sex = as.factor(Sex),
    CpG_Position = as.factor(CpG_Position),
    CogTel = as.numeric(CogTel),
    rauchen = as.factor(rauchen)
  ) %>%
  drop_na()

str(psd95_subset)
```

```{r}
summary(psd95_subset$methylation)
range(psd95_subset$methylation)
sum(psd95_subset$methylation == 1)  # Count how many exact 1s
sum(psd95_subset$methylation == 0)  # For reference on zeros, ]
```

Ein einzelner Wert = 1

```{r}
#N <- nrow(psd95_subset)  # Sample size
#psd95_subset$methylation_trans <- (psd95_subset$methylation * (N - 1) + 0.5) / N
```

```{r}
psd95_subset$methylation <- ifelse(
  psd95_subset$methylation == 1,
  0.999999,
  psd95_subset$methylation
)
```

```{r}
# Center/scale continuous predictors for stability
psd95_subset$Age_start_sc <- scale(psd95_subset$Age_start)
psd95_subset$CogTel_sc    <- scale(psd95_subset$CogTel)

psd95_glmm_CpG_ID <- glmmTMB(
  methylation ~ standort_aus_andere_quelle + Gruppe * Zeitpunkt + 
                  Sex + rauchen + Age_start_sc + CogTel_sc +
                  (1 | CpG_Position) + (1 | ID) + (0 + Zeitpunkt || ID),
  ziformula = ~ Zeitpunkt + Gruppe,
  dispformula = ~ 1 + (1 | CpG_Position),   # optional: CpG-specific overdispersion
  family = beta_family(link = "logit"),
  data = psd95_subset
)
summary(psd95_glmm_CpG_ID)
```


```{r}
library(DHARMa)
sim_psd95 <- simulateResiduals(psd95_glmm_CpG_ID, n = 2000)
plot(sim_psd95)
testUniformity(sim_psd95)
testDispersion(sim_psd95)
testZeroInflation(sim_psd95)

```




```{r}
psd95_glmm_CpG_ID_phi <- update(psd95_glmm_CpG_ID,
  dispformula = ~ Zeitpunkt + Gruppe + (1 | CpG_Position)
)

```


```{r}
sim_phi <- simulateResiduals(psd95_glmm_CpG_ID_phi, n=1500)
testDispersion(sim_phi)
```

```{r}
psd95_subset$obs <- seq_len(nrow(psd95_subset))

psd95_glmm_CpG_ID_olre <- update(
  psd95_glmm_CpG_ID,
  formula = . ~ . + (1 | obs)
)

library(DHARMa)
sim_olre <- simulateResiduals(psd95_glmm_CpG_ID_olre, n = 1500)
testDispersion(sim_olre)

```



```{r}
psd95_glmm_CpG_ID_olre <- glmmTMB(
  methylation ~ standort_aus_andere_quelle + Gruppe * Zeitpunkt +
    Sex + rauchen + Age_start_sc + CogTel_sc +
    (1 | CpG_Position) + (1 | ID) + (0 + Zeitpunkt || ID) + (1 | obs),
  ziformula = ~ Zeitpunkt + Gruppe,
  dispformula = ~ 1 + (1 | CpG_Position),
  family = beta_family(link = "logit"),
  data = psd95_subset,
  control = glmmTMBControl(
    optimizer = optim,
    optArgs = list(method = "BFGS"),
    parallel = 1
  )
)
summary(psd95_glmm_CpG_ID_olre)
```

```{r}
psd95_glmm_CpG_ID_olre <- update(
  psd95_glmm_CpG_ID,
  formula = . ~ . + (1 | obs),
  start = list(theta = c(log(0.05)))  # tiny variance
)
```
```{r}
psd95_glmm_CpG_ID_olre <- update(
  psd95_glmm_CpG_ID,
  formula = . ~ . + (1 | obs),
  control = glmmTMBControl(optimizer = optim, optArgs = list(method="nlminb"))
)

```

```{r}
psd95_glmm_CpG_OLREtest <- glmmTMB(
  methylation ~ standort_aus_andere_quelle + Gruppe * Zeitpunkt +
    Sex + rauchen + Age_start_sc + CogTel_sc +
    (1 | CpG_Position) + (1 | obs) + (1 | ID) + (0 + Zeitpunkt || ID),
  ziformula = ~ Zeitpunkt + Gruppe,
  family = beta_family(link="logit"),
  data = psd95_subset
)

summary(psd95_glmm_CpG_OLREtest)
```
```{r}
library(DHARMa)
sim_olre <- simulateResiduals(psd95_glmm_CpG_OLREtest, n = 1500)
testDispersion(sim_olre)

```


Frustrane Modelle wegen bleibender Übderdispersion

Versuche auf CpGs zu verzichten gemäß PCA.
Vielleicht für Tau auch notwendig.

Nach Chat:
✅ It’s justifiable if you do it transparently and carefully

You can absolutely restrict your inferential modeling to those 3 CpGs provided that:

You pre-specify or justify the reduction.
e.g., “PCA indicated that 3 CpG positions explained >95 % of variance in PSD95 methylation; subsequent mixed modeling focused on these sites.”

You verify that these 3 CpGs are not statistical outliers or sequencing artefacts.
(Check their read quality and call rate.)

You describe them biologically.
e.g., “CpG7–CpG9 near the transcription start site were the most variable.”

You note that other CpGs were excluded because they carried redundant or negligible information.