---
title: "Zero Inflated Beta Regression for Tau Methylation Data"
author: "Jonas J. R. Koberschinski"
date: "25-10-2025"
lastmod: "25-10-2025"
output: html
---

# Preparation
## Load Required Libraries 
```{r}
library(tidyverse) # For data manipulation and visualization
library(dplyr) # For data manipulation
library(readxl) # For reading Excel files
library(glmmTMB) # For generalized linear mixed models
library(performance) # For model performance assessment
library(DHARMa)
# optional: library(emmeans)# For residual diagnostics
library(ggplot2) # For data visualization
```

## Load Data
```{r}
tau <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/TAU_CpGs_mit_verhalten_ohne_p61_p59_p34_p29_umstruk.xlsx"))
 
head(tau)
#summary(tau)
#str(tau)
```

## Subset Data and Drop NA

```{r}
# Create a subset of the data with relevant columns and drop rows with NA values
tau_subset <- tau %>%
  select(
    ID,
    standort_aus_andere_quelle,
    Zeit,
    Gruppe,
    Sex,
    Age_start,
    CogTel,
    CpG_position,
    Psycholgical,
    rauchen,
    methylation
  ) %>%
  mutate(
    ID = as.factor(ID),
    standort_aus_andere_quelle = as.factor(standort_aus_andere_quelle),
    Zeit = as.factor(Zeit),
    Gruppe = as.factor(Gruppe),
    Sex = as.factor(Sex),
    CpG_position = as.factor(CpG_position),
    CogTel = as.numeric(CogTel),
    Psycholgical = as.numeric(Psycholgical),
    rauchen = as.factor(rauchen),
    methylation = as.numeric(methylation)
  ) %>%
  drop_na()

str(tau_subset)
```

```{r}
summary(tau_subset$methylation)
range(tau_subset$methylation)
sum(tau_subset$methylation == 1)  # Count how many exact 1s
sum(tau_subset$methylation == 0)  # For reference on zeros
```

Ein einzelner Wert = 1

```{r}
tau_subset$methylation[tau_subset$methylation == 1] <- 1 - 1e-9
```


## 

# Model Fitting

```{r}
# model
tau_glmm_1 <- glmmTMB(
  methylation ~ Gruppe * Zeit + Sex + rauchen + Age_start + CogTel + Psycholgical +
    (1 + Zeit | ID) + (1 | CpG_position),            # random intercepts + Zeit slope by ID
  ziformula = ~ Gruppe + Zeit + (1 | CpG_position),  # drivers of exact zeros
  dispformula = ~ 1 + (1 | CpG_position),            # allow CpG-specific dispersion (optional)
  family = beta_family(link = "logit"),
  data = tau_subset,
  REML = FALSE
)

summary(tau_glmm_1)

# diagnostics
tau_res <- simulateResiduals(tau_glmm_1, n = 1000)
plot(tau_res)

#(Optional) Marginal means / contrasts for Gruppe × Zeit
# em <- emmeans(tau_glmm_1, ~ Gruppe * Zeit, type = "response")
# pairs(em, by = "Zeit")  # or other contrasts

```


```{r}
tau_glmm_1 <- glmmTMB(
  methylation ~ Gruppe + Zeit + Sex + rauchen + (1|ID),            # random intercepts + Zeit slope by ID
  ziformula = ~ Zeit + (1 | CpG_position),  # drivers of exact zeros
  dispformula = ~ 1 + (1 | CpG_position), 
  family = beta_family(link = "logit"),
  data = tau_subset,
  REML = FALSE
)

summary(tau_glmm_1)

# diagnostics
tau_glmm_1 <- simulateResiduals(fit, n = 1000)
plot(tau_glmm_1)

#(Optional) Marginal means / contrasts for Gruppe × Zeit
# em <- emmeans(tau_glmm_1, ~ Gruppe * Zeit, type = "response")
# pairs(em, by = "Zeit")  # or other contrasts

```


```{r}
library(DHARMa)
tau_glmm_1 <- simulateResiduals(fit, n = 1000)
plot(tau_glmm_1)
testUniformity(tau_glmm_1)
testZeroInflation(tau_glmm_1)   # should NOT suggest extra zeros beyond your ziformula
testDispersion(tau_glmm_1)

```

```{r}
fit_phi <- update(
  fit,
  dispformula = ~ Zeit + Gruppe + (1 | ID) + (1 | CpG_position)
)
library(DHARMa)
sim_phi <- simulateResiduals(fit_phi, n = 2000)
plot(sim_phi); testUniformity(sim_phi); testDispersion(sim_phi); testZeroInflation(sim_phi)
```



```{r}
glmmTMB(
  methylation ~ Gruppe * Zeit + Sex + rauchen + Age_start + CogTel +
    (1 + Zeit | ID) + (1 | CpG_position),            # random intercepts + optional slope
  ziformula = ~ Gruppe + Zeit + (1 | CpG_position),  # what drives exact zeros
  dispformula = ~ 1 + (1 | CpG_position),            # allow CpG-specific dispersion (optional)
  family = beta_family(link = "logit"),
  data = d
)

```

```{r}
tau_glmm_CpG_ZeitID <- glmmTMB(
  methylation ~ standort_aus_andere_quelle + Gruppe + Sex + Age_start + CogTel + Zeit + (1|CpG_position) + (1|Zeit:ID),
  ziformula = ~1,
  family = beta_family(link = "logit"),
  data = tau_subset
)
summary(tau_glmm_CpG_ZeitID)
```

```{r}
tau_glmm_CpG_zeit <- glmmTMB(
  methylation ~ standort_aus_andere_quelle + Gruppe + Sex + Age_start + CogTel + Zeit + (1|CpG_position),
  ziformula = ~1,
  family = beta_family(link = "logit"),
  data = tau_subset
)
summary(tau_glmm_CpG_zeit)
```


```{r}
tau_glmm_CpG_ID <- glmmTMB(
  methylation ~ standort_aus_andere_quelle + Gruppe + Sex + Age_start + CogTel + Zeit + (1|CpG_position) + (1|ID) + (1|Zeit:Gruppe),
  ziformula = ~Zeit + Gruppe,
  family = beta_family(link = "logit"),
  data = tau_subset
)
summary(tau_glmm_CpG_ID)


```


## Fever CpGs
| Rank | CpG  |        |
|------|------|-------:|
| 1    | m331 | 0.4438 |
| 2    | m337 | 0.4236 |
| 3    | m310 | 0.3856 |
| 4    | m315 | 0.3843 |
| 5    | m323 | 0.3771 |
| -    | ---- | ------ |
| 6    | m306 | 0.2459 |
| 7    | m219 | 0.1652 |
| 8    | m300 | 0.1489 |
| 9    | m293 | 0.1295 |
| 10   | m207 | 0.1246 |
| 11   | m249 | 0.1151 |
| 12   | m277 | 0.1086 |
| 13   | m210 | 0.0785 |

### select CpGs for new data frame

```{r}

tau_top5 <- tau_subset %>%
  filter(CpG_position %in% c("m331", "m337", "m310", "m315", "m323"))
```

```{r}
summary(tau_top5)
```

```{r}
summary(tau_top5$methylation)
range(tau_top5$methylation)
sum(tau_top5$methylation == 1)  # Count how many exact 1s
sum(tau_top5$methylation == 0) 
```

```{r}
tau_glmm_top5 <- glmmTMB(
  methylation ~ Gruppe * Zeit + Sex + rauchen + Age_start + CogTel +
    (0 + Zeit | ID) + CpG_position,            # random intercepts + Zeit slope by ID
  ziformula = ~ Zeit,  # drivers of exact zeros
  family = beta_family(link = "logit"),
  data = tau_top5,
  REML = FALSE
)
summary(tau_glmm_top5)


```

```{r}
tau_top5$Zeit_num <- as.numeric(tau_top5$Zeit) - 1  # T0=0, T1=1, T2=2

fit_numSlope <- glmmTMB(
  methylation ~ CpG_position + Gruppe * Zeit + Sex + rauchen + scale(Age_start) + scale(CogTel) +
                (1 + Zeit_num | ID),
  ziformula   = ~ Zeit,                      # keep as in your current model first
  family      = beta_family(link="logit"),
  data        = tau_top5
)
summary(fit_numSlope)
```
```{r}
fit_tau_best <- fit_numSlope         # your first model: (1 + Zeit_num | ID), ziformula = ~ Zeit
fit_tau_ZI0  <- update(fit_tau_best, ziformula = ~ 1)

bbmle::AICtab(fit_tau_best, fit_tau_ZI0, sort = TRUE)
# If ΔAIC < ~2, prefer fit_tau_ZI0 (simpler ZI).

```
```{r}
fit_tau_ZI0$sdr$pdHess  # want TRUE
VarCorr(fit_tau_ZI0)    # look at |corr| < ~0.95 and sensible variances

```

```{r}
tau_top5$Zeit_c <- as.numeric(tau_top5$Zeit) - mean(as.numeric(tau_top5$Zeit))
fit_tau_ZI0_c <- update(fit_tau_ZI0, . ~ . - (1 + Zeit_num | ID) + (1 + Zeit_c | ID))
summary(fit_tau_ZI0_c)
```
ChatGPT
We selected a zero-inflated beta mixed model with constant zero inflation (≈4%) and participant-specific intercepts and linear time slopes (corr ≈ −0.82). Overall expected methylation 
𝐸
[
𝑦
]
=
(
1
−
𝑝
0
)
𝜇
E[y]=(1−p
0
	​

)μ was similar between groups at T0 and T2, but lower in Piano at T1 (Gruppe×Zeit interaction on the conditional mean significant at T1), while zeros were rare and stable over time.

```{r}
# Zeit_num = as.numeric(Zeit) - 1 (T0=0, T1=1, T2=2)
tau_best <- glmmTMB(
  methylation ~ CpG_position + Gruppe * Zeit + Sex + rauchen +
    scale(Age_start) + scale(CogTel) +
    (1 + Zeit_num | ID),          # correlated intercept + linear time slope
  ziformula   = ~ 1,              # zeros are rare (~4%), time effects NS → intercept-only
  family = beta_family(link="logit"),
  data = tau_top5
)
summary(tau_best)

```



