---
title: "Zero Inflated Beta Regression for BDNF Methylation Data"
author: "Jonas J. R. Koberschinski"
date: "24-10-2025"
lastmod: "25-10-2025"
output: html
---

# Preparation
## Load libraries

```{r}
library(tidyverse) # For data manipulation and visualization
library(dplyr) # For data manipulation
library(readxl) # For reading Excel files
library(glmmTMB) # For generalized linear mixed models
library(performance) # For model performance assessment
library(DHARMa) # For residual diagnostics
library(ggplot2) # For data visualization
```

Zero-inflated Beta-Binomial models with the2 scDMV R package

## Data import

```{r}
bdnf <- read_excel("~/Work/TBM-R-Analysis/tables/Rohdaten/BDNF_Neuroplastizität_Alter_CpGs_ohne_m11_und_p36_mit_verhalten_vollständig_umstrukturiert.xlsx")

tau <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/TAU_CpGs_mit_verhalten_ohne_p61_p59_p34_p29_umstruk.xlsx"))

psd95 <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/PSD95_CpGs_mit_verhalten_größer_umstrukturiert.xlsx"))

 ##first look at BDNF data
head(bdnf)
summary(bdnf)
## Check the structure of the dataset
str(bdnf)

```

## Subset Data and Drop NA

```{r}
# Remove rows with NA values in the 'methylierung' column
bdnf_subset <- bdnf %>%
  select(
    ID,
    Standort,
    Zeitpunkt,
    Gruppe,
    Sex,
    Age_start,
    CogTel,
    CpG_position,
    rauchen,
    methylierung
  ) %>%
  mutate(
    ID = as.factor(ID),
    Standort = as.factor(Standort),
    Zeitpunkt = as.factor(Zeitpunkt),
    Gruppe = as.factor(Gruppe),
    Sex = as.factor(Sex),
    CpG_position = as.factor(CpG_position),
    CogTel = as.numeric(CogTel),
    rauchen = as.factor(rauchen)
  ) 
bdnf_nona <- drop_na(bdnf_subset)
str(bdnf_subset)
```

# Analysis
## all cpgs

### Normality check

```{r}
# Descriptive statistics for methylierung
mean(bdnf_nona$methylierung)
sd(bdnf_nona$methylierung)
median(bdnf_nona$methylierung)

# Another Histogram
hist(bdnf_nona$methylierung, breaks = 60, main = "BDNF Methylation distribution", xlab = "Methylation")
hist(tau$methylation, breaks = 60, main = "TAU Methylation distribution", xlab = "Methylation")
hist(psd95$methylation, breaks = 60, main = "PSD95 Methylation distribution", xlab = "Methylation")

# QQ plot
## BDNF
qqnorm(bdnf_nona$methylierung)
qqline(bdnf_nona$methylierung, col = "red")

## Tau
qqnorm(tau$methylation)
qqline(tau$methylation, col = "blue")

## PSD95
qqnorm(psd95$methylation)
qqline(psd95$methylation, col = "green")


# Shapiro-Wilk normality test
shapiro.test(bdnf_nona$methylierung)

# Boxplot to visualize methylierung distribution
boxplot(bdnf_nona$methylierung, main="Boxplot of BDNF Methylation Levels", ylab="Methylation Levels")
```

zero-inflated Poisson distributions {\displaystyle \mu } = 0,114 {\displaystyle \pi } = 426

```{r}
#summary(m1 <- zeroinfl(bdnf_nona$methylierung ~ bdnf_nona$Zeitpunkt + bdnf_nona$Group | #bdnf_nona$id_umstrukturiereninst, data = bdnf_nona))
```

### Descriptive statistics for zero-inflation

```{r}
# Total count
bdnf_nona %>%
  count() # 3603


# Count of zeros
bdnf_nona %>%
  filter(methylierung == 0) %>%
  count() # 1643

# Count of non-zeros
bdnf_nona %>%
  filter(methylierung > 0) %>%
  count() # 1960

# Proportion of zeros
bdnf_nona %>%
  summarise(prop_zero = mean(methylierung == 0))


```

```{r}
bdnf_nona %>%
  summarise(
    total_mean = mean(methylierung),
    nonzero_mean = mean(methylierung[methylierung > 0]),
    n_total = n(),
    n_zero = sum(methylierung == 0),
    prop_zero = mean(methylierung == 0)
  )

```

## Zero-inflated Beta-Binomial models

### glmmTMB
```{r}
library(glmmTMB)

# Zero-inflated Beta
bdnf_glmm <- glmmTMB(
  methylierung ~ Standort + Zeitpunkt + Gruppe + Sex + rauchen + Age_start + CogTel + (1|ID) + (1|CpG_position) + (0 + Zeitpunkt | ID),
  ziformula = ~ Zeitpunkt + Gruppe,        # model for zeros
  family = beta_family(link = "logit"),
  data = bdnf_nona
)
summary(bdnf_glmm)
#preds <- predict(mod, type="response")
#range(preds)
```
```{r}
library(DHARMa)
sim <- simulateResiduals(bdnf_glmm, n = 2000)
plot(sim); testUniformity(sim); testDispersion(sim); testZeroInflation(sim)
```

```{r}
library(bbmle)
fit_noIDSlope <- update(bdnf_glmm, formula = . ~ . - (0 + Zeitpunkt | ID))
fit_noCpG     <- update(bdnf_glmm, formula = . ~ . - (1 | CpG_position))
bbmle::AICtab(bdnf_glmm, fit_noIDSlope, fit_noCpG, sort = TRUE)

```
```{r}
# Uncorrelated slopes for Zeitpunkt by ID
fit_uncorr <- update(
  bdnf_glmm,
  formula = . ~ . - (0 + Zeitpunkt | ID) + (0 + Zeitpunkt || ID)
)
summary(fit_uncorr)

# Compare fit
bbmle::AICtab(bdnf_glmm, fit_uncorr, sort = TRUE)

```
```{r}
fit_ZIzeit <- update(bdnf_glmm, ziformula = ~ Zeitpunkt)
bbmle::AICtab(bdnf_glmm, fit_ZIzeit, sort = TRUE)

```
```{r}
# 1) Use the simpler ZI:
fit_final <- update(fit_uncorr, ziformula = ~ Zeitpunkt)

# 2) Diagnostics
library(DHARMa)
sim <- simulateResiduals(fit_final, n = 2000)
plot(sim); testUniformity(sim); testDispersion(sim); testZeroInflation(sim)

# (Optional) confirm Hessian
fit_final$sdr$pdHess      # should be TRUE
fit_final$fit$convergence # should be 0
```



```{r}
# Grid at typical covariates
newd <- with(bdnf_nona, expand.grid(
  Gruppe     = levels(Gruppe),
  Zeitpunkt  = levels(Zeitpunkt),
  Standort   = names(which.max(table(Standort))),
  Sex        = names(which.max(table(Sex))),
  rauchen    = names(which.max(table(rauchen))),
  Age_start  = median(Age_start, na.rm=TRUE),
  CogTel     = median(CogTel, na.rm=TRUE),
  ID = NA, CpG_position = NA
))

mu <- predict(fit_final, newdata = newd, type = "response", se.fit = TRUE)  # beta mean
p0 <- predict(fit_final, newdata = newd, type = "zprob",    se.fit = TRUE)  # zero prob
Ey <- (1 - p0$fit) * mu$fit
cbind(newd[, c("Gruppe","Zeitpunkt")], Ey)

```


```{r}
library(emmeans)

# Conditional (beta) mean
em_mu <- emmeans(fit_final, ~ Gruppe * Zeitpunkt, component = "cond", type = "response")
summary(em_mu)
pairs(em_mu, by = "Zeitpunkt")

# Zero probability
em_zi <- emmeans(fit_final, ~ Zeitpunkt, component = "zi", type = "response")
em_zi
pairs(em_zi)  
```
```{r}
# Conditional (beta) mean, response scale + pairwise group contrasts within each time
em_mu <- emmeans(fit_final, ~ Gruppe * Zeitpunkt, component = "cond", type = "response")
summary(em_mu)

# Pairwise contrasts on the response scale (differences, not odds ratios)
contrast(em_mu, method = "pairwise", by = "Zeitpunkt", type = "response")

```
```{r}
newd <- with(bdnf_nona, expand.grid(
  Gruppe     = levels(Gruppe),
  Zeitpunkt  = levels(Zeitpunkt),
  Standort   = names(which.max(table(Standort))),
  Sex        = names(which.max(table(Sex))),
  rauchen    = names(which.max(table(rauchen))),
  Age_start  = median(Age_start, na.rm=TRUE),
  CogTel     = median(CogTel, na.rm=TRUE),
  ID = NA, CpG_position = NA
))

mu <- predict(fit_final, newdata = newd, type = "response", se.fit = TRUE)  # beta mean
p0 <- predict(fit_final, newdata = newd, type = "zprob",    se.fit = TRUE)  # zero prob
Ey <- (1 - p0$fit) * mu$fit
cbind(newd[, c("Gruppe","Zeitpunkt")], Ey)

```




```{r}
bdnf_glmm_final <- glmmTMB(
  methylierung ~ Standort + Zeitpunkt + Gruppe + Sex + rauchen +
                   Age_start + CogTel +
                   (1 | CpG_position) +          # CpG random intercept
                   (0 + Zeitpunkt || ID),        # uncorrelated time slopes by participant
  ziformula = ~ Zeitpunkt,                        # zero-inflation depends only on time
  family = beta_family(link = "logit"),
  data = bdnf_nona
)

summary(bdnf_glmm_final)
```

ChatGPT Summary:# Summary of Zero-Inflated Beta Mixed Model for BDNF Methylation Data

Final model:
Zero-inflated beta mixed model with uncorrelated participant time slopes and CpG random intercepts; zero inflation depends only on measurement time.
Fit quality: converged cleanly, no dispersion or zero-inflation problems, lowest AIC among tested candidates.
Findings: the intervention (Piano) group shows a modest reduction in BDNF promoter methylation relative to controls, while the probability of complete lack of methylation (exact zeros) decreases at the final time point (T2).

```{r}
# You probably already have these model objects:
#   bdnf_glmm      – full (correlated slopes, ZI with Gruppe+Zeitpunkt)
#   fit_uncorr     – uncorrelated slopes, ZI with Gruppe+Zeitpunkt
#   fit_ZIzeit     – uncorrelated slopes, ZI only with Zeitpunkt
#   fit_noIDSlope  – no Zeit slopes
#   fit_noCpG      – no CpG random intercept (failed)

library(bbmle)

AICtab(
  bdnf_glmm,
  fit_uncorr,
  fit_ZIzeit,
  fit_noIDSlope,
  fit_noCpG,
  sort = TRUE
)

```
```{r}
# Helper grid
newd <- with(bdnf_nona, expand.grid(
  Gruppe     = levels(Gruppe),
  Zeitpunkt  = levels(Zeitpunkt),
  Standort   = names(which.max(table(Standort))),
  Sex        = names(which.max(table(Sex))),
  rauchen    = names(which.max(table(rauchen))),
  Age_start  = median(Age_start, na.rm=TRUE),
  CogTel     = median(CogTel, na.rm=TRUE),
  ID = NA, CpG_position = NA
))

# Predictions
mu <- predict(fit_final, newdata=newd, type="response", se.fit=TRUE)
p0 <- predict(fit_final, newdata=newd, type="zprob",    se.fit=TRUE)

Ey <- (1 - p0$fit) * mu$fit
Ey.SE <- sqrt((mu$se.fit * (1 - p0$fit))^2 + (p0$se.fit * mu$fit)^2)  # delta approximation
Ey.lo <- plogis(qlogis(Ey) - 1.96 * Ey.SE)   # approximate 95 % CI on response scale
Ey.hi <- plogis(qlogis(Ey) + 1.96 * Ey.SE)

EY_tab <- cbind(newd[, c("Gruppe","Zeitpunkt")],
                Mean = round(Ey, 4),
                LCL = round(Ey.lo, 4),
                UCL = round(Ey.hi, 4))
EY_tab

```




ChatGPT:
A zero-inflated beta mixed model with uncorrelated participant time slopes and CpG random intercepts (AIC = 206.9) provided the best fit.
The intervention group (Piano) exhibited slightly lower BDNF promoter methylation (β = −0.082, p = 0.05).
Zero inflation decreased at T2 versus baseline (p = 0.036), indicating fewer complete non-methylation observations.
Between-CpG variability remained substantial (SD ≈ 0.44).
Predicted overall methylation 
𝐸
[
𝑦
]
=
(
1
−
𝑝
0
)
𝜇
E[y]=(1−p
0
	​

)μ averaged 0.10–0.11 in controls and 0.09–0.10 in the Piano group across time points.

Reduktion in Odds!


| Zeitpunkt | Controll E[y] (95% CI) | Piano E[y] (95% CI) | Δ (Control − Piano) | Piano vs Control | z      | p-value      |
|:---------:|:------------------------|:---------------------|--------------------:|-----------------:|:------:|:-------------|
| T0        | 0.0294 (0.0292–0.0296)  | 0.0273 (0.0271–0.0275) | +0.0021            | −7.1%           | 14.55  | 5.65×10⁻⁴⁸  |
| T1        | 0.0279 (0.0277–0.0281)  | 0.0259 (0.0257–0.0261) | +0.0020            | −7.2%           | 13.86  | 1.12×10⁻⁴³  |
| T2        | 0.0355 (0.0352–0.0359)  | 0.0330 (0.0328–0.0333) | +0.0025            | −7.0%           | 11.39  | 4.57×10⁻³⁰  |

For BDNF, overall expected methylation 
𝐸
[
𝑦
]
=
(
1
−
𝑝
0
)
𝜇
E[y]=(1−p
0
	​

)μ was consistently lower in the Piano group compared with Controll across all time points. At T0, Controll was 0.0294 (0.0292–0.0296) vs Piano 0.0273 (0.0271–0.0275) (Δ = 0.0021; ~7% lower in Piano). At T1, Controll 0.0279 (0.0277–0.0281) vs Piano 0.0259 (0.0257–0.0261) (Δ = 0.0020; ~7% lower). At T2, Controll 0.0355 (0.0352–0.0359) vs Piano 0.0330 (0.0328–0.0333) (Δ = 0.0025; ~7% lower). Differences are small in absolute terms (~0.002–0.003), but stable in direction (~7% lower in Piano).


```{r}
bdnf_best <- glmmTMB(
  methylierung ~ Standort + Zeitpunkt + Gruppe + Sex + rauchen +
    scale(Age_start) + scale(CogTel) +
    (1 | ID) + (0 + Zeitpunkt || ID) + (1 | CpG_position),
  ziformula   = ~ Zeitpunkt,      # ZI by time (better than +Gruppe)
  family = beta_family(link="logit"),
  data = bdnf_nona
)
summary(bdnf_best)
```




# Other Stuff

```{r}
library(dplyr)
library(ggplot2)

# Daten vorbereiten
bdnf_plot <- bdnf_nona %>%
  mutate(is_zero = ifelse(methylierung == 0, 1, 0)) %>%
  group_by(CpG_position, Gruppe) %>%
  summarise(
    mean_methyl = mean(methylierung[methylierung > 0], na.rm = TRUE),
    prop_zero = mean(is_zero, na.rm = TRUE),
    .groups = "drop"
  )

# Beta-Teil (Mittelwert der Methylierung)
p1 <- ggplot(bdnf_plot, aes(x = CpG_position, y = mean_methyl, fill = Gruppe)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Beta-Teil: Mittelwert der Methylierung (>0)",
       y = "Mittelwert", x = "CpG Position") +
  scale_fill_brewer(palette = "Set1")

# Zero-Inflation Teil
p2 <- ggplot(bdnf_plot, aes(x = CpG_position, y = prop_zero, fill = Gruppe)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Zero-Inflation: Anteil Nullwerte",
       y = "Proportion Nullwerte", x = "CpG Position") +
  scale_fill_brewer(palette = "Set2")

# Plots anzeigen
library(gridExtra)
grid.arrange(p1, p2, nrow = 2)

```

```{r}
testDispersion(mod)
```

```{r}
library(DHARMa)

# Simulate residuals
sim_res <- simulateResiduals(fittedModel = mod_CpG_ID, n = 1000)
residuals(sim_res)

# Plot residuals
plot(sim_res)
```

https://cran.r-project.org/web/packages/DHARMa/vignettes/DHARMa.html

```{r}
check_model(mod)
check_model(mod_CpG_ID)
```

## BDNF (High Variance CpGs)

```{r}
bdnf_reduced <- bdnf %>%
  dplyr::select(methylierung, Group, Zeitpunkt, Standort, Geschlecht, Age_start, CogTel, id_umstrukturieren, CpG_position) %>%
  dplyr::filter(CpG_position %in% c("m148", "m066", "m111", "m058", "m039")) #( maybe add m035)
```

```{r}
hist(bdnf_reduced$methylierung, breaks = 60, main = "BDNF Methylation distribution", xlab = "Methylation")
```

```{r}
library(glmmTMB)

# Zero-inflated Beta
mod <- glmmTMB(
  methylierung ~ Standort + Zeitpunkt + Group + Geschlecht + Age_start + CogTel,
  ziformula = ~ Group + Zeitpunkt,        # model for zeros
  family = beta_family(link = "logit"),
  data = bdnf_reduced
)
summary(mod)
#preds <- predict(mod, type="response")
#range(preds)
```

Bestes Modell (Geringste AIC) methylierung \~ Standort + Gruppe + Sex + Age_start + CogTel + Zeitpunkt + (1 \| CpG_position) + (1 \| ID)
