---
title: "Zero Inflated Beta Regression for PSD95 Methylation Data"
author: "Jonas J. R. Koberschinski"
date: "25-10-2025"
lastmod: "25-10-2025"
output: html
editor: 
  markdown: 
    wrap: 72
---

# Preparation

## Load libraries

```{r}
library(tidyverse) # For data manipulation and visualization
library(dplyr) # For data manipulation
library(readxl) # For reading Excel files
library(glmmTMB) # For generalized linear mixed models
library(performance) # For model performance assessment
library(DHARMa) # For residual diagnostics
library(ggplot2) # For data visualization
```

## Data import

```{r}
psd95 <- suppressWarnings(read_xlsx("~/Work/TBM-R-Analysis/tables/Rohdaten/PSD95_CpGs_mit_verhalten_größer_umstrukturiert.xlsx"))

head(psd95)
```

## Subset Data and Drop NA

```{r}
# Create a subset of the data with relevant columns and drop rows with NA values
psd95_subset <- psd95 %>%
  select(
    ID,
    standort_aus_andere_quelle,
    Zeitpunkt,
    Gruppe,
    Sex,
    Age_start,
    CogTel,
    CpG_Position,
    Psycholgical,
    rauchen,
    methylation
  ) %>%
  mutate(
    ID = as.factor(ID),
    standort_aus_andere_quelle = as.factor(standort_aus_andere_quelle),
    Zeitpunkt = as.factor(Zeitpunkt),
    Gruppe = as.factor(Gruppe),
    Sex = as.factor(Sex),
    CpG_Position = as.factor(CpG_Position),
    CogTel = as.numeric(CogTel),
    rauchen = as.factor(rauchen)
  ) %>%
  drop_na()

str(psd95_subset)
```

```{r}
summary(psd95_subset$methylation)
range(psd95_subset$methylation)
sum(psd95_subset$methylation == 1)  # Count how many exact 1s
sum(psd95_subset$methylation == 0)  # For reference on zeros, ]
```

Ein einzelner Wert = 1

```{r}
#N <- nrow(psd95_subset)  # Sample size
#psd95_subset$methylation_trans <- (psd95_subset$methylation * (N - 1) + 0.5) / N
```

```{r}
psd95_subset$methylation <- ifelse(
  psd95_subset$methylation == 1,
  0.999999,
  psd95_subset$methylation
)
```

```{r}
# Center/scale continuous predictors for stability
psd95_subset$Age_start_sc <- scale(psd95_subset$Age_start)
psd95_subset$CogTel_sc    <- scale(psd95_subset$CogTel)

psd95_glmm_CpG_ID <- glmmTMB(
  methylation ~ standort_aus_andere_quelle + Gruppe * Zeitpunkt + 
                  Sex + rauchen + Age_start_sc + CogTel_sc +
                  (1 | CpG_Position) + (1 | ID) + (0 + Zeitpunkt || ID),
  ziformula = ~ Zeitpunkt + Gruppe,
  dispformula = ~ 1 + (1 | CpG_Position),   # optional: CpG-specific overdispersion
  family = beta_family(link = "logit"),
  data = psd95_subset
)
summary(psd95_glmm_CpG_ID)
```

```{r}
library(DHARMa)
sim_psd95 <- simulateResiduals(psd95_glmm_CpG_ID, n = 2000)
plot(sim_psd95)
testUniformity(sim_psd95)
testDispersion(sim_psd95)
testZeroInflation(sim_psd95)

```

```{r}
psd95_glmm_CpG_ID_phi <- update(psd95_glmm_CpG_ID,
  dispformula = ~ Zeitpunkt + Gruppe + (1 | CpG_Position)
)

```

```{r}
sim_phi <- simulateResiduals(psd95_glmm_CpG_ID_phi, n=1500)
testDispersion(sim_phi)
```

```{r}
#psd95_subset$obs <- seq_len(nrow(psd95_subset))

#psd95_glmm_CpG_ID_olre <- update(
#  psd95_glmm_CpG_ID,
#  formula = . ~ . + (1 | obs)
#)

#library(DHARMa)
#sim_olre <- simulateResiduals(psd95_glmm_CpG_ID_olre, n = 1500)
#testDispersion(sim_olre)
```

```{r}
#psd95_glmm_CpG_ID_olre <- glmmTMB(
#  methylation ~ standort_aus_andere_quelle + Gruppe * Zeitpunkt +
#    Sex + rauchen + Age_start_sc + CogTel_sc +
#    (1 | CpG_Position) + (1 | ID) + (0 + Zeitpunkt || ID) + (1 | obs),
#  ziformula = ~ Zeitpunkt + Gruppe,
#  dispformula = ~ 1 + (1 | CpG_Position),
#  family = beta_family(link = "logit"),
#  data = psd95_subset,
#  control = glmmTMBControl(
#    optimizer = optim,
#    optArgs = list(method = "BFGS"),
#    parallel = 1
#  )
#)
#summary(psd95_glmm_CpG_ID_olre)
```

```{r}
#psd95_glmm_CpG_ID_olre <- update(
#  psd95_glmm_CpG_ID,
#  formula = . ~ . + (1 | obs),
#  start = list(theta = c(log(0.05)))  # tiny variance
#)
```

```{r}
#psd95_glmm_CpG_ID_olre <- update(
#  psd95_glmm_CpG_ID,
#  formula = . ~ . + (1 | obs),
#  control = glmmTMBControl(optimizer = optim, optArgs = list(method="nlminb"))
#)

```

```{r}
#psd95_glmm_CpG_OLREtest <- glmmTMB(
#  methylation ~ standort_aus_andere_quelle + Gruppe * Zeitpunkt +
#    Sex + rauchen + Age_start_sc + CogTel_sc +
#    (1 | CpG_Position) + (1 | obs) + (1 | ID) + (0 + Zeitpunkt || ID),
#  ziformula = ~ Zeitpunkt + Gruppe,
#  family = beta_family(link="logit"),
#  data = psd95_subset
#)

#summary(psd95_glmm_CpG_OLREtest)
```

```{r}
#library(DHARMa)
#sim_olre <- simulateResiduals(psd95_glmm_CpG_OLREtest, n = 1500)
#testDispersion(sim_olre)

```

Frustrane Modelle wegen bleibender Übderdispersion

Versuche auf CpGs zu verzichten gemäß PCA. Vielleicht für Tau auch
notwendig.

Nach Chat: ✅ It’s justifiable if you do it transparently and carefully

You can absolutely restrict your inferential modeling to those 3 CpGs
provided that:

You pre-specify or justify the reduction. e.g., “PCA indicated that 3
CpG positions explained \>95 % of variance in PSD95 methylation;
subsequent mixed modeling focused on these sites.”

You verify that these 3 CpGs are not statistical outliers or sequencing
artefacts. (Check their read quality and call rate.)

You describe them biologically. e.g., “CpG7–CpG9 near the transcription
start site were the most variable.”

You note that other CpGs were excluded because they carried redundant or
negligible information.

## Fever CpGs

| Rank | CpG  |       |
|------|------|------:|
| 1    | p95  | 0.664 |
| 2    | p60  | 0.484 |
| 3    | p23  | 0.445 |
| \-   | ---  | ----- |
| 4    | m35  | 0.153 |
| 5    | m28  | 0.103 |
| 6    | m418 | 0.097 |
| 7    | m24  | 0.094 |
| 8    | p91  | 0.094 |
| 9    | m400 | 0.086 |
| 10   | m11  | 0.078 |

### select CpGs for new data frame

```{r}
psd95_top3 <- psd95_subset %>%
  filter(CpG_Position %in% c("p95", "p60", "p23"))
```

```{r}
summary(psd95_top3)
```

### fit model with selected CpGs

```{r}
  # 3) Fit zero-inflated beta mixed model (CpG as fixed effect)
library(glmmTMB)
fit_top3 <- glmmTMB(
  methylation ~ Gruppe * Zeitpunkt + Sex + rauchen + Age_start_sc + CogTel_sc +
                 (1 | ID) + (0 + Zeitpunkt || ID),
  ziformula   = ~ Zeitpunkt + Gruppe,
  family      = beta_family(link = "logit"),
  data        = psd95_top3
)

summary(fit_top3)
```

```{r}
library(DHARMa)
sim <- simulateResiduals(fit_top3, n = 2000)
plot(sim); testUniformity(sim); testDispersion(sim); testZeroInflation(sim)
```

```{r}
# 1) Keep CpG fixed; simplify the zero part (Gruppe was NS there)
fit_top3_ZIzeit <- update(fit_top3, ziformula = ~ Zeitpunkt)

# 2) Let dispersion vary a bit by time (soaks up the slight under-dispersion)
fit_top3_phi <- update(fit_top3_ZIzeit, dispformula = ~ Zeitpunkt)

# Compare and re-check DHARMa
bbmle::AICtab(fit_top3, fit_top3_ZIzeit, fit_top3_phi, sort = TRUE)

library(DHARMa)
sim <- simulateResiduals(fit_top3_phi, n = 2000)
plot(sim); testUniformity(sim); testDispersion(sim); testZeroInflation(sim)

```

```{r}
library(emmeans)
best <- fit_top3_phi  # your chosen model

# 1) Get component-wise estimates on the response scale
em_mu <- as.data.frame(emmeans(best, ~ Gruppe * Zeitpunkt, component="cond", type="response"))
em_zi <- as.data.frame(emmeans(best, ~ Zeitpunkt,          component="zi",   type="response"))

# 2) Merge zero-prob by time (same for both groups)
names(em_zi)[names(em_zi)=="response"] <- "p0"
names(em_zi)[names(em_zi)=="SE"]       <- "p0_SE"
EY <- merge(em_mu, em_zi[, c("Zeitpunkt","p0","p0_SE")], by="Zeitpunkt", all.x=TRUE)

# 3) Combine: Ey = (1 - p0)*mu, with delta-method SE (ignoring tiny cross-covariance)
EY$Ey    <- (1 - EY$p0) * EY$response
EY$Ey_SE <- sqrt( (1 - EY$p0)^2 * EY$SE^2 + (EY$response^2) * EY$p0_SE^2 )

# 4) 95% CI on response scale (simple Normal approx)
EY$LCL <- pmax(0, EY$Ey - 1.96*EY$Ey_SE)
EY$UCL <- pmin(1, EY$Ey + 1.96*EY$Ey_SE)

# (Optional) slightly better-bounded CI via logit transform
logit   <- function(x) log(x/(1-x)); invlogit <- function(x) 1/(1+exp(-x))
g       <- logit(pmin(pmax(EY$Ey, 1e-6), 1-1e-6))
dg_dEy  <- 1/(EY$Ey*(1 - EY$Ey))
se_g    <- dg_dEy * EY$Ey_SE
EY$LCL2 <- invlogit(g - 1.96*se_g)
EY$UCL2 <- invlogit(g + 1.96*se_g)

# 5) Tidy output
out <- EY[, c("Gruppe","Zeitpunkt","Ey","LCL","UCL","LCL2","UCL2")]
out[order(out$Zeitpunkt, out$Gruppe), ]

```

ChatGPT: What to report (template)

Model: zero-inflated beta with CpG (top-3) as fixed effect; random
(1\|ID) and uncorrelated (0+Zeitpunkt\|\|ID); ziformula = \~ Zeitpunkt;
dispformula = \~ Zeitpunkt.

Diagnostics: dispersion test \~ OK (slight under-dispersion), ZI OK, KS
OK.

Effects: report (i) group contrasts at each time from pairs(em_mu,
by="Zeitpunkt"), (ii) ZI by time from em_zi, and (iii) overall expected
methylation table EY_tab with 95% bootstrapped CIs.

Sensitivity: include the no-CpG model and the CpG-fixed/no-phi model in
the appendix—AICs worse, conclusions unchanged.

That’s a publishable endpoint. If you want, I can help format EY_tab
into a nice table (means ± CI) and give a ggplot snippet for a clean
figure.

## Report ChatGPT

Overall expected methylation

| Zeitpunkt | Controll E\[y\] (95% CI) | Piano E\[y\] (95% CI) | Δ (Controll − Piano) | Piano vs Controll |
|:-------------:|:-------------:|:-------------:|:-------------:|:-------------:|
| T0 | 0.136 (0.110–0.167) | 0.146 (0.119–0.178) | −0.010 | +7.1% (p = 0.65) |
| T1 | 0.161 (0.134–0.192) | 0.156 (0.129–0.186) | +0.005 | −3.2% (p = 0.80) |
| T2 | 0.190 (0.157–0.229) | 0.134 (0.108–0.165) | **+0.056** | **−29.6% (p = 0.016)** |

Notes

CIs are the bounded (logit-transformed) intervals you computed
(LCL2–UCL2), so they stay within \[ 0 , 1\] \[0,1\].

p-values use a quick delta method combining the component SEs (no
bootstrap). T2 remains clearly different; T0/T1 are not.

One-line result you can paste

Overall expected methylation 𝐸 \[ 𝑦\] E\[y\] was similar between groups
at T0 and T1, but at T2 Control exceeded Piano (0.190 \[0.157–0.229\] vs
0.134 \[0.108–0.165\]; Δ = 0.056; \~30% lower in Piano; p ≈ 0.016).


```{r}
psd95_best <- glmmTMB(
  methylation ~ Gruppe * Zeitpunkt + Sex + rauchen + scale(Age_start) + scale(CogTel) +
    (1 | ID) + (0 + Zeitpunkt || ID),          # uncorrelated time slopes
  ziformula   = ~ Zeitpunkt,                    # ZI varies by time
  dispformula = ~ Zeitpunkt,                    # phi varies by time (best by AIC)
  family = beta_family(link="logit"),
  data = psd95_top3
)
summary(psd95_best)
```

```{r}
report(psd95_best)
```


```{r}
library(report)
# 1) minimal subset + clean types
dat <- psd95 %>% 
  transmute(
    ID = factor(ID),
    Gruppe = factor(Gruppe),
    Zeitpunkt = factor(Zeitpunkt),
    CpG = factor(CpG_Position),                    # rename for convenience
    progress = case_when(                          # uses your mapping idea
      as.character(Zeitpunkt) == "T1" ~ as.numeric(Piano_Fortschritt_Teacher_6Mon),
      as.character(Zeitpunkt) == "T2" ~ as.numeric(Piano_Fortschritt_Teacher_12Mon),
      TRUE ~ NA_real_
    ),
    CogTel = suppressWarnings(as.numeric(CogTel)),
    meth_raw = suppressWarnings(as.numeric(methylation))
  ) %>%
  filter(!is.na(ID), !is.na(Gruppe), !is.na(Zeitpunkt), !is.na(CpG), !is.na(meth_raw)) %>%
  filter(!is.na(progress), !is.na(CogTel))         # keep only rows where you have progress + CogTel

# 2) model-only copy in (0,1) for beta part (raw stays unchanged)
eps <- 1e-6
dat <- dat %>%
  mutate(meth = pmin(pmax(meth_raw, eps), 1 - eps))

# 3) ZIBR model (zero-inflated beta)
m_zibr <- glmmTMB(
  meth ~ Gruppe + Zeitpunkt + progress + CogTel + (1 | ID/CpG),
  ziformula = ~ 1,
  family = beta_family(link = "logit"),
  data = dat
)

summary(m_zibr)
car::Anova(m_zibr, type = 2)  # 
report
```

```{r}
library(dplyr)
library(glmmTMB)

dat <- psd95 %>% 
  transmute(
    ID = factor(ID),
    Gruppe = factor(Gruppe),
    Zeitpunkt = factor(Zeitpunkt),
    CpG_Position = factor(CpG_Position),

    progress = case_when(
      as.character(Zeitpunkt) == "T1" ~ suppressWarnings(as.numeric(Piano_Fortschritt_Teacher_6Mon)),
      as.character(Zeitpunkt) == "T2" ~ suppressWarnings(as.numeric(Piano_Fortschritt_Teacher_12Mon)),
      TRUE ~ NA_real_
    ),

    CogTel = suppressWarnings(as.numeric(CogTel)),
    meth_raw = suppressWarnings(as.numeric(methylation))
  ) %>%
  # keep only the CpGs you want
  filter(CpG_Position %in% c("p95", "p60", "p23")) %>%
  # keep essential rows
  filter(!is.na(ID), !is.na(Gruppe), !is.na(Zeitpunkt), !is.na(CpG_Position), !is.na(meth_raw)) %>%
  # progress only exists for T1/T2 with your mapping
  filter(!is.na(progress), !is.na(CogTel))

# model-only copy for beta likelihood (raw stays unchanged)
eps <- 1e-6
dat <- dat %>% mutate(meth = pmin(pmax(meth_raw, eps), 1 - eps))

m_zibr <- glmmTMB(
  meth ~ Gruppe * Zeitpunkt + progress + CogTel + (1 | ID/CpG_Position),
  ziformula = ~ 1,
  family = beta_family(link = "logit"),
  data = dat
)

summary(m_zibr)
anova(m_zibr)

# sanity check
c(
  n = nrow(dat),
  zeros = sum(dat$meth_raw == 0),
  ones  = sum(dat$meth_raw == 1)
)

```

# New

```{r}
library(dplyr)
library(glmmTMB)
library(report)
library(performance)

```
performance)

```

